---
title: "Fish Reproduction Calibration Diagnostics"
author: "ZooMSS Calibration Framework"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6
)
```

```{r libraries}
devtools::load_all()
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
```

# Load Calibration Results

```{r load-results}
cache_dir <- here::here("calibration_repro_cache")
calib <- readRDS(file.path(cache_dir, "calibration_results.rds"))

cat("=== Calibration Settings ===\n")
cat(sprintf("  LHS samples:      %d\n", calib$settings$n_samples))
cat(sprintf("  SST:              %.0f deg C\n", calib$settings$sst))
cat(sprintf("  SST amplitude:    %.1f\n", calib$settings$sst_amplitude))
cat(sprintf("  Chl amplitude:    %.1f\n", calib$settings$chl_amplitude))
cat(sprintf("  Benchmark years:  %d\n", calib$settings$benchmark_years))
cat(sprintf("  Screening years:  %d\n", calib$settings$screening_years))
cat(sprintf("  Refinement years: %d\n", calib$settings$refinement_years))
cat(sprintf("  Assessment years: %d (final)\n", calib$settings$assess_years))
cat(sprintf("  Best score:       %.4f\n", calib$best$score))
```

# Part 1: LHS Exploration Overview

## Score Distribution

```{r lhs-scores, fig.height=8}
lhs <- calib$lhs_results

p1 <- ggplot(lhs, aes(x = score)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = calib$best$score, colour = "red", linetype = "dashed") +
  annotate("text", x = calib$best$score, y = Inf, label = "Best",
           vjust = 2, colour = "red") +
  theme_bw() +
  labs(x = "Composite Score", y = "Count", title = "LHS Score Distribution")

metrics <- lhs[, c("coexistence", "stability", "zoo_comp", "fish_ratio", "spectrum")]
metrics_long <- pivot_longer(metrics, everything(), names_to = "metric", values_to = "value")

p2 <- ggplot(metrics_long, aes(x = metric, y = value)) +
  geom_boxplot(fill = "lightblue") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(x = "Metric", y = "Score (0 = best)", title = "Metric Score Distributions")

p3 <- ggplot(lhs, aes(x = coexistence, y = zoo_comp, colour = score)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_colour_viridis_c(option = "plasma", direction = -1) +
  geom_hline(yintercept = 0.3, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0.01, linetype = "dashed", colour = "grey50") +
  annotate("rect", xmin = 0, xmax = 0.01, ymin = 0, ymax = 0.3,
           fill = "green", alpha = 0.1) +
  theme_bw() +
  labs(x = "Coexistence Penalty", y = "Zoo Composition Penalty",
       colour = "Score", title = "Filter Region (green = passes both constraints)")

wrap_plots(p1, p2, p3, ncol = 1)
```

## Parameter Sensitivity

```{r param-sensitivity, fig.height=10}
param_names <- calib$param_space$name
plots <- lapply(param_names, function(pn) {
  if (pn %in% names(lhs)) {
    ggplot(lhs, aes(x = .data[[pn]], y = score)) +
      geom_point(alpha = 0.2, size = 0.5) +
      geom_smooth(method = "loess", se = FALSE, colour = "red") +
      theme_bw() + labs(title = pn) +
      theme(plot.title = element_text(size = 9))
  }
})
plots <- plots[!sapply(plots, is.null)]
wrap_plots(plots, ncol = 3)
```

# Part 2: Best Calibrated Parameters

```{r best-params}
best <- calib$best
par <- best$par

cat("=== Best Parameters ===\n")
cat("\nShared (all fish groups):\n")
for (nm in c("PPMR", "FeedWidth", "K_growth", "f_M", "repro_eff")) {
  cat(sprintf("  %-12s = %.4f\n", nm, par[nm]))
}
cat(sprintf("\n  R_frac = 1 - f_M - K_growth = %.4f\n",
            1 - par["f_M"] - par["K_growth"]))

cat("\nGroup-specific:\n")
cat(sprintf("  %-12s  Small    Med      Large\n", ""))
for (base in c("Wmat", "ZSpre", "ZSexp")) {
  vals <- par[paste0(base, "_", c("S", "M", "L"))]
  cat(sprintf("  %-12s  %-8.4f %-8.4f %-8.4f\n", base, vals[1], vals[2], vals[3]))
}

cat("\n=== Metric Breakdown ===\n")
for (nm in names(best$details$metric_scores)) {
  cat(sprintf("  %-15s = %.4f\n", nm, best$details$metric_scores[nm]))
}
```

# Part 3: Validation Runs

```{r validation-setup}
benchmark <- calib$benchmark
chl_levels <- benchmark$chl_levels
log10_chl <- benchmark$log10_chl
assess_years <- benchmark$assess_years %||% 100

# Seasonal forcing parameters from benchmark
sst_amp <- benchmark$sst_amplitude %||% 4
chl_amp <- benchmark$chl_amplitude %||% 0.5

Groups <- getGroups()
Groups <- apply_repro_params(best$par, Groups)

cache_dir_val <- file.path(cache_dir, "validation")
if (!dir.exists(cache_dir_val)) dir.create(cache_dir_val, recursive = TRUE)

validation_chl_idx <- seq(1, length(chl_levels), by = 2)
validation_results <- list()

for (i in validation_chl_idx) {
  cache_file <- file.path(cache_dir_val, sprintf("val_chl%d.rds", i))
  if (file.exists(cache_file)) {
    validation_results[[as.character(i)]] <- readRDS(cache_file)
  } else {
    input_params <- create_seasonal_input(
      n_years = 400, dt = 0.1,
      base_sst = benchmark$sst, base_chl = chl_levels[i],
      sst_amplitude = sst_amp, chl_amplitude = chl_amp
    )
    mdl <- zoomss_model(input_params = input_params, Groups = Groups, isave = 2)
    saveRDS(mdl, cache_file)
    validation_results[[as.character(i)]] <- mdl
  }
}
```

## Zooplankton Community Composition Comparison

```{r zoo-comparison, fig.height=8}
zoo_names <- benchmark$zoo_names
n_zoo <- length(zoo_names)

calib_zoo_prop <- matrix(NA, nrow = length(validation_chl_idx), ncol = n_zoo,
                         dimnames = list(NULL, zoo_names))
for (j in seq_along(validation_chl_idx)) {
  ci <- validation_chl_idx[j]
  mdl <- validation_results[[as.character(ci)]]
  avg <- averageTimeSeries(mdl, var = "biomass", n_years = assess_years)
  avg_total <- rowSums(avg)  # sum across size bins -> per group total
  zoo_idx <- which(mdl$param$Groups$Type == "Zooplankton")
  zoo_bm <- avg_total[zoo_idx]
  zoo_total <- sum(zoo_bm)
  if (zoo_total > 0) calib_zoo_prop[j, ] <- zoo_bm / zoo_total
}

comp_data <- data.frame()
for (j in seq_along(validation_chl_idx)) {
  ci <- validation_chl_idx[j]
  for (sp in seq_along(zoo_names)) {
    comp_data <- rbind(comp_data, data.frame(
      log10_chl = log10_chl[ci], Species = zoo_names[sp],
      Legacy = benchmark$zoo_proportions[ci, sp],
      Calibrated = calib_zoo_prop[j, sp]
    ))
  }
}

legacy_long <- comp_data %>%
  select(log10_chl, Species, proportion = Legacy) %>%
  mutate(Model = "Legacy (repro_on=0)")
calib_long <- comp_data %>%
  select(log10_chl, Species, proportion = Calibrated) %>%
  mutate(Model = "Calibrated")
all_comp <- rbind(legacy_long, calib_long)
all_comp$Species <- factor(all_comp$Species, levels = zoo_names)

Groups_ref <- getGroups()
zoo_colours <- Groups_ref$PlotColour[Groups_ref$Type == "Zooplankton"]
names(zoo_colours) <- zoo_names

ggplot(all_comp, aes(x = log10_chl, y = proportion, fill = Species)) +
  geom_area(position = "stack") +
  scale_fill_manual(values = zoo_colours) +
  facet_wrap(~Model, ncol = 1) +
  theme_bw() +
  labs(x = expression(log[10](Chlorophyll~a)), y = "Biomass Proportion",
       title = "Zooplankton Community Composition: Legacy vs Calibrated")
```

## Fish Biomass Across Gradient

```{r fish-gradient, fig.height=6}
fish_names <- benchmark$fish_names
fish_data <- data.frame()

for (j in seq_along(validation_chl_idx)) {
  ci <- validation_chl_idx[j]
  mdl <- validation_results[[as.character(ci)]]
  avg <- averageTimeSeries(mdl, var = "biomass", n_years = assess_years)
  avg_total <- rowSums(avg)  # sum across size bins -> per group total
  fish_grps <- mdl$param$fish_grps
  for (f in seq_along(fish_names)) {
    fish_data <- rbind(fish_data, data.frame(
      log10_chl = log10_chl[ci], Fish_Group = fish_names[f],
      Legacy = benchmark$fish_biomass[ci, f],
      Calibrated = avg_total[fish_grps[f]]
    ))
  }
}

fish_long <- fish_data %>%
  pivot_longer(cols = c(Legacy, Calibrated), names_to = "Model", values_to = "Biomass")

ggplot(fish_long, aes(x = log10_chl, y = log10(pmax(Biomass, 1e-30)),
                       colour = Fish_Group, linetype = Model)) +
  geom_line(linewidth = 1) + geom_point(size = 2) + theme_bw() +
  labs(x = expression(log[10](Chlorophyll~a)), y = expression(log[10](Biomass)),
       title = "Fish Biomass: Legacy vs Calibrated")
```

## Biomass Time Series at Representative Chlorophyll Levels

```{r timeseries, fig.height=12}
repr_idx <- c(1, ceiling(length(validation_chl_idx)/3),
              ceiling(2*length(validation_chl_idx)/3),
              length(validation_chl_idx))
repr_idx <- unique(pmin(repr_idx, length(validation_chl_idx)))

ts_plots <- lapply(repr_idx, function(j) {
  ci <- validation_chl_idx[j]
  mdl <- validation_results[[as.character(ci)]]
  plotTimeSeries(mdl, by = "biomass",
                 species = c("Fish_Small", "Fish_Med", "Fish_Large"),
                 transform = "log10") +
    ggtitle(sprintf("log10(chl) = %.1f", log10_chl[ci])) + xlim(300, 400)
})
wrap_plots(ts_plots, ncol = 1)
```

## Reproduction Diagnostics

```{r repro-diag, fig.height=8}
mid_j <- ceiling(length(validation_chl_idx) / 2)
mid_ci <- validation_chl_idx[mid_j]
mdl_mid <- validation_results[[as.character(mid_ci)]]

p_ssb <- plotReproduction(mdl_mid, by = "SSB") +
  ggtitle(sprintf("SSB (log10 chl = %.1f)", log10_chl[mid_ci])) + xlim(300, 400)
p_rec <- plotReproduction(mdl_mid, by = "recruitment") +
  ggtitle("Recruitment") + xlim(300, 400)
p_repro <- plotReproduction(mdl_mid, by = "repro_output") +
  ggtitle("Reproductive Output") + xlim(300, 400)

wrap_plots(p_ssb, p_rec, p_repro, ncol = 1)
```

# Part 4: Refinement Comparison

```{r refinement-table}
n_refined <- length(calib$refined)
ref_summary <- data.frame(
  Candidate = seq_len(n_refined),
  Score = sapply(calib$refined, function(x) x$score),
  Convergence = sapply(calib$refined, function(x) x$convergence)
)
for (nm in names(calib$refined[[1]]$details$metric_scores)) {
  ref_summary[[nm]] <- sapply(calib$refined, function(x) x$details$metric_scores[nm])
}
knitr::kable(ref_summary, digits = 4, caption = "Refined candidate comparison")
```

# Part 5: Post-hoc Yield Curve Validation

```{r yield-curves, fig.height=6}
yield_data <- yield_curve_validation(
  par = best$par, fmort_levels = seq(0, 2, by = 0.2),
  chl = 1.0, sst = benchmark$sst, n_years = 400,
  sst_amplitude = sst_amp, chl_amplitude = chl_amp
)

ggplot(yield_data, aes(x = Fmort, y = Yield, colour = Fish_Group)) +
  geom_line(linewidth = 1) + geom_point(size = 2) + theme_bw() +
  labs(x = "Fishing Mortality (F)", y = "Yield",
       title = "Yield Curves at Calibrated Parameters (Chl = 1 mg/m3)")
```

# Session Info

```{r}
sessionInfo()
```
