---
title: "ZooMSS Development & Research Progress Report"
subtitle: "Marine Ecosystem Modeling and Antarctic Regional Studies"
author: "Your Name"
date: last-modified
date-format: "MMMM D, YYYY"
bibliography: references.bib
---

```{r setup}
#| include: false

# Load required packages
required_packages <- c("knitr", "dplyr", "ggplot2", "plotly", "DT", "kableExtra", 
                      "tidyr", "viridis", "grid")

for(pkg in required_packages) {
  if(!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, quiet = TRUE)
  }
  library(pkg, character.only = TRUE, quietly = TRUE)
}

# Set global chunk options
opts_chunk$set(
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  dpi = 300,
  out.width = "100%"
)
```

::: {.callout-note appearance="simple"}
## Executive Summary

This report outlines progress and future directions for two major research projects: **ZooMSS marine ecosystem model development** and **ACEAS regional modeling in East Antarctica**. Key milestones include implementing fish reproduction for FishMIP collaboration, presenting to UQ lab group (December 2024), and showcasing new capabilities at CMIP workshop (March 2026).
:::

# Introduction

Marine ecosystem modeling faces increasing demands for biological realism and integration with global climate projections. This report details progress on the **ZooMSS (Zooplankton Model of Size Spectra)** framework^[ZooMSS represents marine ecosystems through size-structured population dynamics] alongside complementary work in Antarctic regional modeling through the **ACEAS (Antarctic Climate & Ecosystems CRC)** program.

```{r timeline-setup}
#| echo: false
# Create timeline data
timeline_data <- data.frame(
  Project = c("ZooMSS", "ZooMSS", "ZooMSS", "ACEAS", "ZooMSS", "ACEAS"),
  Task = c("Fish Reproduction Implementation", "UQ Lab Presentation", "CMIP Workshop Presentation", 
           "Regional Modeling Phase 1", "Long-term Scenarios (2300)", "Regional Modeling Phase 2"),
  Start = as.Date(c("2024-09-01", "2024-12-01", "2026-03-01", 
                    "2024-10-01", "2025-01-01", "2025-06-01")),
  End = as.Date(c("2024-11-30", "2024-12-15", "2026-03-15", 
                  "2025-05-31", "2025-08-31", "2025-12-31")),
  Priority = c("Critical", "High", "Critical", "Medium", "Medium", "High"),
  Status = c("In Progress", "Planned", "Planned", "Planned", "Planned", "Planned")
)
```

# ZooMSS Model Development {#sec-zoomss}

## Current Status & Achievements

The ZooMSS model has undergone significant development in 2024, with major advances in **reproduction dynamics implementation** for fish functional groups. This enhancement is crucial for the model's continued participation in international collaborations.

::: {.column-margin}
**FishMIP** (Fisheries and Marine Ecosystem Model Intercomparison Project) requires all participating models to include fishing scenarios, which necessitates realistic fish population dynamics including reproduction.
:::

### Fish Reproduction Implementation {#sec-reproduction}

The implementation of fish reproduction represents a fundamental advancement in ZooMSS capabilities:

```{r reproduction-features}
#| echo: false
reproduction_features <- data.frame(
  Component = c("Spawning Stock Biomass (SSB)", "Recruitment Dynamics", "Seasonal Spawning", 
                "Size-dependent Maturity", "Energy Allocation", "Population Feedback"),
  Status = c("✅ Complete", "✅ Complete", "✅ Complete", 
             "✅ Complete", "🔄 In Progress", "🔄 Testing"),
  Description = c("Biomass of reproductively mature individuals",
                  "Beverton-Holt stock-recruitment relationship",
                  "Temperature and photoperiod-driven spawning cycles",
                  "Probabilistic maturation by body size",
                  "Trade-offs between growth and reproduction",
                  "Density-dependent population regulation")
)

kable(reproduction_features, 
      caption = "Fish Reproduction Implementation Status",
      col.names = c("Component", "Status", "Description")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  column_spec(2, width = "15%") %>%
  column_spec(3, width = "50%")
```

### Extended Testing Framework

A comprehensive testing framework has been developed to validate reproduction dynamics:

- **500-year simulations** to assess long-term stability
- **Chlorophyll gradient testing** (oligotrophic to eutrophic conditions)
- **Seasonal variation analysis** across environmental gradients
- **Multi-species interaction** validation for 12 functional groups

::: {.column-margin}
The testing framework explores model behavior across chlorophyll levels from 0.01 mg/m³ (oligotrophic) to 10.0 mg/m³ (eutrophic), representing the full range of marine productivity.
:::

## FishMIP Integration Requirements

ZooMSS participation in FishMIP depends on the successful implementation of fishing scenarios, which require:

1. **Realistic fish population dynamics** ✅ *Complete*
2. **Spawning stock biomass tracking** ✅ *Complete*  
3. **Recruitment variability** ✅ *Complete*
4. **Fishing mortality implementation** 🔄 *Next Phase*
5. **Economic scenario coupling** 📋 *Planned*

## Visualization Capabilities

The model now includes comprehensive visualization tools showing both zooplankton and fish dynamics:

```{r vis-capabilities}
#| echo: false
vis_types <- data.frame(
  `Plot Type` = c("Size Spectra", "Time Series - Abundance", "Time Series - Biomass", 
                  "Fish Reproduction Metrics", "Environmental Forcing", "Trophic Interactions"),
  `Groups Displayed` = c("All 12 functional groups", "All groups (9 zooplankton + 3 fish)", 
                         "All groups with biomass", "Fish-specific (SSB, recruitment)", 
                         "Temperature, chlorophyll", "Predator-prey relationships"),
  `Key Insights` = c("Community size structure", "Population dynamics", "Ecosystem biomass",
                     "Reproductive success", "Environmental drivers", "Food web structure")
)

kable(vis_types,
      caption = "ZooMSS Visualization Capabilities") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = TRUE)
```

# ACEAS Regional Modeling {#sec-aceas}

## Project Overview

The **Antarctic Climate & Ecosystems CRC (ACEAS)** regional modeling project focuses on understanding climate impacts on East Antarctic ecosystems through high-resolution modeling approaches.

::: {.column-margin}
**ACEAS** represents a collaborative effort to understand how climate change affects Antarctic marine ecosystems, with particular focus on sea ice dynamics and ecosystem responses.
:::

### Objectives

- Develop high-resolution regional ocean-biogeochemical models
- Assess climate change impacts on Antarctic marine ecosystems  
- Integrate sea ice dynamics with ecosystem processes
- Provide projections for policy and conservation planning

### Methodology

The ACEAS approach combines:

- **Regional ocean modeling** with enhanced spatial resolution
- **Biogeochemical cycle simulation** including nutrient dynamics
- **Sea ice-ecosystem coupling** for realistic environmental forcing
- **Multi-scenario analysis** under different climate pathways

# Project Roadmap & Timeline {#sec-roadmap}

## Time Allocation Strategy

The dual-project structure requires careful time management and strategic coordination:

```{r time-allocation}
#| echo: false
#| fig-cap: "Monthly time allocation between ZooMSS and ACEAS projects"

months <- c("Oct 2024", "Nov 2024", "Dec 2024", "Jan 2025", "Feb 2025", "Mar 2025", 
            "Apr 2025", "May 2025", "Jun 2025", "Jul 2025", "Aug 2025", "Sep 2025")

zoomss_time <- c(70, 80, 90, 60, 50, 85, 60, 50, 40, 60, 70, 60)
aceas_time <- c(30, 20, 10, 40, 50, 15, 40, 50, 60, 40, 30, 40)

time_df <- data.frame(
  Month = factor(months, levels = months),
  ZooMSS = zoomss_time,
  ACEAS = aceas_time
)

time_long <- time_df %>%
  tidyr::pivot_longer(cols = c(ZooMSS, ACEAS), 
                      names_to = "Project", 
                      values_to = "Percentage")

ggplot(time_long, aes(x = Month, y = Percentage, fill = Project)) +
  geom_col(position = "stack", alpha = 0.8) +
  scale_fill_manual(values = c("ZooMSS" = "#2E86AB", "ACEAS" = "#A23B72")) +
  labs(x = "Month", y = "Time Allocation (%)", 
       title = "Project Time Allocation Strategy") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14, face = "bold"),
        legend.position = "bottom") +
  geom_hline(yintercept = 50, linetype = "dashed", alpha = 0.5)
```

## Key Milestones {#sec-milestones}

### Short-term (October 2024 - March 2025)

::: {.callout-important}
## Critical Deliverables

**December 2024**: UQ Lab Group Presentation
- Showcase fish reproduction implementation
- Demonstrate extended testing framework  
- Present validation results across environmental gradients
:::

1. **Complete fish reproduction validation** (October-November 2024)
2. **Prepare UQ presentation materials** (November-December 2024)  
3. **Initiate ACEAS regional model setup** (January-February 2025)
4. **Begin long-term scenario preparations** (February-March 2025)

### Medium-term (April 2025 - September 2025)

1. **ACEAS Phase 1 completion** - Regional model calibration and validation
2. **ZooMSS long-term scenarios** - Simulations extending to 2300
3. **Fishing scenario implementation** - Preparation for FishMIP protocols
4. **CMIP workshop preparation** - Materials and analysis for March 2026

### Long-term (October 2025 - March 2026)

::: {.callout-tip}
## Major Presentation Opportunity

**March 2026**: CMIP Workshop
- Present ZooMSS with full fish reproduction capabilities
- Demonstrate FishMIP-ready fishing scenarios
- Showcase long-term projection capabilities to 2300
:::

1. **ACEAS Phase 2 implementation** - Advanced regional scenarios
2. **ZooMSS-FishMIP integration** - Full protocol implementation  
3. **Manuscript preparation** - Multiple publications from both projects
4. **CMIP workshop presentation** - International model intercomparison showcase

## Risk Assessment & Mitigation

```{r risk-assessment}
#| echo: false
risks <- data.frame(
  Risk = c("Fish reproduction bugs", "ACEAS model instability", "Time management conflicts", 
           "CMIP deadline pressure", "Computing resource limitations"),
  Impact = c("High", "Medium", "High", "Medium", "Low"),
  Probability = c("Low", "Medium", "Medium", "Low", "Medium"),
  Mitigation = c("Extensive testing framework", "Incremental development approach", 
                 "Clear time allocation strategy", "Early preparation timeline",
                 "Cloud computing backup plan")
)

kable(risks,
      caption = "Project Risk Assessment Matrix") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(2, background = ifelse(risks$Impact == "High", "#ffcccc", 
                                   ifelse(risks$Impact == "Medium", "#fff2cc", "#ccffcc"))) %>%
  column_spec(3, background = ifelse(risks$Probability == "High", "#ffcccc", 
                                   ifelse(risks$Probability == "Medium", "#fff2cc", "#ccffcc")))
```

# Technical Implementation Details {#sec-technical}

## ZooMSS Architecture Enhancements

The fish reproduction implementation required significant architectural changes:

```{r arch-diagram}
#| echo: false
#| fig-cap: "ZooMSS model architecture with reproduction module"
#| fig-height: 6

# Create a conceptual architecture diagram
arch_data <- data.frame(
  Component = c("Environmental Forcing", "Size-Structured Dynamics", "Feeding Interactions", 
                "Growth & Mortality", "Fish Reproduction", "Population Feedback"),
  Level = c(1, 2, 2, 3, 3, 4),
  x = c(3, 1, 5, 1, 5, 3),
  y = c(4, 3, 3, 2, 2, 1),
  Type = c("Input", "Core", "Core", "Process", "Process", "Output")
)

ggplot(arch_data, aes(x = x, y = y, label = Component, fill = Type)) +
  geom_point(size = 20, alpha = 0.7, shape = 21, stroke = 2) +
  geom_text(size = 3, fontface = "bold") +
  scale_fill_viridis_d(option = "plasma", alpha = 0.7) +
  xlim(0, 6) + ylim(0, 5) +
  theme_void() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  labs(title = "ZooMSS Model Architecture") +
  geom_segment(aes(x = 3, y = 3.5, xend = 1, yend = 3.5), 
               arrow = arrow(length = unit(0.3, "cm")), 
               linewidth = 1, alpha = 0.6, inherit.aes = FALSE) +
  geom_segment(aes(x = 3, y = 3.5, xend = 5, yend = 3.5), 
               arrow = arrow(length = unit(0.3, "cm")), 
               linewidth = 1, alpha = 0.6, inherit.aes = FALSE)
```

## Computational Performance

The extended testing framework requires significant computational resources:

- **500-year simulations**: ~45 minutes per run on standard hardware
- **Gradient testing**: Multiple scenarios across environmental conditions  
- **Memory optimization**: Efficient storage of time series data
- **Parallel processing**: Multiple scenario execution capabilities

::: {.column-margin}
Current testing uses chlorophyll gradients with three levels: 0.01 mg/m³ (oligotrophic), 1.0 mg/m³ (mesotrophic), and 10.0 mg/m³ (eutrophic).
:::

# Future Directions {#sec-future}

## ZooMSS Evolution

### Immediate Priorities (Next 6 months)

1. **Fishing mortality implementation** for FishMIP compliance
2. **Economic scenario coupling** for policy relevance  
3. **Model intercomparison** with other FishMIP participants
4. **Long-term stability** assessment through 2300

### Long-term Vision (1-2 years)

- **Regional implementations** for specific marine ecosystems
- **Climate coupling** with Earth System Models
- **Machine learning integration** for parameter optimization
- **Uncertainty quantification** frameworks

## ACEAS Development

### Phase 1 Objectives

- High-resolution regional ocean model for East Antarctica
- Validation against observational data
- Baseline ecosystem state characterization

### Phase 2 Expansion  

- Multi-scenario climate projections
- Ecosystem service quantification
- Policy-relevant impact assessments
- Integration with global Earth System Models

# Conclusions {#sec-conclusions}

The dual-project approach represents a strategic investment in both **fundamental marine ecosystem modeling** (ZooMSS) and **regional climate impact assessment** (ACEAS). The implementation of fish reproduction in ZooMSS marks a critical milestone enabling participation in international model intercomparison projects, while the ACEAS work provides essential regional perspectives on Antarctic ecosystem changes.

::: {.callout-note}
## Success Metrics

**ZooMSS**: Successful FishMIP integration, stable long-term projections, international recognition at CMIP workshop

**ACEAS**: Validated regional model, policy-relevant projections, contribution to Antarctic science community

**Integration**: Balanced time allocation, complementary skill development, enhanced research profile
:::

The roadmap balances immediate technical requirements with long-term scientific objectives, positioning both projects for maximum impact in marine ecosystem science and climate change research.

---

## Appendices {.appendix}

### A. Technical Specifications

**ZooMSS Model Configuration:**
- 12 functional groups (9 zooplankton, 3 fish)
- 191 size classes spanning 6 orders of magnitude
- Monthly time steps with sub-daily environmental forcing
- Reproduction dynamics with seasonal spawning cycles

**ACEAS Model Requirements:**
- High-resolution regional domain (East Antarctica)
- Coupled ocean-biogeochemical-sea ice system
- Multiple climate scenario ensemble
- Observational data validation framework

### B. Publication Strategy

**Planned Manuscripts:**

1. "Fish reproduction in size-spectrum marine ecosystem models: Implementation and validation in ZooMSS" - *Target: Ecological Modelling*
2. "Long-term marine ecosystem projections under climate change: A ZooMSS perspective to 2300" - *Target: Global Change Biology*  
3. "Regional impacts of climate change on East Antarctic marine ecosystems" - *Target: Antarctic Science*

### C. Collaboration Network

**ZooMSS Collaborations:**
- FishMIP international network
- University of Queensland research group
- CMIP modeling community

**ACEAS Partnerships:**
- Antarctic Climate & Ecosystems CRC
- Australian Antarctic Division
- International Antarctic research community
