---
title: "Fishing Dynamics and Ecosystem Effects in ZooMSS"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Fishing Dynamics and Ecosystem Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r message=FALSE, warning=FALSE}
library(zoomss)
library(ggplot2)
library(patchwork)
library(dplyr)
```

## Introduction

This vignette explores how effort-driven fishing mortality affects the ZooMSS marine ecosystem over multi-decadal timescales. We examine:

1. **Trophic cascades** — how removing fish biomass propagates through zooplankton communities
2. **Size spectra distortion** — how fishing steepens the community size spectrum
3. **Stock–recruitment dynamics** — how SSB and recruitment respond under sustained exploitation
4. **Fishing intensity gradients** — equilibrium responses across a range of effort levels
5. **Pulse fishing and recovery** — ecosystem resilience after fishing cessation

Throughout, we use the effort-driven fishing framework where fishing mortality is calculated dynamically as $F(w,t) = \text{Effort}(t) \times q \times \text{Selectivity}(w)$.

## Helper Functions

We define a few helper functions used throughout the vignette for extracting and plotting fishing-specific outputs.

```{r}
# Extract catch data frame from model output
extract_catch <- function(mdl, fish_names = c("Fish_Small", "Fish_Med", "Fish_Large")) {
  data.frame(
    time = rep(mdl$time, times = length(fish_names)),
    catch = as.vector(mdl$catch),
    group = rep(fish_names, each = length(mdl$time))
  )
}

# Extract total biomass per group over time
extract_biomass_ts <- function(mdl) {
  species <- mdl$param$Groups$Species
  ngrps <- mdl$param$ngrps
  nsave <- length(mdl$time)
  
  # Sum biomass across size classes for each group
  bm <- matrix(NA, nrow = nsave, ncol = ngrps)
  for (g in 1:ngrps) {
    bm[, g] <- rowSums(mdl$biomass[g, , drop = FALSE] *
                          matrix(mdl$param$w, nrow = 1)[rep(1, nsave), ] *
                          mdl$param$dx)
  }
  # Actually use the abundance array directly
  bm <- matrix(NA, nrow = nsave, ncol = ngrps)
  for (g in 1:ngrps) {
    bm[, g] <- apply(mdl$abundance[, g, , drop = FALSE], 1, function(row) {
      sum(row * mdl$param$w * mdl$param$dx)
    })
  }
  
  df <- data.frame(
    time = rep(mdl$time, times = ngrps),
    biomass = as.vector(bm),
    species = rep(species, each = nsave),
    type = rep(mdl$param$Groups$Type, each = nsave)
  )
  df$species <- factor(df$species, levels = species)
  df
}

# Get consistent fish colours
get_fish_colours <- function(mdl) {
  fc <- mdl$param$Groups$PlotColour[mdl$param$fish_grps]
  names(fc) <- mdl$param$Groups$Species[mdl$param$fish_grps]
  fc
}

# Get all group colours
get_all_colours <- function(mdl) {
  ac <- mdl$param$Groups$PlotColour
  names(ac) <- mdl$param$Groups$Species
  ac
}
```

## Part 1: Unfished Baseline

First, we establish an unfished equilibrium as a reference. We run for 200 years with constant environment to reach steady state.

```{r}
Groups <- getGroups()

env_base <- createInputParams(
  time = seq(0, 200, by = 0.1),
  sst = 15,
  chl = 1.0
)

mdl_unfished <- zoomss_model(input_params = env_base, Groups = Groups, isave = 5)
```

```{r fig.height=8}
p1 <- plotTimeSeries(mdl_unfished, by = "biomass", transform = "log10") +
  ggtitle("Unfished Baseline — All Groups")

p2 <- plotReproduction(mdl_unfished, by = "SSB") +
  ggtitle("Unfished SSB")

p3 <- plotReproduction(mdl_unfished, by = "recruitment") +
  ggtitle("Unfished Recruitment")

wrap_plots(p1, p2, p3, ncol = 1)
```

## Part 2: Constant Fishing — Trophic Cascades

We now apply constant fishing across all three fish groups and observe how the ecosystem responds. We use moderate catchability values and compare against the unfished baseline.

```{r}
Groups_fished <- setFishingParams(Groups,
                                  q_small = 0.01,
                                  q_med   = 0.02,
                                  q_large = 0.03)

env_fished <- createInputParams(
  time = seq(0, 200, by = 0.1),
  sst = 15,
  chl = 1.0
)

env_fished <- addFishingEffort(env_fished,
                               effort_small = 1.0,
                               effort_med   = 1.0,
                               effort_large = 1.0)

mdl_fished <- zoomss_model(input_params = env_fished, Groups = Groups_fished, isave = 5)
```

### Fish Biomass: Fished vs Unfished

```{r fig.height=6}
fish_names <- c("Fish_Small", "Fish_Med", "Fish_Large")

p_unfished <- plotTimeSeries(mdl_unfished, by = "biomass",
                             species = fish_names, transform = "log10") +
  ggtitle("Unfished — Fish Biomass") +
  xlim(50, 200)

p_fished <- plotTimeSeries(mdl_fished, by = "biomass",
                           species = fish_names, transform = "log10") +
  ggtitle("Fished — Fish Biomass") +
  xlim(50, 200)

wrap_plots(p_unfished, p_fished, ncol = 1, guides = "collect")
```

### Catch Time Series

```{r}
fish_colours <- get_fish_colours(mdl_fished)

catch_df <- extract_catch(mdl_fished)

ggplot(catch_df, aes(x = time, y = catch, colour = group)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Time (years)", y = "Catch (g WW)",
       colour = "Fish Group", title = "Predicted Catch Under Constant Fishing")
```

### Community Composition: Trophic Cascade Effects

Removing fish biomass through fishing releases predation pressure on zooplankton, potentially allowing zooplankton biomass to increase — a trophic cascade. We compare community composition between fished and unfished scenarios.

```{r fig.height=6}
p_unfished_fill <- plotTimeSeries(mdl_unfished, by = "biomass", type = "fill") +
  ggtitle("Unfished — Community Composition") +
  xlim(100, 200)

p_fished_fill <- plotTimeSeries(mdl_fished, by = "biomass", type = "fill") +
  ggtitle("Fished — Community Composition") +
  xlim(100, 200)

wrap_plots(p_unfished_fill, p_fished_fill, ncol = 1, guides = "collect")
```

### Size Spectra Distortion

Fishing selectively removes larger individuals, steepening the community size spectrum relative to the unfished state.

```{r fig.height=6}
p_spec_unfished <- plotSizeSpectra(mdl_unfished, n_years = 10) +
  ggtitle("Unfished — Size Spectrum")

p_spec_fished <- plotSizeSpectra(mdl_fished, n_years = 10) +
  ggtitle("Fished — Size Spectrum")

wrap_plots(p_spec_unfished, p_spec_fished, ncol = 1, guides = "collect")
```

## Part 3: Stock–Recruitment Dynamics Under Fishing

Fishing reduces adult biomass, which reduces SSB, which reduces reproductive output and recruitment. This creates a feedback loop that determines the equilibrium population size under exploitation.

```{r fig.height=9}
p_ssb_comp <- wrap_plots(
  plotReproduction(mdl_unfished, by = "SSB") +
    ggtitle("Unfished — SSB") + xlim(50, 200),
  plotReproduction(mdl_fished, by = "SSB") +
    ggtitle("Fished — SSB") + xlim(50, 200),
  ncol = 1, guides = "collect"
)

p_rec_comp <- wrap_plots(
  plotReproduction(mdl_unfished, by = "recruitment") +
    ggtitle("Unfished — Recruitment") + xlim(50, 200),
  plotReproduction(mdl_fished, by = "recruitment") +
    ggtitle("Fished — Recruitment") + xlim(50, 200),
  ncol = 1, guides = "collect"
)

wrap_plots(p_ssb_comp, p_rec_comp, ncol = 2)
```

### Reproductive Investment by Size Under Fishing

Fishing removes the larger, more fecund individuals, shifting where reproductive investment peaks in the size spectrum.

```{r fig.height=6}
p_repro_unfished <- plotReproSizeSpectra(mdl_unfished, n_years = 10) +
  ggtitle("Unfished — Reproductive Rate by Size")

p_repro_fished <- plotReproSizeSpectra(mdl_fished, n_years = 10) +
  ggtitle("Fished — Reproductive Rate by Size")

wrap_plots(p_repro_unfished, p_repro_fished, ncol = 1, guides = "collect")
```

## Part 4: Fishing Intensity Gradient

To understand how the ecosystem responds across a range of fishing pressures, we run a series of simulations with increasing effort multipliers. This reveals the shape of the yield–effort curve and identifies when stock collapse becomes likely.

```{r}
effort_levels <- c(0, 0.25, 0.5, 1.0, 1.5, 2.0, 3.0, 5.0)

Groups_grad <- setFishingParams(Groups,
                                q_small = 1,
                                q_med   = 1,
                                q_large = 1)

results_effort <- list()

for (i in seq_along(effort_levels)) {
  env_eff <- createInputParams(
    time = seq(0, 200, by = 0.1),
    sst = 15,
    chl = 1.0
  )
  
  env_eff <- addFishingEffort(env_eff,
                              effort_small = effort_levels[i],
                              effort_med   = effort_levels[i],
                              effort_large = effort_levels[i])
  
  results_effort[[i]] <- zoomss_model(
    input_params = env_eff,
    Groups = Groups_grad,
    isave = 5
  )
}
```

### Equilibrium Yield, Biomass, SSB and Recruitment vs Effort

We extract equilibrium values (averaged over the final 20 years) for each effort level.

```{r fig.height=10}
num_fish <- 3
fish_names <- c("Fish_Small", "Fish_Med", "Fish_Large")

effort_summary <- data.frame(
  effort = rep(effort_levels, each = num_fish),
  fish_group = rep(fish_names, length(effort_levels)),
  eq_catch = NA,
  eq_biomass = NA,
  eq_SSB = NA,
  eq_recruitment = NA
)

for (i in seq_along(effort_levels)) {
  mdl_i <- results_effort[[i]]
  nsave <- length(mdl_i$time)
  fish_grps <- mdl_i$param$fish_grps
  
  # Average over final 20% of time series
  final_idx <- max(1, floor(nsave * 0.8)):nsave
  
  for (f in 1:num_fish) {
    row_idx <- (i - 1) * num_fish + f
    fg <- fish_grps[f]
    
    # Equilibrium biomass (from abundance)
    avg_biomass <- mean(apply(mdl_i$abundance[final_idx, fg, , drop = FALSE], 1, function(row) {
      sum(row * mdl_i$param$w * mdl_i$param$dx)
    }))
    effort_summary$eq_biomass[row_idx] <- avg_biomass
    
    # Equilibrium SSB and recruitment
    effort_summary$eq_SSB[row_idx] <- mean(mdl_i$SSB[final_idx, f], na.rm = TRUE)
    effort_summary$eq_recruitment[row_idx] <- mean(mdl_i$recruitment[final_idx, f], na.rm = TRUE)
    
    # Equilibrium catch
    if (!is.null(mdl_i$catch)) {
      effort_summary$eq_catch[row_idx] <- mean(mdl_i$catch[final_idx, f], na.rm = TRUE)
    } else {
      effort_summary$eq_catch[row_idx] <- 0
    }
  }
}

fish_colours <- get_fish_colours(results_effort[[1]])

# Yield–Effort curve
p_yield <- ggplot(effort_summary, aes(x = effort, y = eq_catch, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Effort Multiplier", y = "Equilibrium Catch (g WW)",
       title = "Yield–Effort Curve", colour = "Fish Group")

# Biomass depletion
p_depletion <- ggplot(effort_summary, aes(x = effort, y = eq_biomass, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Effort Multiplier", y = "Equilibrium Biomass",
       title = "Biomass Depletion Under Increasing Effort", colour = "Fish Group")

# SSB depletion
p_ssb_depl <- ggplot(effort_summary, aes(x = effort, y = eq_SSB, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Effort Multiplier", y = "Equilibrium SSB",
       title = "SSB Depletion Under Increasing Effort", colour = "Fish Group")

# Recruitment response
p_rec_grad <- ggplot(effort_summary, aes(x = effort, y = eq_recruitment, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Effort Multiplier", y = "Equilibrium Recruitment",
       title = "Recruitment Response to Fishing Intensity", colour = "Fish Group")

wrap_plots(p_yield, p_depletion, p_ssb_depl, p_rec_grad, ncol = 1, guides = "collect")
```

The yield–effort curve is expected to show a dome shape for each fish group: at low effort, increasing fishing increases catch; beyond some optimum, overfishing reduces the stock so severely that total catch declines. The effort level at the peak approximates Maximum Sustainable Yield (MSY) for each group.

### Community Composition Across the Effort Gradient

How does the relative dominance of zooplankton vs fish change with increasing fishing pressure?

```{r fig.height=6}
# Extract equilibrium proportional composition at each effort level
comp_data <- data.frame()

for (i in seq_along(effort_levels)) {
  mdl_i <- results_effort[[i]]
  nsave <- length(mdl_i$time)
  final_idx <- max(1, floor(nsave * 0.8)):nsave
  ngrps <- mdl_i$param$ngrps
  species <- mdl_i$param$Groups$Species
  
  group_biomass <- numeric(ngrps)
  for (g in 1:ngrps) {
    group_biomass[g] <- mean(apply(
      mdl_i$abundance[final_idx, g, , drop = FALSE], 1, function(row) {
        sum(row * mdl_i$param$w * mdl_i$param$dx)
      }))
  }
  
  comp_data <- rbind(comp_data, data.frame(
    effort = effort_levels[i],
    species = species,
    biomass = group_biomass,
    type = mdl_i$param$Groups$Type
  ))
}

# Calculate proportions
comp_data <- comp_data %>%
  group_by(effort) %>%
  mutate(proportion = biomass / sum(biomass, na.rm = TRUE)) %>%
  ungroup()

comp_data$species <- factor(comp_data$species, levels = results_effort[[1]]$param$Groups$Species)

all_colours <- get_all_colours(results_effort[[1]])

ggplot(comp_data, aes(x = factor(effort), y = proportion, fill = species)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = all_colours) +
  theme_bw() +
  labs(x = "Effort Multiplier", y = "Proportion of Total Biomass",
       title = "Community Composition Across Fishing Intensity Gradient",
       fill = "Species")
```

At higher effort levels, fish biomass becomes a smaller fraction of the total community. Zooplankton groups may increase in relative (and possibly absolute) abundance as predation pressure from fish is released.

## Part 5: Selective Fishing — Targeting Large Fish Only

In practice, fishing is often size-selective. What happens if only the largest fish are targeted?

```{r}
Groups_large_only <- setFishingParams(Groups,
                                      q_small = 0,
                                      q_med   = 0,
                                      q_large = 0.05)

env_large <- createInputParams(
  time = seq(0, 200, by = 0.1),
  sst = 15,
  chl = 1.0
)

env_large <- addFishingEffort(env_large,
                              effort_small = 1.0,
                              effort_med   = 1.0,
                              effort_large = 1.0)

mdl_large_only <- zoomss_model(input_params = env_large, Groups = Groups_large_only, isave = 5)
```

```{r fig.height=9}
p_bm_all <- plotTimeSeries(mdl_large_only, by = "biomass", transform = "log10") +
  ggtitle("Large Fish Only Fishing — All Groups") +
  xlim(50, 200)

p_bm_fish <- plotTimeSeries(mdl_large_only, by = "biomass",
                            species = fish_names, transform = "log10") +
  ggtitle("Large Fish Only — Fish Biomass") +
  xlim(50, 200)

# Catch (only large fish have non-zero catch)
catch_large <- extract_catch(mdl_large_only)
p_catch_large <- ggplot(catch_large, aes(x = time, y = catch, colour = group)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Time (years)", y = "Catch (g WW)",
       colour = "Fish Group", title = "Catch — Large Fish Only")

wrap_plots(p_bm_all, p_bm_fish, p_catch_large, ncol = 1, guides = "collect")
```

Removing only the top predator (Fish_Large) may release medium and small fish from predation, producing a mesopredator release effect that propagates down through the food web.

## Part 6: Pulse Fishing and Recovery

How quickly does the ecosystem recover after fishing stops? We simulate a fishing period followed by a recovery period to assess resilience.

```{r}
n_total <- 3001  # 300 years at dt = 0.1
time_vec <- seq(0, 300, by = 0.1)

# Fish for years 50-150, then stop
effort_pulse <- ifelse(time_vec >= 50 & time_vec <= 150, 1.0, 0.0)

Groups_pulse <- setFishingParams(Groups,
                                 q_small = 0.015,
                                 q_med   = 0.025,
                                 q_large = 0.04)

env_pulse <- createInputParams(time = time_vec, sst = 15, chl = 1.0)
env_pulse <- addFishingEffort(env_pulse,
                              effort_small = effort_pulse,
                              effort_med   = effort_pulse,
                              effort_large = effort_pulse)

mdl_pulse <- zoomss_model(input_params = env_pulse, Groups = Groups_pulse, isave = 5)
```

```{r fig.height=9}
# Effort time series for context
effort_df <- data.frame(time = mdl_pulse$time,
                        effort = approx(time_vec, effort_pulse, xout = mdl_pulse$time)$y)

p_effort <- ggplot(effort_df, aes(x = time, y = effort)) +
  geom_area(fill = "grey80", alpha = 0.7) +
  geom_line(linewidth = 0.5) +
  theme_bw() +
  labs(x = "Time (years)", y = "Effort", title = "Fishing Effort (Pulse)")

p_bm_pulse <- plotTimeSeries(mdl_pulse, by = "biomass",
                             species = fish_names, transform = "log10") +
  ggtitle("Fish Biomass — Pulse Fishing Experiment")

p_ssb_pulse <- plotReproduction(mdl_pulse, by = "SSB") +
  ggtitle("SSB — Pulse Fishing")

p_rec_pulse <- plotReproduction(mdl_pulse, by = "recruitment") +
  ggtitle("Recruitment — Pulse Fishing")

wrap_plots(p_effort, p_bm_pulse, p_ssb_pulse, p_rec_pulse, ncol = 1)
```

Key questions to examine:
- How quickly does biomass decline once fishing starts?
- Does recovery begin immediately when fishing stops, or is there a lag?
- Does the ecosystem return to its unfished state, or does it reach a different equilibrium?
- Which fish groups recover fastest — small, medium, or large?

### Community Composition: Before, During, and After Fishing

```{r fig.height=5}
p_fill_pulse <- plotTimeSeries(mdl_pulse, by = "biomass", type = "fill") +
  geom_vline(xintercept = c(50, 150), linetype = "dashed", colour = "black") +
  annotate("text", x = 25, y = 0.95, label = "Pre-fishing", size = 3) +
  annotate("text", x = 100, y = 0.95, label = "Fishing", size = 3) +
  annotate("text", x = 225, y = 0.95, label = "Recovery", size = 3) +
  ggtitle("Community Composition — Pulse Fishing Experiment")

p_fill_pulse
```

## Part 7: Ramping Effort (Historical Fishing Expansion)

A more realistic scenario: fishing effort increases gradually over time, mimicking the historical expansion of industrial fishing.

```{r}
time_ramp <- seq(0, 200, by = 0.1)
n_ramp <- length(time_ramp)

# No fishing for first 50 years (spin-up), then linear ramp from 0 to 3
effort_ramp <- pmax(0, (time_ramp - 50) / 50)  # reaches 1.0 at year 100, 3.0 at year 200

Groups_ramp <- setFishingParams(Groups,
                                q_small = 0.01,
                                q_med   = 0.02,
                                q_large = 0.03)

env_ramp <- createInputParams(time = time_ramp, sst = 15, chl = 1.0)
env_ramp <- addFishingEffort(env_ramp,
                             effort_small = effort_ramp,
                             effort_med   = effort_ramp,
                             effort_large = effort_ramp)

mdl_ramp <- zoomss_model(input_params = env_ramp, Groups = Groups_ramp, isave = 5)
```

```{r fig.height=12}
# Effort ramp
effort_ramp_df <- data.frame(
  time = mdl_ramp$time,
  effort = approx(time_ramp, effort_ramp, xout = mdl_ramp$time)$y
)

p_eff_ramp <- ggplot(effort_ramp_df, aes(x = time, y = effort)) +
  geom_area(fill = "grey80", alpha = 0.7) +
  geom_line(linewidth = 0.5) +
  theme_bw() +
  labs(x = "Time (years)", y = "Effort", title = "Ramping Effort Scenario")

p_bm_ramp <- plotTimeSeries(mdl_ramp, by = "biomass", transform = "log10") +
  ggtitle("All Groups — Biomass Under Ramping Effort")

# Catch
catch_ramp <- extract_catch(mdl_ramp)
p_catch_ramp <- ggplot(catch_ramp, aes(x = time, y = catch, colour = group)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Time (years)", y = "Catch (g WW)",
       colour = "Fish Group", title = "Catch Under Ramping Effort")

p_ssb_ramp <- plotReproduction(mdl_ramp, by = "SSB") +
  ggtitle("SSB Under Ramping Effort")

p_fill_ramp <- plotTimeSeries(mdl_ramp, by = "biomass", type = "fill") +
  ggtitle("Community Composition Under Ramping Effort")

wrap_plots(p_eff_ramp, p_bm_ramp, p_catch_ramp, p_ssb_ramp, p_fill_ramp, ncol = 1)
```

As effort ramps up, watch for:
- The point where catch for each group peaks and then declines (approaching or exceeding MSY)
- Progressive depletion of fish groups from largest to smallest
- Shifts in community composition toward zooplankton dominance

## Part 8: Catchability Sensitivity Analysis

To understand how catchability (q) influences equilibrium biomass and yield, we hold effort constant and vary q across a range of values for each size group.
This helps identify thresholds where increased catchability leads to disproportionate biomass decline or yield collapse.

```{r}
q_levels <- c(0.001, 0.005, 0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1)

results_q <- list()

for (i in seq_along(q_levels)) {
  Groups_q <- setFishingParams(Groups,
                               q_small = q_levels[i],
                               q_med   = q_levels[i],
                               q_large = q_levels[i])
  
  env_q <- createInputParams(
    time = seq(0, 200, by = 0.1),
    sst = 15,
    chl = 1.0
  )
  
  env_q <- addFishingEffort(env_q,
                            effort_small = 1.0,
                            effort_med   = 1.0,
                            effort_large = 1.0)
  
  results_q[[i]] <- zoomss_model(
    input_params = env_q,
    Groups = Groups_q,
    isave = 5
  )
}
```


### Equilibrium Yield, Biomass, SSB and Recruitment vs Catchability

We extract equilibrium values (averaged over the final 20% of the time series) for each catchability level — the same approach used for the effort gradient in Part 4, but now sweeping across $q$ while effort is fixed at 1.0. Since $F = q \times \text{Effort}$, this directly reveals the sensitivity of equilibrium outcomes to the catchability parameter estimated during calibration.

```{r fig.height=10}
num_fish <- 3
fish_names <- c("Fish_Small", "Fish_Med", "Fish_Large")

q_summary <- data.frame(
  q = rep(q_levels, each = num_fish),
  fish_group = rep(fish_names, length(q_levels)),
  eq_catch = NA,
  eq_biomass = NA,
  eq_SSB = NA,
  eq_recruitment = NA
)

for (i in seq_along(q_levels)) {
  mdl_i <- results_q[[i]]
  nsave <- length(mdl_i$time)
  fish_grps <- mdl_i$param$fish_grps

  # Average over final 20% of time series
  final_idx <- max(1, floor(nsave * 0.8)):nsave

  for (f in 1:num_fish) {
    row_idx <- (i - 1) * num_fish + f
    fg <- fish_grps[f]

    # Equilibrium biomass (from abundance)
    avg_biomass <- mean(apply(mdl_i$abundance[final_idx, fg, , drop = FALSE], 1, function(row) {
      sum(row * mdl_i$param$w * mdl_i$param$dx)
    }))
    q_summary$eq_biomass[row_idx] <- avg_biomass

    # Equilibrium SSB and recruitment
    q_summary$eq_SSB[row_idx] <- mean(mdl_i$SSB[final_idx, f], na.rm = TRUE)
    q_summary$eq_recruitment[row_idx] <- mean(mdl_i$recruitment[final_idx, f], na.rm = TRUE)

    # Equilibrium catch
    if (!is.null(mdl_i$catch)) {
      q_summary$eq_catch[row_idx] <- mean(mdl_i$catch[final_idx, f], na.rm = TRUE)
    } else {
      q_summary$eq_catch[row_idx] <- 0
    }
  }
}

fish_colours <- get_fish_colours(results_q[[1]])

# Yield-Catchability curve
p_yield_q <- ggplot(q_summary, aes(x = q, y = eq_catch, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = fish_colours) +
  scale_x_log10() +
  theme_bw() +
  labs(x = "Catchability (q)", y = "Equilibrium Catch (g WW)",
       title = "Yield-Catchability Curve", colour = "Fish Group")

# Biomass depletion vs q
p_biomass_q <- ggplot(q_summary, aes(x = q, y = eq_biomass, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = fish_colours) +
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Catchability (q)", y = "Equilibrium Biomass",
       title = "Biomass Depletion vs Catchability", colour = "Fish Group")

# SSB vs q
p_ssb_q <- ggplot(q_summary, aes(x = q, y = eq_SSB, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = fish_colours) +
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Catchability (q)", y = "Equilibrium SSB",
       title = "SSB vs Catchability", colour = "Fish Group")

# Recruitment vs q
p_rec_q <- ggplot(q_summary, aes(x = q, y = eq_recruitment, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = fish_colours) +
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Catchability (q)", y = "Equilibrium Recruitment",
       title = "Recruitment vs Catchability", colour = "Fish Group")

wrap_plots(p_yield_q, p_biomass_q, p_ssb_q, p_rec_q, ncol = 1, guides = "collect")
```

The yield-catchability curve should show a similar dome shape to the yield-effort curve: at low $q$ values, increasing catchability increases catch proportionally; beyond some critical threshold, stock depletion dominates and total yield declines. Because $q$ acts as a multiplier on effort, the $q$ at peak yield approximates $q_{\text{MSY}}$ when effort is held at 1.0.

Note the log-scaled x-axis — catchability spans over two orders of magnitude, and the nonlinear response highlights how sensitive equilibrium outcomes are to this parameter. This underscores why accurate $q$ estimation during calibration is critical.

### Biomass Depletion Ratio

To express the impact of catchability in relative terms, we normalise equilibrium biomass against the unfished reference (lowest $q$ level, which has negligible fishing mortality).

```{r fig.height=5}
# Calculate depletion ratio relative to lowest q (effectively unfished)
ref_biomass <- q_summary %>%
  filter(q == min(q_levels)) %>%
  select(fish_group, ref_biomass = eq_biomass)

q_depletion <- q_summary %>%
  left_join(ref_biomass, by = "fish_group") %>%
  mutate(depletion = eq_biomass / ref_biomass)

ggplot(q_depletion, aes(x = q, y = depletion, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = 0.2, linetype = "dotted", colour = "red") +
  annotate("text", x = max(q_levels), y = 0.52,
           label = "B/B0 = 0.5", hjust = 1, size = 3) +
  annotate("text", x = max(q_levels), y = 0.22,
           label = "B/B0 = 0.2", hjust = 1, size = 3, colour = "red") +
  scale_colour_manual(values = fish_colours) +
  scale_x_log10() +
  theme_bw() +
  labs(x = "Catchability (q)",
       y = expression(B / B[0]),
       title = "Biomass Depletion Ratio Across Catchability Gradient",
       colour = "Fish Group")
```

The dashed line at $B/B_0 = 0.5$ marks a common reference point for maximum sustainable yield. The dotted red line at $B/B_0 = 0.2$ indicates severely depleted stock levels. The $q$ value at which each fish group crosses these thresholds provides practical guidance on the range of catchability estimates that are ecologically sustainable.

### Size Spectra: Low vs High Catchability

Comparing size spectra at the extremes of the catchability gradient reveals how gear efficiency reshapes community structure.

```{r fig.height=6}
idx_low <- 1                  # lowest q
idx_high <- length(q_levels)  # highest q

p_spec_low_q <- plotSizeSpectra(results_q[[idx_low]], n_years = 10) +
  ggtitle(paste0("Low Catchability (q = ", q_levels[idx_low], ") - Size Spectrum"))

p_spec_high_q <- plotSizeSpectra(results_q[[idx_high]], n_years = 10) +
  ggtitle(paste0("High Catchability (q = ", q_levels[idx_high], ") - Size Spectrum"))

wrap_plots(p_spec_low_q, p_spec_high_q, ncol = 1, guides = "collect")
```

At low catchability the size spectrum retains its unfished shape with abundant large-bodied organisms. At high catchability, selective removal steepens the spectrum — large fish are severely depleted while smaller groups and zooplankton may increase due to competitive and predation release.


## Summary

This vignette demonstrated the ecological consequences of fishing in ZooMSS:

1. **Trophic cascades**: Fishing reduces fish biomass, which can release zooplankton from predation pressure, altering community composition throughout the size spectrum.

2. **Size spectrum distortion**: Fishing selectively removes biomass from the larger size classes, steepening the slope of the community size spectrum relative to the unfished state.

3. **Stock–recruitment feedbacks**: Reduced SSB under fishing leads to lower reproductive output and recruitment, creating a negative feedback that determines equilibrium population size.

4. **Yield–effort relationships**: The yield–effort curve shows the classic dome shape — moderate fishing maximises catch, while overfishing depletes stocks to the point where total catch declines.

5. **Selective fishing effects**: Targeting only large fish produces mesopredator release effects that propagate through the food web.

6. **Recovery dynamics**: After fishing cessation, ecosystem recovery is not instantaneous. Recovery timescales depend on the severity of depletion and the life-history characteristics of each fish group.

7. **Historical fishing expansion**: Gradually increasing effort reproduces the pattern of sequential depletion from large to small fish, consistent with "fishing down the food web" observations.

### Connection to FishMIP Calibration

The effort-driven framework demonstrated here is designed for calibration against observed catch data within the FishMIP (Fisheries and Marine Ecosystem Model Intercomparison Project) framework. The calibration process estimates catchability ($q$) and selectivity ($w_{min}$) by minimising the difference between predicted and observed catch:

$$\text{SSE} = \sum_t \left( \text{Catch}_{\text{predicted}}(t) - \text{Catch}_{\text{observed}}(t) \right)^2$$

See the calibration vignette (forthcoming) for details on parameter estimation using `calibrateFishing()`.

## Session Info

```{r}
sessionInfo()
```
