---
title: "Fish Reproduction and Energy Budget in ZooMSS"
vignette: >
  %\VignetteIndexEntry{Fish Reproduction and Energy Budget}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r message=FALSE, warning=FALSE}
library(zoomss)
library(ggplot2)
library(patchwork)
```

## Introduction

This vignette demonstrates the explicit energy budget and fish reproduction system in ZooMSS. The model now implements a DBPM-style (Dynamic Benthic-Pelagic Model) energy budget that explicitly tracks:
- **Defecation / non-nutritive fraction**: Energy lost as faeces plus water and structural material (varies with prey quality via Carbon content)
- **Assimilation**: Energy absorbed through the gut wall
- **Metabolism**: Energy spent on maintenance (f_M fraction)
- **Growth**: Energy allocated to somatic growth (K_growth fraction)
- **Reproduction**: Energy allocated to reproductive output (R_frac fraction)

For fish groups with `repro_on = 1`, reproductive investment drives population dynamics through explicit recruitment.

> **Terminology note:** The parameters `def_high` and `def_low` represent the **non-nutritive fraction** of ingested prey — encompassing water content, structural material (chitin, cellulose), and true fecal losses — rather than defecation in the strict physiological sense.

## The Energy Budget Structure

The energy budget follows a two-stage partitioning:

**Stage 1: Ingestion Partitioning**
$$
\text{Ingestion} = \text{Defecation} + \text{Assimilation}
$$

Where defecation varies continuously based on prey Carbon content:
$$
D_{prey} = D_{high} + (D_{low} - D_{high}) \times \left(1 - \frac{C_{prey}}{C_{max}}\right)
$$

This formulation interpolates between low defecation for high-quality (high Carbon) prey and high defecation for low-quality (low Carbon) prey. The approach is based on the Dynamic Benthic-Pelagic Model (DBPM; Blanchard et al. 2009, 2012). We retain `def_high = 0.30` from the DBPM (Peters, 1983) and extend `def_low` to **0.95** to capture the extreme water content of gelatinous prey (jellyfish are ~95% water, salps/larvaceans ~96%). Carbon:wet weight ratio serves as a continuous proxy for food quality.

**Stage 2: Assimilation Partitioning**
$$
\text{Assimilated} = \text{Metabolism} + \text{Growth} + \text{Reproduction}
$$

Where: $f_M + K + R_{frac} = 1$

## Model Setup

Let's load the default groups and examine the energy budget parameters:

```{r}
Groups <- getGroups()

# Display energy budget parameters
energy_params <- Groups[, c("Species", "Type", "def_high", "def_low", 
                            "f_M", "K_growth", "repro_on", "Carbon")]
print(energy_params)
```

Notice that:
- **Energy budget fractions are group-specific**: `f_M`, `K_growth`, and `R_frac = 1 - f_M - K_growth` vary across functional groups
- `K_growth` ranges from 0.25 (Euphausiids) to 0.50 (Jellyfish), reflecting different life-history strategies
- **Zooplankton** have `repro_on = 0` (all production goes to somatic growth)
- **Fish** have `repro_on = 1` (reproduction enabled), with `R_frac = 1 - f_M - K_growth` allocated to reproduction
- `def_low = 0.95` across all groups — gelatinous prey (jellyfish, salps) incur very high non-nutritive fractions

## Part 1: Constant Environmental Conditions

First, let's run the model with constant temperature and chlorophyll (1 mg m⁻³) to establish baseline reproduction dynamics. We run for 200 years to ensure the model reaches an oscillating steady-state.

```{r}
# Create constant environmental conditions
env_constant <- createInputParams(
  time = seq(0, 200, by = 0.1),
  sst = 15,
  chl = 1.0  # 1 mg/m³ chlorophyll
)

# Plot the environmental forcing
plotEnvironment(env_constant)
```

```{r}
# Run the model with constant environment
mdl_constant <- zoomss_model(
  input_params = env_constant, 
  Groups = Groups, 
  isave = 2
)
```

### Visualizing the Energy Budget

Before examining the model dynamics, let's visualize the energy budget structure.

#### Energy Budget Allocation

This plot shows how assimilated energy is partitioned into metabolism, growth, and reproduction for each functional group:

```{r fig.height=5}
plotEnergyBudget(mdl_constant)
```

Note that energy budget fractions are group-specific. For fish, the allocation depends on maturity: **immature fish** allocate all production to growth (K_growth + R_frac), while **mature fish** split production between growth (K_growth) and reproduction (R_frac). The transition is governed by a size-based maturity ogive.

#### Defecation by Prey Quality

This plot shows how defecation rate varies with prey Carbon content:

```{r fig.height=5}
plotDefecation(mdl_constant)
```

The non-nutritive fraction decreases linearly with increasing prey Carbon content. High-Carbon prey (like ciliates/flagellates at 0.15 g C/g WW) have low non-nutritive fraction (~30%), while low-Carbon prey (like jellyfish at 0.005 g C/g WW) have very high non-nutritive fraction (~94%). This wide range (0.30–0.95) captures the fundamental difference between carbon-dense crustacean prey and water-laden gelatinous prey.

#### Prey-Specific Growth Efficiency Matrix

This heatmap shows the net growth efficiency for each predator-prey combination:

```{r fig.height=6}
plotAssimilationMatrix(mdl_constant)
```

This matrix displays the growth efficiency for each predator-prey combination, calculated as $(1 - D_{prey}) \times K_{predator}$. Prey are ordered by Carbon content along the x-axis (low Carbon like Larvaceans and Salps on the left, high Carbon like Ciliates/Flagellates on the right). Higher values (brighter colours) indicate that a predator gains more somatic growth per unit prey consumed. Growth efficiency varies across groups because K_growth is group-specific (0.25–0.50).

### Biomass Dynamics

Let's examine the biomass dynamics over time:

```{r fig.height=9}
# Plot biomass time series for all groups
p1 <- plotTimeSeries(mdl_constant, by = "biomass", transform = "log10") +
  ggtitle("All Groups - Biomass Time Series")

# Focus on fish groups only
p2 <- plotTimeSeries(mdl_constant, by = "biomass", 
                     species = c("Fish_Small", "Fish_Med", "Fish_Large"),
                     transform = "log10") +
  ggtitle("Fish Groups - Biomass Time Series")

# Stacked biomass showing proportions
p3 <- plotTimeSeries(mdl_constant, by = "biomass", type = "fill") +
  ggtitle("Relative Biomass Proportions")

wrap_plots(p1, p2, p3, ncol = 1)
```

### Reproduction Metrics

Now let's examine the reproduction metrics for fish:

```{r fig.height=9}
# Spawning Stock Biomass (SSB)
p_ssb <- plotReproduction(mdl_constant, by = "SSB") +
  ggtitle("Spawning Stock Biomass (SSB)")

# Recruitment
p_recruit <- plotReproduction(mdl_constant, by = "recruitment") +
  ggtitle("Recruitment")

# Total reproductive output
p_repro <- plotReproduction(mdl_constant, by = "repro_output") +
  ggtitle("Total Reproductive Output")

wrap_plots(p_ssb, p_recruit, p_repro, ncol = 1)
```

### Growth Rate Comparison

Compare growth rates between fish and zooplankton:

```{r fig.height=6}
# Growth rates for zooplankton (higher K_growth = 0.50)
p_zoo <- plotGrowthComparison(mdl_constant, groups = "zooplankton") +
  ggtitle("Zooplankton Growth Rates")

# Growth rates for fish (lower K_growth = 0.36, with reproduction)
p_fish <- plotGrowthComparison(mdl_constant, groups = "fish") +
  ggtitle("Fish Growth Rates")

wrap_plots(p_zoo, p_fish, ncol = 1, guides = "collect")
```

### Reproductive Investment by Size

Examine how reproductive investment varies with body size:

```{r}
plotReproSizeSpectra(mdl_constant, n_years = 10) +
  ggtitle("Fish Reproductive Rate by Body Size")
```

The reproductive rate increases with body size within each fish group, reflecting:
1. The maturity ogive (larger individuals are more likely to be mature)
2. The body-size scaling of growth rates

### Size Spectra

```{r fig.height=6}
# Abundance size spectrum
p_abund <- plotSizeSpectra(mdl_constant, by = "abundance", n_years = 10) +
  ggtitle("Abundance Size Spectrum")

# Biomass size spectrum  
p_biom <- plotSizeSpectra(mdl_constant, by = "biomass", n_years = 10) +
  ggtitle("Biomass Size Spectrum")

wrap_plots(p_abund, p_biom, ncol = 1, guides = "collect")
```

### Food Web Diagnostics

#### Predator-Prey Mass Ratio (PPMR)

The PPMR describes the size relationship between predators and their prey. In ZooMSS, zooplankton PPMR varies with body size following Wirtz (2012), while fish have a fixed PPMR of 100. The community-level PPMR distribution, weighted by biomass, reveals the dominant feeding relationships in the food web.

```{r fig.height=5}
# Plot PPMR distribution at the final saved time step
final_idx <- length(mdl_constant$time)
plotPPMR(mdl_constant, idx = final_idx) +
  ggtitle("Community PPMR Distribution (Final Time Step)")
```

The vertical black line shows the biomass-weighted mean log₁₀ PPMR across the zooplankton community. Coloured lines show the PPMR profile for each zooplankton group across its size range.

#### Emergent Trophic Levels

Trophic levels emerge from the realised diet composition at each time step, calculated iteratively using the Gauss-Seidel method. Phytoplankton have TL = 1, and consumer trophic levels are computed as 1 + the biomass-weighted mean TL of their prey.

```{r fig.height=6}
# Extract trophic levels across all saved time steps
tl_matrix <- extractTrophicLevels(mdl_constant)

# Create a data frame for plotting
species_names <- mdl_constant$param$Groups$Species
plot_colours <- mdl_constant$param$Groups$PlotColour
names(plot_colours) <- species_names

tl_df <- data.frame(
  time = rep(mdl_constant$time, times = ncol(tl_matrix)),
  trophic_level = as.vector(tl_matrix),
  species = rep(species_names, each = nrow(tl_matrix))
)

# Time series of trophic levels
p_tl_ts <- ggplot(tl_df, aes(x = time, y = trophic_level, colour = species)) +
  geom_line(linewidth = 0.6) +
  scale_colour_manual(values = plot_colours) +
  theme_bw() +
  labs(x = "Time (years)", y = "Trophic Level",
       title = "Emergent Trophic Levels Over Time",
       colour = "Species")

# Bar chart of final trophic levels
final_tl <- tl_matrix[nrow(tl_matrix), ]
tl_final_df <- data.frame(
  species = factor(species_names, levels = species_names[order(final_tl)]),
  trophic_level = final_tl
)

p_tl_bar <- ggplot(tl_final_df, aes(x = species, y = trophic_level, fill = species)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = plot_colours) +
  coord_flip() +
  theme_bw() +
  labs(x = NULL, y = "Trophic Level",
       title = "Final Trophic Levels by Group") +
  geom_hline(yintercept = c(2, 3, 4), linetype = "dashed", alpha = 0.3)

wrap_plots(p_tl_ts, p_tl_bar, ncol = 1, heights = c(1.2, 1))
```

Herbivores and omnivores (flagellates, ciliates, omnivorous copepods) have trophic levels near 2, while carnivores (chaetognaths, fish) reach higher trophic levels reflecting secondary and tertiary consumption.

### Time-Averaged Biomass (Final 50 Years)

To assess the steady-state community structure, we compute time-averaged biomass over the final 50 years of the simulation using `getBiomass()`.

```{r fig.height=8}
# Get wet-weight biomass array (time x groups x sizes)
biomass_ww <- getBiomass(mdl_constant, units = "ww")

# Identify time indices for the final 50 years
all_times <- mdl_constant$time
final_50_idx <- which(all_times >= max(all_times) - 50)

# Average biomass over the final 50 years: sum across sizes for each group
biomass_final50 <- apply(biomass_ww[final_50_idx, , , drop = FALSE], c(2, 3), mean, na.rm = TRUE)
total_biomass_by_group <- rowSums(biomass_final50, na.rm = TRUE)

# Also get carbon biomass
biomass_c <- getBiomass(mdl_constant, units = "carbon")
biomass_c_final50 <- apply(biomass_c[final_50_idx, , , drop = FALSE], c(2, 3), mean, na.rm = TRUE)
total_carbon_by_group <- rowSums(biomass_c_final50, na.rm = TRUE)

# Build summary data frame
biomass_summary_df <- data.frame(
  species = factor(species_names, levels = species_names),
  type = mdl_constant$param$Groups$Type,
  biomass_ww = total_biomass_by_group,
  biomass_carbon = total_carbon_by_group
)

# Bar chart of total wet-weight biomass by group
p_bm_bar <- ggplot(biomass_summary_df, aes(x = species, y = biomass_ww, fill = species)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = plot_colours) +
  scale_y_log10() +
  coord_flip() +
  theme_bw() +
  labs(x = NULL, y = "Wet Weight Biomass (log scale)",
       title = "Time-Averaged Biomass by Group (Final 50 Years)")

# Stacked proportional biomass (zooplankton vs fish)
p_bm_prop <- ggplot(biomass_summary_df, aes(x = type, y = biomass_ww, fill = species)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = plot_colours) +
  theme_bw() +
  labs(x = NULL, y = "Wet Weight Biomass",
       title = "Biomass Composition: Zooplankton vs Fish",
       fill = "Species")

# Size spectrum averaged over last 50 years
biomass_spectrum_df <- data.frame(
  log10_weight = rep(log10(mdl_constant$param$w), each = length(species_names)),
  species = rep(species_names, times = length(mdl_constant$param$w)),
  biomass = as.vector(t(biomass_final50))
)
biomass_spectrum_df <- biomass_spectrum_df[biomass_spectrum_df$biomass > 0, ]

p_bm_spectrum <- ggplot(biomass_spectrum_df,
                        aes(x = log10_weight, y = biomass, colour = species)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = plot_colours) +
  scale_y_log10() +
  theme_bw() +
  labs(x = expression("log"[10] * " body weight (g)"),
       y = "Biomass density (log scale)",
       title = "Time-Averaged Biomass Size Spectrum (Final 50 Years)",
       colour = "Species")

wrap_plots(p_bm_bar, p_bm_prop, p_bm_spectrum, ncol = 1, heights = c(1, 0.8, 1))
```

```{r}
# Print summary table
cat("Time-averaged biomass summary (final 50 years):\n\n")
print(data.frame(
  Species = biomass_summary_df$species,
  Type = biomass_summary_df$type,
  WetWeight = signif(biomass_summary_df$biomass_ww, 4),
  Carbon = signif(biomass_summary_df$biomass_carbon, 4),
  Pct_WW = round(100 * biomass_summary_df$biomass_ww / sum(biomass_summary_df$biomass_ww), 1)
), row.names = FALSE)
```

## Part 2: Seasonal Environmental Forcing

Now let's compare with seasonal environmental conditions to see how reproduction responds to environmental variability. We use the same base chlorophyll (1 mg m⁻³) but with seasonal variation.

```{r}
# Create seasonal environmental conditions using createEnviroData
env_seasonal <- createEnviroData(
  n_years = 100,
  dt = 0.1,
  base_sst = 15,
  base_chl = 1.0,  # Same base as constant run
  seasonal = TRUE,
  sst_amplitude = 4,  # ±4°C seasonal variation
  chl_amplitude = 0.5  # ±0.5 mg/m³ seasonal variation
)

# Convert to input params format
env_seasonal_params <- createInputParams(
  time = env_seasonal$time,
  sst = env_seasonal$sst,
  chl = env_seasonal$chl
)

# Plot the seasonal environment
plotEnvironment(env_seasonal_params)
```

```{r}
# Run the model with seasonal environment
mdl_seasonal <- zoomss_model(
  input_params = env_seasonal_params, 
  Groups = Groups, 
  isave = 2
)
```

### Comparing Constant vs Seasonal: Biomass

```{r fig.height=9}
# Constant environment biomass
p_const <- plotTimeSeries(mdl_constant, by = "biomass", 
                          species = c("Fish_Small", "Fish_Med", "Fish_Large"),
                          transform = "log10") +
  ggtitle("Constant Environment - Fish Biomass") +
  xlim(50, 100)

# Seasonal environment biomass
p_seas <- plotTimeSeries(mdl_seasonal, by = "biomass", 
                         species = c("Fish_Small", "Fish_Med", "Fish_Large"),
                         transform = "log10") +
  ggtitle("Seasonal Environment - Fish Biomass") +
  xlim(50, 100)

# Show the seasonal forcing for context
p_env <- plotEnvironment(env_seasonal_params) & xlim(50, 100)

wrap_plots(p_const, p_seas, p_env, ncol = 1, heights = c(1, 1, 1))
```

### Comparing Constant vs Seasonal: Reproduction

```{r fig.height=9}
# SSB comparison
p_ssb_const <- plotReproduction(mdl_constant, by = "SSB") +
  ggtitle("Constant - SSB") +
  xlim(50, 100)

p_ssb_seas <- plotReproduction(mdl_seasonal, by = "SSB") +
  ggtitle("Seasonal - SSB") +
  xlim(50, 100)

# Recruitment comparison  
p_rec_const <- plotReproduction(mdl_constant, by = "recruitment") +
  ggtitle("Constant - Recruitment") +
  xlim(50, 100)

p_rec_seas <- plotReproduction(mdl_seasonal, by = "recruitment") +
  ggtitle("Seasonal - Recruitment") +
  xlim(50, 100)

wrap_plots(p_ssb_const, p_ssb_seas, p_rec_const, p_rec_seas, 
           ncol = 2, guides = "collect")
```

## Part 3: Exploring Chlorophyll Gradients

Let's explore how productivity (chlorophyll concentration) affects reproduction metrics and community composition across a wide gradient from oligotrophic (0.1 mg m⁻³) to eutrophic (10 mg m⁻³) conditions. We run each scenario for 100 years to ensure steady-state.

```{r}
# Run models across an extended chlorophyll gradient (0.1 to 10 mg/m³)
chl_levels <- c(0.1, 0.2, 0.5, 1.0, 2.0, 5.0, 10.0)
results_chl <- list()

for (i in seq_along(chl_levels)) {
  env_chl <- createInputParams(
    time = seq(0, 100, by = 0.1),
    sst = 15,
    chl = chl_levels[i]
  )
  
  results_chl[[i]] <- zoomss_model(
    input_params = env_chl,
    Groups = Groups,
    isave = 2
  )
}
```

### Fish Biomass, SSB and Recruitment Across the Gradient

```{r fig.height=10}
# Extract final biomass, SSB and recruitment values for each chlorophyll level
num_fish <- 3
fish_names <- c("Fish_Small", "Fish_Med", "Fish_Large")
fish_grps <- results_chl[[1]]$param$fish_grps

fish_summary <- data.frame(
  chl = rep(chl_levels, each = num_fish),
  fish_group = rep(fish_names, length(chl_levels)),
  final_biomass = NA,
  final_SSB = NA,
  final_recruitment = NA
)

for (i in seq_along(chl_levels)) {
  mdl <- results_chl[[i]]
  final_idx <- length(mdl$time)
  
  # Get average biomass over last 10 years
  avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 10)
  fish_biomass <- rowSums(avg_biomass[fish_grps, , drop = FALSE])
  
  for (f in 1:num_fish) {
    row_idx <- (i - 1) * num_fish + f
    fish_summary$final_biomass[row_idx] <- fish_biomass[f]
    fish_summary$final_SSB[row_idx] <- mdl$SSB[final_idx, f]
    fish_summary$final_recruitment[row_idx] <- mdl$recruitment[final_idx, f]
  }
}

# Get fish colours
fish_colours <- results_chl[[1]]$param$Groups$PlotColour[fish_grps]
names(fish_colours) <- fish_names

# Plot Fish Biomass vs chlorophyll
p_biomass <- ggplot(fish_summary, aes(x = chl, y = final_biomass, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll (mg m"^-3*")"),
       y = "Biomass",
       title = "Fish Biomass Response to Productivity Gradient",
       colour = "Fish Group")

# Plot SSB vs chlorophyll
p_ssb <- ggplot(fish_summary, aes(x = chl, y = final_SSB, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll (mg m"^-3*")"),
       y = "Spawning Stock Biomass (SSB)",
       title = "Fish SSB Response to Productivity Gradient",
       colour = "Fish Group")

# Plot Recruitment vs chlorophyll
p_recruit <- ggplot(fish_summary, aes(x = chl, y = final_recruitment, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll (mg m"^-3*")"),
       y = "Recruitment",
       title = "Fish Recruitment Response to Productivity Gradient",
       colour = "Fish Group")

wrap_plots(p_biomass, p_ssb, p_recruit, ncol = 1, guides = "collect")
```

### Community Composition Across the Gradient

Let's examine how the relative proportions of zooplankton and fish change across the productivity gradient.

```{r fig.height=10}
# Extract biomass for all groups at steady state (average last 10 years)
all_species <- results_chl[[1]]$param$Groups$Species
ngrps <- length(all_species)

biomass_summary <- data.frame(
  chl = rep(chl_levels, each = ngrps),
  species = rep(all_species, length(chl_levels)),
  biomass = NA,
  type = rep(results_chl[[1]]$param$Groups$Type, length(chl_levels))
)

for (i in seq_along(chl_levels)) {
  mdl <- results_chl[[i]]
  # Average biomass over last 10 years
  avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 10)
  total_biomass <- rowSums(avg_biomass)
  
  for (g in 1:ngrps) {
    row_idx <- (i - 1) * ngrps + g
    biomass_summary$biomass[row_idx] <- total_biomass[g]
  }
}

# Calculate proportions within each chl level
biomass_summary <- biomass_summary %>%
  dplyr::group_by(chl) %>%
  dplyr::mutate(proportion = biomass / sum(biomass, na.rm = TRUE)) %>%
  dplyr::ungroup()

# Set species factor levels for plotting
biomass_summary$species <- factor(biomass_summary$species, levels = all_species)

# Get colours from the model
plot_colours <- results_chl[[1]]$param$Groups$PlotColour
names(plot_colours) <- all_species

# Create stacked proportion plot
p_prop <- ggplot(biomass_summary, aes(x = factor(chl), y = proportion, fill = species)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = plot_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll (mg m"^-3*")"),
       y = "Proportion of Total Biomass",
       title = "Community Composition Across Productivity Gradient",
       fill = "Species") +
  theme(legend.position = "right")

# Also show absolute biomass
p_abs <- ggplot(biomass_summary, aes(x = chl, y = biomass, colour = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = plot_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll (mg m"^-3*")"),
       y = "Biomass",
       title = "Absolute Biomass Across Productivity Gradient",
       colour = "Species")

wrap_plots(p_prop, p_abs, ncol = 1, guides = "collect")
```

## Part 4: Temperature Effects

Explore how temperature affects fish reproduction through metabolic scaling. We run 100-year simulations at 1 mg m⁻³ chlorophyll across a temperature gradient.

```{r}
# Run models across a temperature gradient
sst_levels <- c(5, 10, 15, 20, 25)
results_sst <- list()

for (i in seq_along(sst_levels)) {
  env_sst <- createInputParams(
    time = seq(0, 100, by = 0.1),
    sst = sst_levels[i],
    chl = 1.0
  )
  
  results_sst[[i]] <- zoomss_model(
    input_params = env_sst,
    Groups = Groups,
    isave = 2
  )
}
```

### Fish Biomass, SSB and Recruitment Across the Temperature Gradient

```{r fig.height=10}
# Extract final biomass, SSB and recruitment values for each temperature level
num_fish <- 3
fish_names <- c("Fish_Small", "Fish_Med", "Fish_Large")
fish_grps <- results_sst[[1]]$param$fish_grps

fish_sst_summary <- data.frame(
  sst = rep(sst_levels, each = num_fish),
  fish_group = rep(fish_names, length(sst_levels)),
  final_biomass = NA,
  final_SSB = NA,
  final_recruitment = NA
)

for (i in seq_along(sst_levels)) {
  mdl <- results_sst[[i]]
  final_idx <- length(mdl$time)
  
  # Get average biomass over last 10 years
  avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 10)
  fish_biomass <- rowSums(avg_biomass[fish_grps, , drop = FALSE])
  
  for (f in 1:num_fish) {
    row_idx <- (i - 1) * num_fish + f
    fish_sst_summary$final_biomass[row_idx] <- fish_biomass[f]
    fish_sst_summary$final_SSB[row_idx] <- mdl$SSB[final_idx, f]
    fish_sst_summary$final_recruitment[row_idx] <- mdl$recruitment[final_idx, f]
  }
}

# Get fish colours
fish_colours <- results_sst[[1]]$param$Groups$PlotColour[fish_grps]
names(fish_colours) <- fish_names

# Plot Fish Biomass vs temperature
p_biomass <- ggplot(fish_sst_summary, aes(x = sst, y = final_biomass, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (°C)",
       y = "Biomass",
       title = "Fish Biomass Response to Temperature",
       colour = "Fish Group")

# Plot SSB vs temperature
p_ssb <- ggplot(fish_sst_summary, aes(x = sst, y = final_SSB, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (°C)",
       y = "Spawning Stock Biomass (SSB)",
       title = "Fish SSB Response to Temperature",
       colour = "Fish Group")

# Plot Recruitment vs temperature
p_recruit <- ggplot(fish_sst_summary, aes(x = sst, y = final_recruitment, colour = fish_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (°C)",
       y = "Recruitment",
       title = "Fish Recruitment Response to Temperature",
       colour = "Fish Group")

wrap_plots(p_biomass, p_ssb, p_recruit, ncol = 1, guides = "collect")
```

### Community Composition Across the Temperature Gradient

```{r fig.height=10}
# Extract biomass for all groups at steady state (average last 10 years)
all_species <- results_sst[[1]]$param$Groups$Species
ngrps <- length(all_species)

biomass_sst_summary <- data.frame(
  sst = rep(sst_levels, each = ngrps),
  species = rep(all_species, length(sst_levels)),
  biomass = NA,
  type = rep(results_sst[[1]]$param$Groups$Type, length(sst_levels))
)

for (i in seq_along(sst_levels)) {
  mdl <- results_sst[[i]]
  # Average biomass over last 10 years
  avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 10)
  total_biomass <- rowSums(avg_biomass)
  
  for (g in 1:ngrps) {
    row_idx <- (i - 1) * ngrps + g
    biomass_sst_summary$biomass[row_idx] <- total_biomass[g]
  }
}

# Calculate proportions within each sst level
biomass_sst_summary <- biomass_sst_summary %>%
  dplyr::group_by(sst) %>%
  dplyr::mutate(proportion = biomass / sum(biomass, na.rm = TRUE)) %>%
  dplyr::ungroup()

# Set species factor levels for plotting
biomass_sst_summary$species <- factor(biomass_sst_summary$species, levels = all_species)

# Get colours from the model
plot_colours <- results_sst[[1]]$param$Groups$PlotColour
names(plot_colours) <- all_species

# Create stacked proportion plot
p_prop <- ggplot(biomass_sst_summary, aes(x = factor(sst), y = proportion, fill = species)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = plot_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (°C)",
       y = "Proportion of Total Biomass",
       title = "Community Composition Across Temperature Gradient",
       fill = "Species") +
  theme(legend.position = "right")

# Also show absolute biomass
p_abs <- ggplot(biomass_sst_summary, aes(x = sst, y = biomass, colour = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_manual(values = plot_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (°C)",
       y = "Biomass",
       title = "Absolute Biomass Across Temperature Gradient",
       colour = "Species")

wrap_plots(p_prop, p_abs, ncol = 1, guides = "collect")
```

## Summary

This vignette demonstrated the explicit energy budget and fish reproduction system in ZooMSS:

1. **Energy Budget Components**: The model explicitly tracks non-nutritive fraction (defecation + water/structural losses), metabolism, growth, and reproduction fractions. Energy budget parameters (`f_M`, `K_growth`, `R_frac`) are group-specific, ranging from K_growth = 0.25 (Euphausiids) to 0.50 (Jellyfish).

2. **Prey Quality Effects**: The non-nutritive fraction varies continuously with prey Carbon content from `def_high = 0.30` (high-Carbon prey like ciliates/flagellates at 0.15 g C/g WW) to `def_low = 0.95` (low-Carbon prey like jellyfish at 0.005 g C/g WW). This ~12× dynamic range in growth efficiency across prey types captures the fundamental difference between carbon-dense and gelatinous prey.

3. **Food Web Structure**: Emergent PPMR distributions and trophic levels provide diagnostic checks on the model's food web. Trophic levels range from ~2 for herbivores/omnivores to >3 for top predators.

4. **Fish Reproduction**: Fish groups allocate a group-specific fraction of assimilated energy (R_frac) to reproduction, scaled by a maturity ogive. Immature fish redirect R_frac to somatic growth.

5. **Recruitment**: New fish enter the population through explicit recruitment based on Spawning Stock Biomass (SSB).

6. **Environmental Sensitivity**: Reproduction metrics and community composition respond to:
   - Productivity (chlorophyll concentration)
   - Temperature (through metabolic scaling)
   - Seasonal cycles in environmental conditions

### Key Differences: Zooplankton vs Fish

| Aspect | Zooplankton | Fish |
|--------|-------------|------|
| K_growth | 0.25–0.50 (group-specific) | 0.30 (all fish groups) |
| R_frac | 0.00 (not used) | 0.20 (= 1 - f_M - K_growth) |
| repro_on | 0 (disabled) | 1 (enabled) |
| Recruitment | Boundary condition | SSB-driven |

The explicit energy budget provides a mechanistic foundation for understanding how environmental change affects fish population dynamics through reproduction.

## Session Info

```{r}
sessionInfo()
```
