---
title: "Dual-Pathway Energy Budget and Fish Reproduction in ZooMSS"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Dual-Pathway Energy Budget and Fish Reproduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r message=FALSE, warning=FALSE}
devtools::load_all()
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(kableExtra)
library(ggrepel)
```

```{r cache-setup}
# ============================================================================
# MODEL OUTPUT CACHING
# ============================================================================
# Long model runs are cached to avoid re-running on every knit.
# Pre-compute all models by running: source("vignettes/run_gradient_models.R")
# Delete the cache directory to force re-runs.

cache_dir <- here::here("vignettes", "cache")
if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE)

#' Run or load a cached ZooMSS model
#'
#' @param cache_name Character name for the cache file (without extension)
#' @param input_params Input parameters for the model
#' @param Groups Functional groups
#' @param isave Save frequency
#' @param cache_dir Path to cache directory (explicit for parallel safety)
#' @param force_rerun Logical, if TRUE, re-run even if cache exists
run_or_load <- function(cache_name, input_params, Groups, isave = 2,
                        cache_dir, force_rerun = FALSE) {
  cache_file <- file.path(cache_dir, paste0(cache_name, ".rds"))
  if (file.exists(cache_file) && !force_rerun) {
    message("Loading cached model: ", cache_name)
    return(readRDS(cache_file))
  }
  message("Running model: ", cache_name, " (this may take a while...)")
  mdl <- zoomss_model(
    input_params = input_params,
    Groups = Groups,
    isave = isave
  )
  tmp_file <- tempfile(tmpdir = cache_dir, fileext = ".rds")
  saveRDS(mdl, tmp_file)
  file.rename(tmp_file, cache_file)
  message("Cached model saved: ", cache_file)
  return(mdl)
}
```

## Introduction

This vignette demonstrates the dual-pathway growth model and fish reproduction system in ZooMSS. The model uses two distinct growth pathways depending on organism type:

- **Zooplankton** use the original gross growth efficiency (GGE) approach, where growth efficiency is a prey property: $E_{ij} = \text{GrossGEscale}_j \times C_j$
- **Fish** use an explicit DBPM-style (Dynamic Benthic-Pelagic Model) energy budget that tracks defecation, assimilation, metabolism, growth, and reproduction

This dual-pathway design preserves the well-validated zooplankton community composition from the original ZooMSS while enabling mechanistic fish population dynamics through explicit recruitment.

## The Dual-Pathway Growth Model

### Zooplankton: GGE Pathway

Zooplankton growth uses the original ZooMSS formulation where growth efficiency is determined entirely by prey properties. For predator $i$ consuming prey group $j$:

$$
E_{ij} = \text{GrossGEscale}_j \times C_j
$$

where $C_j$ is the prey Carbon content (g C / g wet weight) and $\text{GrossGEscale}_j$ is a scaling coefficient (typically 2.5). This approach treats growth efficiency as a property of the prey, reflecting the nutritional quality of different food sources.

For phytoplankton consumption, the efficiency is:
$$
E_{i,\text{phyto}} = \text{GrossGEscale}_i \times C_{\text{phyto}}
$$

where $C_{\text{phyto}} = 0.1$ is the Carbon content of phytoplankton.

### Fish: Explicit Energy Budget

Fish groups use a two-stage energy partitioning based on DBPM theory:

**Stage 1: Ingestion Partitioning**
$$
\text{Ingestion} = \text{Defecation} + \text{Assimilation}
$$

Where defecation varies continuously based on prey Carbon content:
$$
D_{prey} = D_{high} + (D_{low} - D_{high}) \times \left(1 - \frac{C_{prey}}{C_{max}}\right)
$$

This interpolates between low defecation for high-quality (high Carbon) prey and high defecation for low-quality (low Carbon) prey (Blanchard et al. 2009, 2012).

**Stage 2: Assimilation Partitioning**
$$
\text{Assimilated} = \text{Metabolism} + \text{Growth} + \text{Reproduction}
$$

Where: $f_M + K_{\text{growth}} + R_{\text{frac}} = 1$

For immature fish (below maturation size $W_{\text{mat}}$), $R_{\text{frac}}$ energy is redirected to somatic growth. For mature fish, $R_{\text{frac}}$ energy drives reproductive investment and recruitment.

### Why Two Pathways?

The zooplankton GGE approach has been extensively validated against observed community composition across environmental gradients (Heneghan et al. 2016, 2020). Implementing the explicit energy budget for zooplankton altered community composition in ways that required additional calibration without clear benefit. The explicit energy budget is needed for fish because it enables mechanistic reproduction — energy partitioning into growth vs. reproduction is essential for modelling fish population dynamics under changing environmental and fishing pressure.

Note the dual-pathway distinction: zooplankton growth is driven entirely by `GrossGEscale × Carbon`. For fish, the allocation depends on maturity: **immature fish** redirect $R_{\text{frac}}$ to somatic growth, while **mature fish** split production between growth ($K_{\text{growth}}$) and reproduction ($R_{\text{frac}}$).

## Model Setup

Let's load the default groups and examine the parameters for both pathways:

```{r}
Groups <- getGroups()

# Display zooplankton GGE parameters
cat("=== Zooplankton: GGE Pathway ===\n")
zoo_params <- Groups[Groups$Type == "Zooplankton",
                     c("Species", "GrossGEscale", "Carbon")]
print(zoo_params)

cat("\n=== Fish: Explicit Energy Budget ===\n")
fish_params <- Groups[Groups$Type == "Fish",
                      c("Species", "def_high", "def_low",
                        "f_M", "K_growth", "repro_on", "Carbon")]
print(fish_params)

# Calculate R_frac for fish
cat("\nFish R_frac = 1 - f_M - K_growth =",
    1 - fish_params$f_M[1] - fish_params$K_growth[1], "\n")
```

```markdown
### Functional Group Parameters

The following parameters are constant across all groups: search coefficient (640),
search exponent (0.8), GrossGEscale (2.5), defecation bounds (D_high = 0.30,
D_low = 0.95), and maturity ogive slope (10). The table below shows the parameters
that vary by group.
```

```{r}
# Key parameters that vary across groups
key_params <- Groups[, c("Species", "Type", "FeedType", "W0", "Wmax", "Wmat",
                          "PPMRscale", "PPMR", "FeedWidth", "Carbon",
                          "f_M", "K_growth", "repro_eff", "repro_on",
                          "ZSpre", "ZSexp")]

knitr::kable(key_params,
  col.names = c("Species", "Type", "Feed Type",
                "W0", "Wmax", "Wmat",
                "PPMR scale", "PPMR", "Feed Width", "Carbon",
                "f_M", "K_growth", "repro_eff", "repro_on",
                "ZSpre", "ZSexp"),
  digits = 3,
  caption = "Key parameters for ZooMSS functional groups. Size parameters (W0, Wmax, Wmat) are log10 grams wet weight. Carbon is g C / g WW. Energy budget parameters (f_M, K_growth, repro_eff) are only used for fish groups in the dual-pathway model; zooplankton growth is governed by GrossGEscale × Carbon_prey."
) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            font_size = 11, full_width = FALSE)
```

Notice that:

- **Zooplankton** use `GrossGEscale × Carbon` for growth efficiency — the efficiency depends on what prey is consumed
- **Fish** have `K_growth = 0.30`, `f_M = 0.50`, and `repro_on = 1` — reproduction is enabled with `R_frac = 0.20`
- Both pathways use the same predation kernels, mortality, and temperature scaling — only the growth efficiency calculation differs

## Part 1: Constant Environmental Conditions

First, let's run the model with constant temperature and chlorophyll (1 mg m⁻³) to establish baseline dynamics. We run for 300 years and average over the final 100 years for diagnostics.

```{r}
# Create constant environmental conditions - 300 years
env_constant <- createInputParams(
  time = seq(0, 300, by = 0.1),
  sst = 15,
  chl = 1.0
)

# Plot the environmental forcing
plotEnvironment(env_constant)
```

```{r}
# Run or load from cache
mdl_constant <- run_or_load("mdl_constant_300yr", env_constant, Groups, isave = 2, cache_dir = cache_dir)
```


#### Defecation by Prey Quality (Fish Pathway)

This plot shows how the defecation fraction experienced by fish varies depending on prey Carbon content:

```{r fig.height=5}
plotDefecation(mdl_constant)

geom_text_repel(aes(label = Species), size = 3, max.overlaps = 20)
```

When a fish predator consumes high-Carbon prey (e.g. ciliates at 0.15 g C/g WW), a larger proportion of the ingested material is nutritionally useful, resulting in low defecation (~30%). Conversely, when consuming low-Carbon prey such as gelatinous organisms (e.g. jellyfish at 0.005 g C/g WW), the high water content means most of the ingested material passes through as waste, resulting in high defecation (~95%). Note: zooplankton growth does not use this defecation pathway — their growth efficiency is determined by `GrossGEscale × Carbon_prey`.

#### Prey-Specific Growth Efficiency Matrix

This heatmap shows the net growth efficiency for each predator-prey combination:

```{r fig.height=6}
# ==========================================================================
# Growth efficiency matrix — dual pathway
# Zooplankton predators: GrossGEscale[prey] * Carbon[prey] (prey property, same for all zoo preds)
# Fish predators: (1 - D_prey) * K_growth (predator-specific)
# ==========================================================================

species <- mdl_constant$param$Groups$Species
carbon <- mdl_constant$param$Groups$Carbon
group_type <- mdl_constant$param$Groups$Type
ngrps <- mdl_constant$param$ngrps
GrossGEscale <- mdl_constant$param$Groups$GrossGEscale

# Build growth efficiency matrix
growth_eff_matrix <- matrix(NA, nrow = ngrps, ncol = ngrps)
rownames(growth_eff_matrix) <- species
colnames(growth_eff_matrix) <- species

for (pred in 1:ngrps) {
  for (prey in 1:ngrps) {
    if (group_type[pred] == "Zooplankton") {
      # GGE pathway: efficiency depends on prey only
      growth_eff_matrix[pred, prey] <- GrossGEscale[prey] * carbon[prey]
    } else {
      # Fish energy budget: (1 - def) * K_growth
      Carbon_max <- max(carbon)
      def_high <- mdl_constant$param$Groups$def_high[pred]
      def_low <- mdl_constant$param$Groups$def_low[pred]
      K_growth_val <- mdl_constant$param$Groups$K_growth[pred]
      carbon_frac <- carbon[prey] / Carbon_max
      def_eff <- def_high + (def_low - def_high) * (1 - carbon_frac)
      growth_eff_matrix[pred, prey] <- (1 - def_eff) * K_growth_val
    }
  }
}

# Order prey by Carbon content (low to high)
prey_order <- species[order(carbon)]

# Order predators by type then Carbon
pred_order <- species[order(group_type, carbon)]

# Convert to long format
assim_long <- expand.grid(predator = species, prey = species)
assim_long$efficiency <- as.vector(growth_eff_matrix)
assim_long$predator <- factor(assim_long$predator, levels = pred_order)
assim_long$prey <- factor(assim_long$prey, levels = prey_order)

ggplot(assim_long, aes(x = prey, y = predator, fill = efficiency)) +
  geom_tile() +
  geom_text(aes(label = round(efficiency, 3)), size = 2.5, colour = "white") +
  scale_fill_viridis_c(option = "plasma", limits = c(0, max(growth_eff_matrix, na.rm = TRUE) * 1.1)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Prey Group (ordered by Carbon content)",
       y = "Predator Group",
       fill = "Growth\nEfficiency",
       title = "Prey-Specific Growth Efficiency (Dual Pathway)",
       subtitle = "Zoo predators: GGE×Carbon | Fish predators: (1-D)×K_growth")
```

This matrix displays the growth efficiency for each predator-prey combination. Note the clear distinction between the two pathways: all zooplankton predator rows share identical values (since GGE efficiency is a prey property), while fish predator rows show different values reflecting the energy budget calculation with predator-specific $K_{\text{growth}}$. Prey are ordered by Carbon content along the x-axis (low Carbon like Larvaceans and Salps on the left, high Carbon like Ciliates on the right).

### Biomass Dynamics

Let's examine the biomass dynamics over time:

```{r fig.height=9}
# Plot biomass time series for all groups
p1 <- plotTimeSeries(mdl_constant, by = "biomass", transform = "log10") +
  ggtitle("All Groups - Biomass Time Series")

# Focus on fish groups only
p2 <- plotTimeSeries(mdl_constant, by = "biomass",
                     species = c("Fish_Small", "Fish_Med", "Fish_Large"),
                     transform = "log10") +
  ggtitle("Fish Groups - Biomass Time Series")

# Stacked biomass showing proportions
p3 <- plotTimeSeries(mdl_constant, by = "biomass", type = "fill") +
  ggtitle("Relative Biomass Proportions")

wrap_plots(p1, p2, p3, ncol = 1)
```

### Reproduction Metrics

Now let's examine the reproduction metrics for fish:

```{r fig.height=9}
# Spawning Stock Biomass (SSB)
p_ssb <- plotReproduction(mdl_constant, by = "SSB") +
  ggtitle("Spawning Stock Biomass (SSB)")

# Recruitment
p_recruit <- plotReproduction(mdl_constant, by = "recruitment") +
  ggtitle("Recruitment")

# Total reproductive output
p_repro <- plotReproduction(mdl_constant, by = "repro_output") +
  ggtitle("Total Reproductive Output")

wrap_plots(p_ssb, p_recruit, p_repro, ncol = 1)
```

### Stock-Recruitment Relationships

The relationship between Spawning Stock Biomass (SSB) and recruitment is fundamental to understanding fish population dynamics. Let's examine the emergent stock-recruitment relationships from the model:

```{r fig.height=8}
# Extract SSB and recruitment data for stock-recruitment plots
fish_grps <- mdl_constant$param$fish_grps
fish_names <- mdl_constant$param$Groups$Species[fish_grps]
fish_colours <- mdl_constant$param$Groups$PlotColour[fish_grps]
names(fish_colours) <- fish_names

# Use final 100 years of data for steady-state S-R relationships
n_save <- length(mdl_constant$time)
dt_save <- mdl_constant$param$isave * mdl_constant$param$dt
n_100yr <- min(n_save, round(100 / dt_save))
start_idx <- max(1, n_save - n_100yr + 1)

sr_data <- data.frame()
for (f in seq_along(fish_names)) {
  df <- data.frame(
    SSB = mdl_constant$SSB[start_idx:n_save, f],
    Recruitment = mdl_constant$recruitment[start_idx:n_save, f],
    Fish_Group = fish_names[f],
    Time = mdl_constant$time[start_idx:n_save]
  )
  sr_data <- rbind(sr_data, df)
}

sr_data$Fish_Group <- factor(sr_data$Fish_Group, levels = fish_names)

# Stock-recruitment scatter plots
p_sr <- ggplot(sr_data, aes(x = SSB, y = Recruitment, colour = Fish_Group)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_path(alpha = 0.15, linewidth = 0.3) +
  scale_colour_manual(values = fish_colours) +
  facet_wrap(~Fish_Group, scales = "free") +
  theme_bw() +
  labs(x = "Spawning Stock Biomass (SSB)",
       y = "Recruitment (individuals)",
       title = "Stock-Recruitment Relationships",
       subtitle = "Final 100 years — paths trace temporal trajectory through S-R space")

# SSB vs Recruitment with log axes
p_sr_log <- ggplot(sr_data %>% filter(SSB > 0 & Recruitment > 0),
                   aes(x = SSB, y = Recruitment, colour = Fish_Group)) +
  geom_point(alpha = 0.3, size = 1) +
  scale_colour_manual(values = fish_colours) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~Fish_Group, scales = "free") +
  theme_bw() +
  labs(x = "SSB (log scale)",
       y = "Recruitment (log scale)",
       title = "Stock-Recruitment (log-log scale)")

wrap_plots(p_sr, p_sr_log, ncol = 1)
```

The stock-recruitment relationships show the emergent dynamics between spawning biomass and recruitment. In the constant environment, the oscillating steady-state creates characteristic trajectories in S-R space. The relationship is driven by the explicit fish energy budget where mature fish invest $R_{\text{frac}}$ of their assimilated energy into reproduction, which is then converted to recruits via the reproductive efficiency parameter.

### Growth Rate Comparison

Compare growth rates between fish and zooplankton:

```{r fig.height=6}
# Growth rates for zooplankton (GGE pathway)
p_zoo <- plotGrowthComparison(mdl_constant, groups = "zooplankton", n_years = 100) +
  ggtitle("Zooplankton Growth Rates (GGE Pathway)")

# Growth rates for fish (energy budget pathway, with reproduction)
p_fish <- plotGrowthComparison(mdl_constant, groups = "fish", n_years = 100) +
  ggtitle("Fish Growth Rates (Energy Budget Pathway)")

wrap_plots(p_zoo, p_fish, ncol = 1, guides = "collect")
```

### Reproductive Investment by Size

Examine how reproductive investment varies with body size:

```{r}
plotReproSizeSpectra(mdl_constant, n_years = 100) +
  ggtitle("Fish Reproductive Rate by Body Size")
```

The reproductive rate increases with body size within each fish group, reflecting:
1. The maturity ogive (larger individuals are more likely to be mature)
2. The body-size scaling of growth rates

### Size Spectra

```{r fig.height=6}
# Abundance size spectrum
p_abund <- plotSizeSpectra(mdl_constant, by = "abundance", n_years = 100) +
  ggtitle("Abundance Size Spectrum")

# Biomass size spectrum
p_biom <- plotSizeSpectra(mdl_constant, by = "biomass", n_years = 100) +
  ggtitle("Biomass Size Spectrum")

wrap_plots(p_abund, p_biom, ncol = 1, guides = "collect")
```

## Part 2: Seasonal Environmental Forcing

Now let's compare with seasonal environmental conditions to see how reproduction responds to environmental variability. We use the same base chlorophyll (1 mg m⁻³) but with seasonal variation, run for 300 years.

```{r}
# Create seasonal environmental conditions
env_seasonal <- createEnviroData(
  n_years = 300,
  dt = 0.1,
  base_sst = 15,
  base_chl = 1.0,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5
)

env_seasonal_params <- createInputParams(
  time = env_seasonal$time,
  sst = env_seasonal$sst,
  chl = env_seasonal$chl
)

```

```{r}
# Run or load from cache
mdl_seasonal <- run_or_load("mdl_seasonal_300yr", env_seasonal_params, Groups, isave = 2, cache_dir = cache_dir)
```

### Comparing Constant vs Seasonal: Biomass

```{r fig.height=9}
# Constant environment biomass - focus on final 100 years
p_const <- plotTimeSeries(mdl_constant, by = "biomass",
                          species = c("Fish_Small", "Fish_Med", "Fish_Large"),
                          transform = "log10") +
  ggtitle("Constant Environment - Fish Biomass") +
  xlim(200, 300)

# Seasonal environment biomass
p_seas <- plotTimeSeries(mdl_seasonal, by = "biomass",
                         species = c("Fish_Small", "Fish_Med", "Fish_Large"),
                         transform = "log10") +
  ggtitle("Seasonal Environment - Fish Biomass") +
  xlim(200, 300)

# Show the seasonal forcing for context
p_env <- plotEnvironment(env_seasonal_params) & xlim(200, 300)

wrap_plots(p_env, p_const, p_seas, ncol = 1, heights = c(1, 1, 1))
```

### Comparing Constant vs Seasonal: Reproduction

```{r fig.height=9}
# SSB comparison
# Get shared y-axis ranges
ssb_range <- range(c(mdl_constant$SSB[start_idx:n_save, ],
                     mdl_seasonal$SSB[start_idx:n_save, ]), na.rm = TRUE)
rec_range <- range(c(mdl_constant$recruitment[start_idx:n_save, ],
                     mdl_seasonal$recruitment[start_idx:n_save, ]), na.rm = TRUE)

# SSB comparison
p_ssb_const <- plotReproduction(mdl_constant, by = "SSB") +
  ggtitle("Constant - SSB") + xlim(200, 300) + ylim(ssb_range)

p_ssb_seas <- plotReproduction(mdl_seasonal, by = "SSB") +
  ggtitle("Seasonal - SSB") + xlim(200, 300) + ylim(ssb_range)

# Recruitment comparison
p_rec_const <- plotReproduction(mdl_constant, by = "recruitment") +
  ggtitle("Constant - Recruitment") + xlim(200, 300) + ylim(rec_range)

p_rec_seas <- plotReproduction(mdl_seasonal, by = "recruitment") +
  ggtitle("Seasonal - Recruitment") + xlim(200, 300) + ylim(rec_range)

wrap_plots(p_ssb_const, p_ssb_seas, p_rec_const, p_rec_seas,
           ncol = 2, guides = "collect")
```

### Stock-Recruitment: Constant vs Seasonal

```{r fig.height=6, fig.width=8}
# Build S-R data for both scenarios
build_sr_data <- function(mdl, scenario_name, n_years_avg = 100) {
  fish_grps <- mdl$param$fish_grps
  fish_names <- mdl$param$Groups$Species[fish_grps]
  n_save <- length(mdl$time)
  dt_save <- mdl$param$isave * mdl$param$dt
  n_avg <- min(n_save, round(n_years_avg / dt_save))
  start_idx <- max(1, n_save - n_avg + 1)

  sr <- data.frame()
  for (f in seq_along(fish_names)) {
    df <- data.frame(
      SSB = mdl$SSB[start_idx:n_save, f],
      Recruitment = mdl$recruitment[start_idx:n_save, f],
      Fish_Group = fish_names[f],
      Scenario = scenario_name
    )
    sr <- rbind(sr, df)
  }
  sr$Fish_Group <- factor(sr$Fish_Group, levels = fish_names)
  return(sr)
}

sr_const <- build_sr_data(mdl_constant, "Constant")
sr_seas <- build_sr_data(mdl_seasonal, "Seasonal")
sr_combined <- rbind(sr_const, sr_seas)

ggplot(sr_combined %>% filter(SSB > 0 & Recruitment > 0),
       aes(x = SSB, y = Recruitment, colour = Scenario)) +
  geom_point(alpha = 0.3, size = 1) +
  facet_wrap(~Fish_Group, scales = "free") +
  scale_colour_manual(values = c("Constant" = "grey40", "Seasonal" = "coral")) +
  theme_bw() +
  labs(x = "Spawning Stock Biomass (SSB)",
       y = "Recruitment",
       title = "Stock-Recruitment: Constant vs Seasonal Environment",
       subtitle = "Final 100 years of 300-year simulations")
```

## Part 3: Exploring Chlorophyll Gradients

Let's explore how productivity (chlorophyll concentration) affects reproduction metrics and community composition across a wide gradient from oligotrophic to eutrophic conditions. For the chlorophyll *a* gradient, a distribution of log~10~(Chlorophyll *a*) values were computed at 0.1 intervals between -1.7 and 0.5 log~10~(Chlorophyll *a*) (n = 23) and the mean proportion calculated for each of the bins. We run each scenario for 300 years to ensure near-steady-state.

```{r}
# Chlorophyll gradient: log10(chl) from -1.7 to 0.5 at 0.1 intervals (n = 23)
log10_chl_seq <- seq(-1.7, 0.5, by = 0.1)
chl_levels <- 10^log10_chl_seq
```

### Multi-Temperature Chlorophyll Gradient Experiments

We run the chlorophyll gradient at three temperatures (10°C, 15°C, 20°C) to examine how temperature modulates the productivity response.

```{r}
# Load pre-computed models across chlorophyll gradient at multiple temperatures
# (69 models computed by run_gradient_models.R)
sst_variants <- c(10, 15, 20)

results_chl <- list()
for (temp in sst_variants) {
  temp_char <- as.character(temp)
  results_chl[[temp_char]] <- lapply(seq_along(chl_levels), function(i) {
    cache_name <- sprintf("mdl_chl_log10_%.1f_sst%d_300yr", log10(chl_levels[i]), temp)
    cache_file <- file.path(cache_dir, paste0(cache_name, ".rds"))
    if (!file.exists(cache_file)) stop("Missing cache: ", cache_name,
      ". Run source('vignettes/run_gradient_models.R') first.")
    readRDS(cache_file)
  })
}
```

### Fish Biomass, SSB and Recruitment Across the Gradient

```{r fig.height=12}
# Extract fish metrics across chlorophyll gradient for all temperatures
num_fish <- 3
fish_names <- c("Fish_Small", "Fish_Med", "Fish_Large")

fish_summary_all <- data.frame()

for (temp in sst_variants) {
  temp_char <- as.character(temp)
  fish_grps <- results_chl[[temp_char]][[1]]$param$fish_grps

  for (i in seq_along(chl_levels)) {
    mdl <- results_chl[[temp_char]][[i]]
    final_idx <- length(mdl$time)

    # Average biomass over last 100 years
    avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 100)
    fish_biomass <- rowSums(avg_biomass[fish_grps, , drop = FALSE])

    for (f in 1:num_fish) {
      fish_summary_all <- rbind(fish_summary_all, data.frame(
        chl = chl_levels[i],
        sst = temp,
        fish_group = fish_names[f],
        final_biomass = fish_biomass[f],
        final_SSB = mdl$SSB[final_idx, f],
        final_recruitment = mdl$recruitment[final_idx, f]
      ))
    }
  }
}

fish_summary_all$sst_label <- paste0(fish_summary_all$sst, "\u00B0C")

# Get fish colours
fish_colours <- results_chl[["15"]][[1]]$param$Groups$PlotColour[
  results_chl[["15"]][[1]]$param$fish_grps]
names(fish_colours) <- fish_names

# Plot Fish Biomass vs chlorophyll across temperatures
p_biomass <- ggplot(fish_summary_all,
                    aes(x = log10(chl), y = final_biomass, colour = fish_group,
                        linetype = sst_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Biomass",
       title = "Fish Biomass Response to Productivity Gradient",
       colour = "Fish Group", linetype = "Temperature")

# SSB
p_ssb <- ggplot(fish_summary_all,
                aes(x = log10(chl), y = final_SSB, colour = fish_group,
                    linetype = sst_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Spawning Stock Biomass (SSB)",
       title = "Fish SSB Response to Productivity Gradient",
       colour = "Fish Group", linetype = "Temperature")

# Recruitment
p_recruit <- ggplot(fish_summary_all,
                    aes(x = log10(chl), y = final_recruitment, colour = fish_group,
                        linetype = sst_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Recruitment",
       title = "Fish Recruitment Response to Productivity Gradient",
       colour = "Fish Group", linetype = "Temperature")

wrap_plots(p_biomass, p_ssb, p_recruit, ncol = 1, guides = "collect")
```

### Stock-Recruitment Across the Chlorophyll Gradient

```{r fig.height=12, fig.width=8}
# Build S-R data across chlorophyll gradient at 15°C
sr_chl_data <- data.frame()
for (i in seq_along(chl_levels)) {
  mdl <- results_chl[["15"]][[i]]
  sr_temp <- build_sr_data(mdl, paste0("Chl = ", round(chl_levels[i], 3), " mg/m\u00B3"), n_years_avg = 100)
  sr_temp$chl <- chl_levels[i]
  sr_chl_data <- rbind(sr_chl_data, sr_temp)
}

ggplot(sr_chl_data %>% filter(SSB > 0 & Recruitment > 0),
       aes(x = SSB, y = Recruitment, colour = log10(chl))) +
  geom_point(alpha = 0.3, size = 1) +
  facet_wrap(~Fish_Group, scales = "free", ncol = 1) +
  scale_colour_viridis_c(option = "viridis",
                         name = expression(log[10] * "(Chl " * italic(a) * ")")) +
  theme_bw() +
  labs(x = "Spawning Stock Biomass (SSB)",
       y = "Recruitment",
       title = "Stock-Recruitment Across Productivity Gradient (15\u00B0C)",
       subtitle = "Final 100 years of 300-year simulations")
```

### Community Composition Across the Gradient

Let's examine how community composition changes across the productivity gradient. Following Heneghan et al., we split the lower trophic levels into two panels: (1) phytoplankton size classes with flagellates and ciliates, and (2) the remaining zooplankton groups.

```{r fig.height=6, fig.width=8}
# ============================================================================
# HELPER: Extract community composition with phytoplankton size classes
# ============================================================================
extract_community_composition <- function(results_list, chl_levels, Groups,
                                          n_years_avg = 100) {

  all_species <- Groups$Species
  ngrps <- length(all_species)

  # ---- Zooplankton/Fish biomass ----
  biomass_summary <- data.frame()
  for (i in seq_along(chl_levels)) {
    mdl <- results_list[[i]]
    avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = n_years_avg)
    total_biomass <- rowSums(avg_biomass)

    for (g in 1:ngrps) {
      biomass_summary <- rbind(biomass_summary, data.frame(
        chl = chl_levels[i],
        species = all_species[g],
        biomass = total_biomass[g],
        type = Groups$Type[g]
      ))
    }
  }

  # ---- Phytoplankton size class biomass ----
  phyto_summary <- data.frame()
  for (i in seq_along(chl_levels)) {
    mdl <- results_list[[i]]
    # Use the phytoplankton parameters to calculate size class biomass
    env_row <- data.frame(chl = chl_levels[i])
    phyto_params <- calculatePhytoParam(env_row)

    phyto_summary <- rbind(phyto_summary, data.frame(
      chl = chl_levels[i],
      species = c("PicoPhytoplankton", "NanoPhytoplankton", "MicroPhytoplankton"),
      biomass = c(phyto_params$pico_biom, phyto_params$nano_biom, phyto_params$micro_biom),
      type = "Phytoplankton"
    ))
  }

  return(list(zoo_fish = biomass_summary, phyto = phyto_summary))
}

# ============================================================================
# HELPER: Create split-panel community composition plot
# ============================================================================
plot_split_community <- function(results_list, chl_levels, Groups,
                                 temperature_label, n_years_avg = 100) {

  comp <- extract_community_composition(results_list, chl_levels, Groups,
                                        n_years_avg = n_years_avg)

  # Get model colours
  plot_colours <- Groups$PlotColour
  names(plot_colours) <- Groups$Species

  # Add phytoplankton colours
  phyto_colours <- c(
    "PicoPhytoplankton" = "#c7e9c0",
    "NanoPhytoplankton" = "#74c476",
    "MicroPhytoplankton" = "#238b45"
  )

  # ---- PANEL 1: Phyto size classes + Flagellates + Ciliates ----
  upper_groups <- c("PicoPhytoplankton", "NanoPhytoplankton", "MicroPhytoplankton",
                    "Flagellates", "Ciliates")

  # Combine phyto and small zoo data
  flag_cil <- comp$zoo_fish %>%
    filter(species %in% c("Flagellates", "Ciliates"))
  upper_data <- rbind(comp$phyto, flag_cil)
  upper_data$species <- factor(upper_data$species, levels = upper_groups)
  
  # Clamp negative biomass before computing proportions
  upper_data$biomass <- pmax(0, upper_data$biomass)

  # Calculate proportions within upper panel
  upper_data <- upper_data %>%
    group_by(chl) %>%
    mutate(proportion = biomass / sum(biomass, na.rm = TRUE)) %>%
    ungroup()
  
  upper_colours <- c(phyto_colours,
                   "Flagellates" = "#006d2c",
                   "Ciliates"    = "#00441b")
  
  p_upper <- ggplot(upper_data,
                    aes(x = log10(chl), y = proportion, fill = species)) +
    geom_area(position = "fill") +
    # geom_area(position = "stack") +
    scale_fill_manual(values = upper_colours, name = "Species") +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_continuous(expand = c(0, 0)) +
    coord_cartesian(xlim = c(-1.3, 0.5)) +
    theme_bw() +
    theme(axis.title.x = element_blank(),
          legend.position = "right") +
    labs(y = "Proportion")

  # ---- PANEL 2: Remaining zooplankton (no Flagellates, Ciliates, no Fish) ----
  lower_groups <- c("OmniCopepods", "Euphausiids", "Jellyfish", "Chaetognaths",
                  "CarnCopepods", "Salps", "Larvaceans")

  lower_data <- comp$zoo_fish %>%
    filter(species %in% lower_groups)
  lower_data$species <- factor(lower_data$species, levels = lower_groups)

  # Calculate proportions within lower panel
  lower_data <- lower_data %>%
    group_by(chl) %>%
    mutate(proportion = biomass / sum(biomass, na.rm = TRUE)) %>%
    ungroup()

  lower_colours <- plot_colours[lower_groups]

  p_lower <- ggplot(lower_data,
                    aes(x = log10(chl), y = proportion, fill = species)) +
    geom_area(position = "stack") +
    scale_fill_manual(values = lower_colours, name = "Species") +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_continuous(expand = c(0, 0)) +
    coord_cartesian(xlim = c(-1.3, 0.5)) +
    theme_bw() +
    theme(legend.position = "right") +
    labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
         y = "Proportion")

  # Combine with title
  combined <- wrap_plots(p_upper, p_lower, ncol = 1) +
    plot_annotation(title = paste0("Community Composition Across Productivity Gradient (",
                                   temperature_label, ")"))

  return(combined)
}

# ---- Plot community composition at 15°C ----
plot_split_community(results_chl[["15"]], chl_levels, Groups,
                     "15\u00B0C", n_years_avg = 100)
```

```{r fig.height=6, fig.width=8}
# ---- Plot community composition at 10°C ----
plot_split_community(results_chl[["10"]], chl_levels, Groups,
                     "10\u00B0C", n_years_avg = 100)
```

```{r fig.height=6, fig.width=8}
# ---- Plot community composition at 20°C ----
plot_split_community(results_chl[["20"]], chl_levels, Groups,
                     "20\u00B0C", n_years_avg = 100)
```


```{r fig.height=10, fig.width=10}
# ============================================================================
# COEFFICIENT OF VARIATION ANALYSIS
# ============================================================================
# Temporal CV: how variable is biomass through time at each chl level?
#
# Gradient CV: how variable is mean biomass across the chl gradient?

# Use 15°C results
results_15 <- results_chl[["15"]]
fish_grps <- results_15[[1]]$param$fish_grps
zoo_grps <- which(Groups$Type == "Zooplankton")
all_grps <- seq_len(nrow(Groups))

# --------------------------------------------------------------------------
# 1. Temporal CV at each chl level (final 100 years)
# --------------------------------------------------------------------------
temporal_cv <- data.frame()

for (i in seq_along(chl_levels)) {
  mdl <- results_15[[i]]
  n_save <- length(mdl$time)
  dt_save <- mdl$param$isave * mdl$param$dt
  n_100yr <- min(n_save, round(100 / dt_save))
  start_idx <- max(1, n_save - n_100yr + 1)

  # Extract biomass time series (ngrps x ngrid saved at each time step)
  # Total biomass per group at each saved time step
  biomass_ts <- mdl$biomass[start_idx:n_save, , ]  # time x group x size

  # Sum across size bins for each group at each time step
  group_biomass_ts <- apply(biomass_ts, c(1, 2), sum)  # time x group

  # Fish by group
  for (f in seq_along(fish_grps)) {
    fg <- fish_grps[f]
    ts <- group_biomass_ts[, fg]
    temporal_cv <- rbind(temporal_cv, data.frame(
      chl = chl_levels[i],
      category = "Fish (by group)",
      group = Groups$Species[fg],
      cv = sd(ts, na.rm = TRUE) / mean(ts, na.rm = TRUE)
    ))
  }

  # Fish total
  fish_total_ts <- rowSums(group_biomass_ts[, fish_grps, drop = FALSE])
  temporal_cv <- rbind(temporal_cv, data.frame(
    chl = chl_levels[i],
    category = "Fish (total)",
    group = "All Fish",
    cv = sd(fish_total_ts, na.rm = TRUE) / mean(fish_total_ts, na.rm = TRUE)
  ))

  # Zoo by group
  for (z in zoo_grps) {
    ts <- group_biomass_ts[, z]
    temporal_cv <- rbind(temporal_cv, data.frame(
      chl = chl_levels[i],
      category = "Zooplankton (by group)",
      group = Groups$Species[z],
      cv = sd(ts, na.rm = TRUE) / mean(ts, na.rm = TRUE)
    ))
  }

  # Zoo total
  zoo_total_ts <- rowSums(group_biomass_ts[, zoo_grps, drop = FALSE])
  temporal_cv <- rbind(temporal_cv, data.frame(
    chl = chl_levels[i],
    category = "Zooplankton (total)",
    group = "All Zooplankton",
    cv = sd(zoo_total_ts, na.rm = TRUE) / mean(zoo_total_ts, na.rm = TRUE)
  ))

  # Total community
  community_ts <- rowSums(group_biomass_ts)
  temporal_cv <- rbind(temporal_cv, data.frame(
    chl = chl_levels[i],
    category = "Total community",
    group = "All Groups",
    cv = sd(community_ts, na.rm = TRUE) / mean(community_ts, na.rm = TRUE)
  ))
}

# Handle NaN/Inf from zero-biomass groups
temporal_cv$cv[!is.finite(temporal_cv$cv)] <- NA

# --------------------------------------------------------------------------
# 2. Gradient CV (variability of mean biomass across chl levels)
# --------------------------------------------------------------------------
gradient_means <- data.frame()

for (i in seq_along(chl_levels)) {
  mdl <- results_15[[i]]
  avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 100)
  total_biomass <- rowSums(avg_biomass)

  for (g in all_grps) {
    gradient_means <- rbind(gradient_means, data.frame(
      chl = chl_levels[i],
      group = Groups$Species[g],
      type = Groups$Type[g],
      biomass = total_biomass[g]
    ))
  }
}

gradient_cv <- gradient_means %>%
  group_by(group, type) %>%
  summarise(
    mean_biomass = mean(biomass, na.rm = TRUE),
    sd_biomass   = sd(biomass, na.rm = TRUE),
    cv           = sd_biomass / mean_biomass,
    .groups = "drop"
  )

# Add aggregate rows
fish_agg <- gradient_means %>%
  filter(type == "Fish") %>%
  group_by(chl) %>%
  summarise(biomass = sum(biomass), .groups = "drop") %>%
  summarise(cv = sd(biomass) / mean(biomass)) %>%
  mutate(group = "All Fish", type = "Fish (total)")

zoo_agg <- gradient_means %>%
  filter(type == "Zooplankton") %>%
  group_by(chl) %>%
  summarise(biomass = sum(biomass), .groups = "drop") %>%
  summarise(cv = sd(biomass) / mean(biomass)) %>%
  mutate(group = "All Zooplankton", type = "Zooplankton (total)")

community_agg <- gradient_means %>%
  group_by(chl) %>%
  summarise(biomass = sum(biomass), .groups = "drop") %>%
  summarise(cv = sd(biomass) / mean(biomass)) %>%
  mutate(group = "All Groups", type = "Total community")

gradient_cv_summary <- bind_rows(gradient_cv, fish_agg, zoo_agg, community_agg)

# --------------------------------------------------------------------------
# PLOTS
# --------------------------------------------------------------------------

# --- Temporal CV across chl gradient: aggregates ---
p_temporal_agg <- temporal_cv %>%
  filter(category %in% c("Fish (total)", "Zooplankton (total)", "Total community")) %>%
  ggplot(aes(x = log10(chl), y = cv, colour = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Temporal CV",
       title = "Temporal Coefficient of Variation Across Productivity Gradient",
       subtitle = "CV of biomass time series (final 100 years) at 15°C",
       colour = "Group")

# --- Temporal CV: fish by group ---
p_temporal_fish <- temporal_cv %>%
  filter(category == "Fish (by group)") %>%
  ggplot(aes(x = log10(chl), y = cv, colour = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Temporal CV",
       title = "Fish Temporal CV by Group",
       colour = "Fish Group")

# --- Temporal CV: zoo by group ---
zoo_colours <- Groups$PlotColour[zoo_grps]
names(zoo_colours) <- Groups$Species[zoo_grps]

# Map FeedType to each zoo group
zoo_feedtype <- setNames(Groups$FeedType[zoo_grps], Groups$Species[zoo_grps])

feedtype_colours <- c(
  "Heterotroph"  = "#1b9e77",
  "Omnivore"     = "#7570b3",
  "Carnivore"    = "#d95f02",
  "FilterFeeder" = "#e7298a"
)

p_temporal_zoo <- temporal_cv %>%
  filter(category == "Zooplankton (by group)") %>%
  mutate(feed_type = zoo_feedtype[group]) %>%
  ggplot(aes(x = log10(chl), y = cv, colour = feed_type, group = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_colour_manual(values = feedtype_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Temporal CV",
       title = "Zooplankton Temporal CV by Group",
       subtitle = "Coloured by feeding type",
       colour = "Feed Type")

# p_temporal_zoo <- temporal_cv %>%
#   filter(category == "Zooplankton (by group)") %>%
#   ggplot(aes(x = log10(chl), y = cv, colour = group)) +
#   geom_line(linewidth = 1) +
#   geom_point(size = 1.5) +
#   scale_colour_manual(values = zoo_colours) +
#   theme_bw() +
#   labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
#        y = "Temporal CV",
#        title = "Zooplankton Temporal CV by Group",
#        colour = "Zoo Group")

# --- Gradient CV: bar chart ---
gradient_cv_summary$group <- factor(gradient_cv_summary$group,
  levels = c(Groups$Species, "All Fish", "All Zooplankton", "All Groups"))


# Build colour map: group-specific for individuals, custom for aggregates
bar_colours <- c(
  setNames(Groups$PlotColour, Groups$Species),
  "All Fish"        = "grey40",
  "All Zooplankton" = "grey60",
  "All Groups"      = "grey80"
)

p_gradient <- ggplot(gradient_cv_summary,
                     aes(x = group, y = cv, fill = group)) +
  geom_col() +
  scale_fill_manual(values = bar_colours) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(x = "Group", y = "Gradient CV",
       title = "Gradient CV: Variability of Mean Biomass Across Chlorophyll Levels",
       subtitle = "How much does each group's mean biomass change across productivity (15°C)")

wrap_plots(p_temporal_agg, p_temporal_fish, p_temporal_zoo, p_gradient,
           ncol = 2, guides = "collect")
```

The temporal CV tells you which groups oscillate most at a given productivity, while the gradient CV tells you which groups are most sensitive to productivity itself. Fish with weak recruitment will typically show high temporal CV (boom-bust cycles) and the gradient CV will highlight whether zooplankton or fish composition is more responsive to chlorophyll.

### Full Community Composition (Including Fish)

For completeness, here is the full community composition including fish groups across the gradient at 15°C:

```{r fig.height=8, fig.width=10}
# Standard full community composition plot (all groups)
all_species <- Groups$Species
ngrps <- length(all_species)
plot_colours <- Groups$PlotColour
names(plot_colours) <- all_species

biomass_full <- data.frame()
for (i in seq_along(chl_levels)) {
  mdl <- results_chl[["15"]][[i]]
  avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 100)
  total_biomass <- rowSums(avg_biomass)

  for (g in 1:ngrps) {
    biomass_full <- rbind(biomass_full, data.frame(
      chl = chl_levels[i],
      species = all_species[g],
      biomass = total_biomass[g],
      type = Groups$Type[g]
    ))
  }
}

biomass_full <- biomass_full %>%
  group_by(chl) %>%
  mutate(proportion = biomass / sum(biomass, na.rm = TRUE)) %>%
  ungroup()

biomass_full$species <- factor(biomass_full$species, levels = all_species)

# Proportion plot
p_prop <- ggplot(biomass_full,
                 aes(x = log10(chl), y = proportion, fill = species)) +
  geom_area(position = "stack") +
  scale_fill_manual(values = plot_colours) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(xlim = c(-1.3, 0.5)) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Proportion of Total Biomass",
       title = "Community Composition Across Productivity Gradient (15\u00B0C)",
       fill = "Species")

# Absolute biomass
p_abs <- ggplot(biomass_full, aes(x = log10(chl), y = biomass, colour = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_x_continuous(expand = c(0, 0)) +
  scale_colour_manual(values = plot_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Biomass",
       title = "Absolute Biomass Across Productivity Gradient (15\u00B0C)",
       colour = "Species")

wrap_plots(p_prop, p_abs, ncol = 1, guides = "collect")
```

## Part 4: Temperature Effects

Explore how temperature affects fish reproduction through metabolic scaling. We run 300-year simulations at 1 mg m⁻³ chlorophyll across a temperature gradient.

```{r fig.height=6, fig.width=10}
# Load pre-computed models across temperature gradient
# (5 models computed by run_gradient_models.R)
sst_levels <- c(5, 10, 15, 20)

results_sst <- lapply(seq_along(sst_levels), function(i) {
  cache_name <- sprintf("mdl_sst%d_chl1_300yr", sst_levels[i])
  cache_file <- file.path(cache_dir, paste0(cache_name, ".rds"))
  if (!file.exists(cache_file)) stop("Missing cache: ", cache_name,
    ". Run source('vignettes/run_gradient_models.R') first.")
  readRDS(cache_file)
})
```

### Fish Biomass, SSB and Recruitment Across the Temperature Gradient

```{r fig.height=10}
# Extract fish metrics across temperature gradient
fish_sst_summary <- data.frame()

for (i in seq_along(sst_levels)) {
  mdl <- results_sst[[i]]
  final_idx <- length(mdl$time)
  fish_grps <- mdl$param$fish_grps

  avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 100)
  fish_biomass <- rowSums(avg_biomass[fish_grps, , drop = FALSE])

  for (f in 1:num_fish) {
    fish_sst_summary <- rbind(fish_sst_summary, data.frame(
      sst = sst_levels[i],
      fish_group = fish_names[f],
      final_biomass = fish_biomass[f],
      final_SSB = mdl$SSB[final_idx, f],
      final_recruitment = mdl$recruitment[final_idx, f]
    ))
  }
}

# Biomass vs temperature
p_biomass <- ggplot(fish_sst_summary,
                    aes(x = sst, y = final_biomass, colour = fish_group)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (\u00B0C)", y = "Biomass",
       title = "Fish Biomass Response to Temperature", colour = "Fish Group")

# SSB vs temperature
p_ssb <- ggplot(fish_sst_summary,
                aes(x = sst, y = final_SSB, colour = fish_group)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (\u00B0C)", y = "SSB",
       title = "Fish SSB Response to Temperature", colour = "Fish Group")

# Recruitment vs temperature
p_recruit <- ggplot(fish_sst_summary,
                    aes(x = sst, y = final_recruitment, colour = fish_group)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (\u00B0C)", y = "Recruitment",
       title = "Fish Recruitment Response to Temperature", colour = "Fish Group")

wrap_plots(p_biomass, p_ssb, p_recruit, ncol = 1, guides = "collect")
```

### Stock-Recruitment Across the Temperature Gradient

```{r fig.height=12, fig.width=6}
# S-R relationships across temperature gradient
sr_sst_data <- data.frame()
for (i in seq_along(sst_levels)) {
  sr_temp <- build_sr_data(results_sst[[i]],
                           paste0(sst_levels[i], "\u00B0C"),
                           n_years_avg = 100)
  sr_temp$sst <- sst_levels[i]
  sr_sst_data <- rbind(sr_sst_data, sr_temp)
}

ggplot(sr_sst_data %>% filter(SSB > 0 & Recruitment > 0),
       aes(x = SSB, y = Recruitment, colour = factor(sst))) +
  geom_point(alpha = 0.3, size = 0.8) +
  facet_wrap(~Fish_Group, scales = "free", ncol = 1) +
  scale_colour_manual(
    values = c("5" = "#2166ac", "10" = "#67a9cf", "15" = "#f7f7f7",
               "20" = "#ef8a62", "25" = "#b2182b"),
    name = "SST (\u00B0C)"
  ) +
  theme_bw() +
  labs(x = "Spawning Stock Biomass (SSB)",
       y = "Recruitment",
       title = "Stock-Recruitment Across Temperature Gradient",
       subtitle = "Final 100 years of 300-year simulations at 1 mg/m\u00B3 chlorophyll")
```

### Community Composition Across the Temperature Gradient

```{r fig.height=10}
# Community composition across temperature gradient
biomass_sst <- data.frame()
for (i in seq_along(sst_levels)) {
  mdl <- results_sst[[i]]
  avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 100)
  total_biomass <- rowSums(avg_biomass)

  for (g in 1:ngrps) {
    biomass_sst <- rbind(biomass_sst, data.frame(
      sst = sst_levels[i],
      species = all_species[g],
      biomass = total_biomass[g],
      type = Groups$Type[g]
    ))
  }
}

biomass_sst <- biomass_sst %>%
  group_by(sst) %>%
  mutate(proportion = biomass / sum(biomass, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(proportion = replace(proportion, is.nan(proportion), 0))

biomass_sst$species <- factor(biomass_sst$species, levels = all_species)

# Proportion plot
p_prop <- ggplot(biomass_sst, aes(x = factor(sst), y = proportion, fill = species)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = plot_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (\u00B0C)",
       y = "Proportion of Total Biomass",
       title = "Community Composition Across Temperature Gradient",
       fill = "Species")

# Absolute biomass
p_abs <- ggplot(biomass_sst, aes(x = sst, y = biomass, colour = species)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_manual(values = plot_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (\u00B0C)",
       y = "Biomass",
       title = "Absolute Biomass Across Temperature Gradient",
       colour = "Species")

wrap_plots(p_prop, p_abs, ncol = 1, guides = "collect")
```

## Summary

This vignette demonstrated the dual-pathway growth model and fish reproduction system in ZooMSS:

1. **Dual Growth Pathways**: Zooplankton use the original GGE approach ($E_{ij} = \text{GrossGEscale} \times C_{\text{prey}}$) preserving validated community composition, while fish use an explicit energy budget with prey-specific defecation and partitioning into metabolism, growth, and reproduction

2. **Prey Quality Effects (Fish)**: Fish defecation varies continuously with prey Carbon content — high-Carbon prey (like ciliates) have low defecation (~30%) while low-Carbon prey (like jellyfish) have high defecation (~95%)

3. **Fish Reproduction**: Fish groups allocate $R_{\text{frac}}$ of assimilated energy to reproduction, scaled by a maturity ogive that transitions smoothly around the maturation size

4. **Recruitment**: New fish enter the population through explicit recruitment based on Spawning Stock Biomass (SSB) and reproductive efficiency

5. **Stock-Recruitment Relationships**: Emergent S-R dynamics arise from the energy budget, with characteristic patterns that vary across environmental gradients

6. **Environmental Sensitivity**: Reproduction metrics respond to productivity (chlorophyll), temperature (metabolic scaling), and seasonal cycles

### Key Differences Between Pathways

| Aspect | Zooplankton | Fish |
|--------|-------------|------|
| Growth efficiency | GrossGEscale × Carbon (prey property) | (1 − D) × K_growth (energy budget) |
| Phyto assimilation | GrossGEscale × cc_phyto | (1 − def_phyto) |
| Diffusion scaling | (GGE)² | eff_K² × (1 − def)² |
| Reproduction | None (boundary condition) | R_frac → SSB → recruitment |
| Boundary condition | Proportional lock | Semi-implicit recruitment |

The dual-pathway design provides a mechanistic foundation for understanding how environmental change and fishing pressure affect fish population dynamics through reproduction, while maintaining consistency with the well-validated zooplankton community composition from the original ZooMSS.

## Session Info

```{r}
sessionInfo()
```
