---
title: "Getting started with ZooMSS"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    number_sections: false
vignette: >
  %\VignetteIndexEntry{zoomss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r message=FALSE, warning=FALSE}
library(zoomss)
library(ggplot2)
```


```{r}

plotTimeSeries_exclude <- function(mdl, exclude = NULL, ...) {
  keep <- setdiff(mdl$param$Groups$Species, exclude)
  plotTimeSeries(mdl, species = keep, ...)
}


# usage
p_biomass_vvlow <- plotTimeSeries_exclude(
  mdl_vvlow,
  exclude = "Fish_Med",
  by = "biomass",
  transform = "log10"
) + theme(legend.position = "none")

```

## Input Data

ZooMSS requires two sets of input data:

1. **Groups** - Contains all taxa-specific parameter values for each model group, including size ranges and functional group properties.

2. **Environmental data** - A Time-series dataframe with time series of environmental conditions  with `time`, `sst`, and `chl` columns.

## Running the default Model

Get the default published `Groups` dataframe using:


```{r}
Groups <- getGroups()

# Groups$repro_eff[Groups$Type == "Fish"] <- 0.1

# Groups$PPMR[Groups$Species == "Fish_Large"] <- 500

# Groups$PPMR[Groups$Species == "Fish_Med"] <- 100

# Groups$FeedWidth[Groups$Species == "Fish_Med"] <- 1.4
# Groups$FeedWidth[Groups$Species == "Fish_Large"] <- 1.3
# Groups$FeedWidth[Groups$Species == "Fish_Small"] <- 1.3

# Groups$Wmat[Groups$Species == "Fish_Med"] <- 1
# Groups$Wmat[Groups$Species == "Fish_Large"] <- 3
# Groups$Wmat[Groups$Species == "Fish_Small"] <- -1

# Groups$ZSpre[Groups$Type == "Fish"] <- 0.1
# Groups$ZSexp[Groups$Type == "Fish"] <- 0.3
# Groups$ZSpre[Groups$Species == "Fish_Med"] <- 0.15
# Groups$ZSpre[Groups$Species == "Fish_Small"] <- 0.1
# 
# Groups$ZSpre[Groups$Species == "Fish_Large"] <- 0.2
# Groups$ZSexp[Groups$Species == "Fish_Large"] <- 0.5

# Check the change
Groups[Groups$Type == "Fish", c("Species", "Type", "repro_on","repro_eff", "ZSpre","ZSexp")]
```

Now create an environmental data time-series using the helper function. This time-series uses a constant sea surface temperature (`sst`) and chlorophyll *a*  (`chl`) with a 0.1 yr-1 timestep (`dt`). 

```{r}
env_data_vvlow <- createInputParams(time = seq(0, 400, by = 0.1) ,
                              sst = 15,
                              chl = .01)

env_data_vlow <- createInputParams(time = seq(0, 400, by = 0.1) ,
                              sst = 15,
                              chl = .05)

env_data_low <- createInputParams(time = seq(0, 400, by = 0.1) ,
                              sst = 15,
                              chl = .1)

env_data_med <- createInputParams(time = seq(0, 400, by = 0.1) ,
                              sst = 15,
                              chl = .5)

env_data_high <- createInputParams(time = seq(0, 400, by = 0.1) ,
                              sst = 15,
                              chl = 1)
```

```{r}
env_seasonal_vvlow_stoch <- createEnviroData(
  n_years = 400,
  dt = 0.1,
  base_sst = 15,
  base_chl = 0.01,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5,
  stochastic = TRUE, seed = 123, sst_noise_sd = 0.3, chl_noise_cv = 0.25, ar1_phi = 0.6
)

env_data_vvlow_seasonal_stoch <- createInputParams(
  time = env_seasonal_vvlow_stoch$time,
  sst = env_seasonal_vvlow_stoch$sst,
  chl = env_seasonal_vvlow_stoch$chl
)

env_seasonal_vlow_stoch <- createEnviroData(
  n_years = 400,
  dt = 0.1,
  base_sst = 15,
  base_chl = 0.05,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5, 
  stochastic = TRUE, seed = 123, sst_noise_sd = 0.3, chl_noise_cv = 0.25, ar1_phi = 0.6
)

env_data_vlow_seasonal_stoch <- createInputParams(
  time = env_seasonal_vlow_stoch$time,
  sst = env_seasonal_vlow_stoch$sst,
  chl = env_seasonal_vlow_stoch$chl
)

env_seasonal_low_stoch <- createEnviroData(
  n_years = 400,
  dt = 0.1,
  base_sst = 15,
  base_chl = 0.1,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5, 
  stochastic = TRUE, seed = 123, sst_noise_sd = 0.3, chl_noise_cv = 0.25, ar1_phi = 0.6
)

env_data_low_seasonal_stoch <- createInputParams(
  time = env_seasonal_low_stoch$time,
  sst = env_seasonal_low_stoch$sst,
  chl = env_seasonal_low_stoch$chl
)

env_seasonal_med_stoch <- createEnviroData(
  n_years = 400,
  dt = 0.1,
  base_sst = 15,
  base_chl = 0.5,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5, 
  stochastic = TRUE, seed = 123, sst_noise_sd = 0.3, chl_noise_cv = 0.25, ar1_phi = 0.6
)

env_data_med_seasonal_stoch <- createInputParams(
  time = env_seasonal_med_stoch$time,
  sst = env_seasonal_med_stoch$sst,
  chl = env_seasonal_med_stoch$chl
)

# Create seasonal environmental conditions
env_seasonal_high_stoch <- createEnviroData(
  n_years = 400,
  dt = 0.1,
  base_sst = 15,
  base_chl = 1.0,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5, 
  stochastic = TRUE, seed = 123, sst_noise_sd = 0.3, chl_noise_cv = 0.25, ar1_phi = 0.6
)

env_data_high_seasonal_stoch <- createInputParams(
  time = env_seasonal_high_stoch$time,
  sst = env_seasonal_high_stoch$sst,
  chl = env_seasonal_high_stoch$chl
)
```



```{r}
env_seasonal_vvlow <- createEnviroData(
  n_years = 400,
  dt = 0.1,
  base_sst = 15,
  base_chl = .01,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5
)

env_data_vvlow_seasonal <- createInputParams(
  time = env_seasonal_vvlow$time,
  sst = env_seasonal_vvlow$sst,
  chl = env_seasonal_vvlow$chl
)

env_seasonal_vlow <- createEnviroData(
  n_years = 400,
  dt = 0.1,
  base_sst = 15,
  base_chl = 0.05,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5
)

env_data_vlow_seasonal <- createInputParams(
  time = env_seasonal_vlow$time,
  sst = env_seasonal_vlow$sst,
  chl = env_seasonal_vlow$chl
)

env_seasonal_low <- createEnviroData(
  n_years = 400,
  dt = 0.1,
  base_sst = 15,
  base_chl = 0.1,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5
)

env_data_low_seasonal <- createInputParams(
  time = env_seasonal_low$time,
  sst = env_seasonal_low$sst,
  chl = env_seasonal_low$chl
)

env_seasonal_med <- createEnviroData(
  n_years = 400,
  dt = 0.1,
  base_sst = 15,
  base_chl = 0.5,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5
)

env_data_med_seasonal <- createInputParams(
  time = env_seasonal_med$time,
  sst = env_seasonal_med$sst,
  chl = env_seasonal_med$chl
)

# Create seasonal environmental conditions
env_seasonal_high <- createEnviroData(
  n_years = 400,
  dt = 0.1,
  base_sst = 15,
  base_chl = 1.0,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5
)

env_data_high_seasonal <- createInputParams(
  time = env_seasonal_high$time,
  sst = env_seasonal_high$sst,
  chl = env_seasonal_high$chl
)
```

We can look at the environment data and check everything is ok with: 

```{r}
plotEnvironment(env_data)
```


```{r}
mdl_vvlow_seas <- zoomss_model(input_params = env_data_vvlow_seasonal, Groups = Groups, isave = 2)
mdl_vlow_seas <- zoomss_model(input_params = env_data_vlow_seasonal, Groups = Groups, isave = 2)
mdl_low_seas <- zoomss_model(input_params = env_data_low_seasonal, Groups = Groups, isave = 2)
mdl_med_seas <- zoomss_model(input_params = env_data_med_seasonal, Groups = Groups, isave = 2)
mdl_high_seas <- zoomss_model(input_params = env_data_high_seasonal, Groups = Groups, isave = 2)

mdl_vvlow_seas_stoch <- zoomss_model(input_params = env_data_vvlow_seasonal_stoch, Groups = Groups, isave = 2)
mdl_vlow_seas_stoch <- zoomss_model(input_params = env_data_vlow_seasonal_stoch, Groups = Groups, isave = 2)
mdl_low_seas_stoch <- zoomss_model(input_params = env_data_low_seasonal_stoch, Groups = Groups, isave = 2)
mdl_med_seas_stoch <- zoomss_model(input_params = env_data_med_seasonal_stoch, Groups = Groups, isave = 2)
mdl_high_seas_stoch <- zoomss_model(input_params = env_data_high_seasonal_stoch, Groups = Groups, isave = 2)
```


Now we run ZooMSS and save every `isave` timestep to reduce storage requirements. 
```{r}
mdl_vvlow <- zoomss_model(input_params = env_data_vvlow, Groups = Groups, isave = 2)
mdl_vlow <- zoomss_model(input_params = env_data_vlow, Groups = Groups, isave = 2)
mdl_low <- zoomss_model(input_params = env_data_low, Groups = Groups, isave = 2)
mdl_med <- zoomss_model(input_params = env_data_med, Groups = Groups, isave = 2)
mdl_high <- zoomss_model(input_params = env_data_high, Groups = Groups, isave = 2)
```

## Plotting

The model includes several built-in plotting functions for analysis and visualization.


### Time Series Analysis

```{r}
library(patchwork)

p_biomass_vvlow <- plotTimeSeries(mdl_vvlow, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_vvlow <- plotTimeSeries(mdl_vvlow, by = "biomass", type = "fill") 

p_biomass_vlow <- plotTimeSeries(mdl_vlow, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_vlow <- plotTimeSeries(mdl_vlow, by = "biomass", type = "fill") 

p_biomass_low <- plotTimeSeries(mdl_low, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_low <- plotTimeSeries(mdl_low, by = "biomass", type = "fill") 

p_biomass_med <- plotTimeSeries(mdl_med, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_med <- plotTimeSeries(mdl_med, by = "biomass", type = "fill") 

p_biomass_high <- plotTimeSeries(mdl_high, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_high <- plotTimeSeries(mdl_high, by = "biomass", type = "fill") 

wrap_plots(p_biomass_vvlow, p_fill_vvlow,
           nrow = 2, guides = "collect")

wrap_plots(p_biomass_vlow, p_fill_vlow,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_low, p_fill_low,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_med, p_fill_med,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_high,p_fill_high,
           nrow = 2, guides = "collect")
```
```{r}
library(patchwork)

p_biomass_vvlow_seas <- plotTimeSeries(mdl_vvlow_seas, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_vvlow_seas <- plotTimeSeries(mdl_vvlow_seas, by = "biomass", type = "fill") 

p_biomass_vlow_seas <- plotTimeSeries(mdl_vlow_seas, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_vlow_seas <- plotTimeSeries(mdl_vlow_seas, by = "biomass", type = "fill") 

p_biomass_low_seas <- plotTimeSeries(mdl_low_seas, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_low_seas <- plotTimeSeries(mdl_low_seas, by = "biomass", type = "fill") 

p_biomass_med_seas <- plotTimeSeries(mdl_med_seas, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_med_seas <- plotTimeSeries(mdl_med_seas, by = "biomass", type = "fill") 

p_biomass_high_seas <- plotTimeSeries(mdl_high_seas, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_high_seas <- plotTimeSeries(mdl_high_seas, by = "biomass", type = "fill") 

wrap_plots(p_biomass_vvlow_seas, p_fill_vvlow_seas,
           nrow = 2, guides = "collect")

wrap_plots(p_biomass_vlow_seas, p_fill_vlow_seas,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_low_seas, p_fill_low_seas,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_med_seas, p_fill_med_seas,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_high_seas,p_fill_high_seas,
           nrow = 2, guides = "collect")
```

```{r}
library(patchwork)

p_biomass_vvlow_seas_stoch <- plotTimeSeries(mdl_vvlow_seas_stoch, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_vvlow_seas_stoch <- plotTimeSeries(mdl_vvlow_seas_stoch, by = "biomass", type = "fill") 

p_biomass_vlow_seas_stoch <- plotTimeSeries(mdl_vlow_seas_stoch, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_vlow_seas_stoch <- plotTimeSeries(mdl_vlow_seas_stoch, by = "biomass", type = "fill") 

p_biomass_low_seas_stoch <- plotTimeSeries(mdl_low_seas_stoch, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_low_seas_stoch <- plotTimeSeries(mdl_low_seas_stoch, by = "biomass", type = "fill") 

p_biomass_med_seas_stoch <- plotTimeSeries(mdl_med_seas_stoch, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_med_seas_stoch <- plotTimeSeries(mdl_med_seas_stoch, by = "biomass", type = "fill") 

p_biomass_high_seas_stoch <- plotTimeSeries(mdl_high_seas_stoch, by = "biomass", transform = "log10") +
  theme(legend.position = "none")
p_fill_high_seas_stoch <- plotTimeSeries(mdl_high_seas_stoch, by = "biomass", type = "fill") 

wrap_plots(p_biomass_vvlow_seas_stoch, p_fill_vvlow_seas_stoch,
           nrow = 2, guides = "collect")

wrap_plots(p_biomass_vlow_seas_stoch, p_fill_vlow_seas_stoch,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_low_seas_stoch, p_fill_low_seas_stoch,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_med_seas_stoch, p_fill_med_seas_stoch,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_high_seas_stoch,p_fill_high_seas_stoch,
           nrow = 2, guides = "collect")
```


```{r}
library(patchwork)

p_biomass <- plotTimeSeries(mdl, by = "biomass", transform = "log10") +
  theme(legend.position = "none")

p_fill <- plotTimeSeries(mdl, by = "biomass", type = "fill") 

wrap_plots(p_biomass, p_fill, nrow = 2, guides = "collect")
```

```{r}
library(patchwork)

p_biomass <- plotTimeSeries(mdl, by = "biomass", transform = "log10") +
  theme(legend.position = "none")

p_fill <- plotTimeSeries(mdl, by = "biomass", type = "fill") 

wrap_plots(p_biomass, p_fill, nrow = 2, guides = "collect")
```

These plots display total abundance and mean growth/mortality across all size classes through time.

```{r}

library(patchwork)
p1 <- plotTimeSeries(mdl, by = "abundance", transform = "log10") # Plot abundance time series
p2 <- plotTimeSeries(mdl, by = "growth") # Plot growth rate time series
p3 <- plotTimeSeries(mdl, by = "mortality") # Plot predation mortality time series

wrap_plots(p1, p2, p3, nrow = 3, guides = "collect")

```


We can also plot total biomass through time.

```{r}
p4 <- plotTimeSeries(mdl, by = "biomass", transform = "log10") + theme(legend.position = "none") # Plot biomass 
p5 <- plotTimeSeries(mdl, by = "biomass", type = "stack", transform = "log10") # Plot stacked biomass 
p6 <- plotTimeSeries(mdl, by = "biomass", type = "fill") # Plot proportional stacked biomass 

wrap_plots(p4, p5, p6, nrow = 3, guides = "collect")

```


### Static Plots for a given model time point

Plot mean species-resolved size spectra for the final `n_years`.
```{r}
plotSizeSpectra(mdl, n_years = 10)
```

Plot predator-prey mass ratios for the `idx` timestep
```{r message=FALSE, warning=FALSE}
plotPPMR(mdl, idx = 500) # Plot final timestep
```



```{r}
env_data <- createInputParams(time = seq(0, 300, by = 0.1) ,
                              sst = 15,
                              chl = 1)
```


```{r}
# Create seasonal environmental conditions
env_seasonal <- createEnviroData(
  n_years = 300,
  dt = 0.1,
  base_sst = 15,
  base_chl = 1.0,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5
)

env_seasonal_params <- createInputParams(
  time = env_seasonal$time,
  sst = env_seasonal$sst,
  chl = env_seasonal$chl
)

```






```{r}
# Create seasonal environmental conditions
env_seasonal_noise <- createEnviroData(
  n_years = 300,
  dt = 0.1,
  base_sst = 15,
  base_chl = 1.0,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5,
  stochastic = TRUE,
  sst_noise_sd = 0.5,
  chl_noise_cv = 0.3,
  ar1_phi = 0.7
)

env_seasonal_params_noise <- createInputParams(
  time = env_seasonal_noise$time,
  sst = env_seasonal_noise$sst,
  chl = env_seasonal_noise$chl
)

```



Now we run ZooMSS and save every `isave` timestep to reduce storage requirements. 
```{r}
mdl_seasonal_noise <- zoomss_model(input_params = env_seasonal_params_noise, Groups = Groups, isave = 2)

mdl_seasonal <- zoomss_model(input_params = env_seasonal_params, Groups = Groups, isave = 2)

mdl <- zoomss_model(input_params = env_data, Groups = Groups, isave = 2)
```



```{r}
library(patchwork)

p_biomass_seas_noise <- plotTimeSeries(mdl_seasonal_noise, by = "biomass", transform = "log10") +
  theme(legend.position = "none")

p_fill_seas_noise <- plotTimeSeries(mdl_seasonal_noise, by = "biomass", type = "fill") 

wrap_plots(p_biomass_seas_noise, p_fill_seas_noise, nrow = 2, guides = "collect")
```


```{r}
library(patchwork)

p_biomass_seasonal <- plotTimeSeries(mdl_seasonal, by = "biomass", transform = "log10") +
  theme(legend.position = "none")

p_fill_seasonal <- plotTimeSeries(mdl_seasonal, by = "biomass", type = "fill") 

wrap_plots(p_biomass_seasonal, p_fill_seasonal, nrow = 2, guides = "collect")
```


```{r}
library(patchwork)

p_biomass <- plotTimeSeries(mdl, by = "biomass", transform = "log10") +
  theme(legend.position = "none")

p_fill <- plotTimeSeries(mdl, by = "biomass", type = "fill") 

wrap_plots(p_biomass, p_fill, nrow = 2, guides = "collect")
```

### Time Series Analysis

```{r}
library(patchwork)

p_biomass_vvlow <- plotTimeSeries_exclude(mdl_vvlow, by = "biomass", transform = "log10",exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_vvlow <- plotTimeSeries_exclude(mdl_vvlow, by = "biomass", type = "fill",exclude = "Fish_Med") 

p_biomass_vlow <- plotTimeSeries_exclude(mdl_vlow, by = "biomass", transform = "log10",exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_vlow <- plotTimeSeries_exclude(mdl_vlow, by = "biomass", type = "fill",exclude = "Fish_Med") 

p_biomass_low <- plotTimeSeries_exclude(mdl_low, by = "biomass", transform = "log10",exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_low <- plotTimeSeries_exclude(mdl_low, by = "biomass", type = "fill",exclude = "Fish_Med") 

p_biomass_med <- plotTimeSeries_exclude(mdl_med, by = "biomass", transform = "log10",exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_med <- plotTimeSeries_exclude(mdl_med, by = "biomass", type = "fill",exclude = "Fish_Med") 

p_biomass_high <- plotTimeSeries_exclude(mdl_high, by = "biomass", transform = "log10",exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_high <- plotTimeSeries_exclude(mdl_high, by = "biomass", type = "fill",exclude = "Fish_Med") 

wrap_plots(p_biomass_vvlow, p_fill_vvlow,
           nrow = 2, guides = "collect")

wrap_plots(p_biomass_vlow, p_fill_vlow,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_low, p_fill_low,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_med, p_fill_med,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_high,p_fill_high,
           nrow = 2, guides = "collect")
```


```{r}
library(patchwork)

p_biomass_vvlow <- plotTimeSeries_exclude(
  mdl_vvlow,
  exclude = "Fish_Med",
  by = "biomass",
  transform = "log10"
) + theme(legend.position = "none")

p_biomass_vvlow_seas <- plotTimeSeries_exclude(mdl_vvlow_seas, by = "biomass", transform = "log10", exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_vvlow_seas <- plotTimeSeries_exclude(mdl_vvlow_seas, by = "biomass", type = "fill", exclude = "Fish_Med") 

p_biomass_vlow_seas <- plotTimeSeries_exclude(mdl_vlow_seas, by = "biomass", transform = "log10", exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_vlow_seas <- plotTimeSeries_exclude(mdl_vlow_seas, by = "biomass", type = "fill", exclude = "Fish_Med") 

p_biomass_low_seas <- plotTimeSeries_exclude(mdl_low_seas, by = "biomass", transform = "log10", exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_low_seas <- plotTimeSeries_exclude(mdl_low_seas, by = "biomass", type = "fill", exclude = "Fish_Med") 

p_biomass_med_seas <- plotTimeSeries_exclude(mdl_med_seas, by = "biomass", transform = "log10", exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_med_seas <- plotTimeSeries_exclude(mdl_med_seas, by = "biomass", type = "fill", exclude = "Fish_Med") 

p_biomass_high_seas <- plotTimeSeries_exclude(mdl_high_seas, by = "biomass", transform = "log10", exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_high_seas <- plotTimeSeries_exclude(mdl_high_seas, by = "biomass", type = "fill", exclude = "Fish_Med") 

wrap_plots(p_biomass_vvlow_seas, p_fill_vvlow_seas,
           nrow = 2, guides = "collect")

wrap_plots(p_biomass_vlow_seas, p_fill_vlow_seas,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_low_seas, p_fill_low_seas,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_med_seas, p_fill_med_seas,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_high_seas,p_fill_high_seas,
           nrow = 2, guides = "collect")
```


```{r}
library(patchwork)

p_biomass_vvlow_seas_stoch <- plotTimeSeries_exclude(mdl_vvlow_seas_stoch, by = "biomass", transform = "log10", exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_vvlow_seas_stoch <- plotTimeSeries_exclude(mdl_vvlow_seas_stoch, by = "biomass", type = "fill", exclude = "Fish_Med") 

p_biomass_vlow_seas_stoch <- plotTimeSeries_exclude(mdl_vlow_seas_stoch, by = "biomass", transform = "log10", exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_vlow_seas_stoch <- plotTimeSeries_exclude(mdl_vlow_seas_stoch, by = "biomass", type = "fill", exclude = "Fish_Med") 

p_biomass_low_seas_stoch <- plotTimeSeries_exclude(mdl_low_seas_stoch, by = "biomass", transform = "log10", exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_low_seas_stoch <- plotTimeSeries_exclude(mdl_low_seas_stoch, by = "biomass", type = "fill", exclude = "Fish_Med") 

p_biomass_med_seas_stoch <- plotTimeSeries_exclude(mdl_med_seas_stoch, by = "biomass", transform = "log10", exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_med_seas_stoch <- plotTimeSeries_exclude(mdl_med_seas_stoch, by = "biomass", type = "fill", exclude = "Fish_Med") 

p_biomass_high_seas_stoch <- plotTimeSeries_exclude(mdl_high_seas_stoch, by = "biomass", transform = "log10", exclude = "Fish_Med") +
  theme(legend.position = "none")
p_fill_high_seas_stoch <- plotTimeSeries_exclude(mdl_high_seas_stoch, by = "biomass", type = "fill", exclude = "Fish_Med") 

wrap_plots(p_biomass_vvlow_seas_stoch, p_fill_vvlow_seas_stoch,
           nrow = 2, guides = "collect")

wrap_plots(p_biomass_vlow_seas_stoch, p_fill_vlow_seas_stoch,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_low_seas_stoch, p_fill_low_seas_stoch,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_med_seas_stoch, p_fill_med_seas_stoch,
           nrow = 2, guides = "collect")

wrap_plots(
           p_biomass_high_seas_stoch,p_fill_high_seas_stoch,
           nrow = 2, guides = "collect")
```







```{r}
devtools::load_all()
source("~/GitHub/zoomss_repro/calibration/dual_path_calibration/zoomss_calibration_repro.R")  # adjust path as needed

# --- Does create_seasonal_input survive inside getGroups/apply flow? ---
Groups <- getGroups()
par_test <- generate_lhs_samples(n_samples = 1, seed = 42)
par_vec <- as.numeric(par_test[1, ])
names(par_vec) <- names(par_test)
Groups_mod <- apply_repro_params(par_vec, Groups)

ip <- create_seasonal_input(n_years = 30, dt = 0.1,
                            base_sst = 15, base_chl = 0.1,
                            sst_amplitude = 4, chl_amplitude = 0.5)
cat("About to run model with LHS params + seasonal...\n")
mdl <- zoomss_model(input_params = ip, Groups = Groups_mod, isave = 2)
cat("Survived single run\n")

# --- Does benchmark generation crash? (sequential, no furrr) ---
cat("About to run benchmark manually (1 chl level)...\n")
Groups_leg <- getGroups()
Groups_leg$repro_on[which(Groups_leg$Type == "Fish")] <- 0L
ip2 <- create_seasonal_input(n_years = 30, dt = 0.1,
                             base_sst = 15, base_chl = 1.0,
                             sst_amplitude = 4, chl_amplitude = 0.5)
mdl2 <- zoomss_model(input_params = ip2, Groups = Groups_leg, isave = 2)
cat("Survived benchmark run\n")

# --- Does the furrr parallel path crash? ---
cat("About to test furrr with 1 worker...\n")
future::plan(future::multisession, workers = 1)
res <- furrr::future_map(1, function(i) {
  devtools::load_all(quiet = TRUE)
  ip <- create_seasonal_input(n_years = 10, dt = 0.1,
                              base_sst = 15, base_chl = 1.0,
                              sst_amplitude = 4, chl_amplitude = 0.5)
  Groups <- getGroups()
  mdl <- zoomss_model(input_params = ip, Groups = Groups, isave = 2)
  "OK"
}, .options = furrr::furrr_options(seed = TRUE))
future::plan(future::sequential)
cat("Survived furrr test:", res[[1]], "\n")
```













