---
title: "Getting started with ZooMSS"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    number_sections: false
vignette: >
  %\VignetteIndexEntry{zoomss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r message=FALSE, warning=FALSE}
library(zoomss)
library(ggplot2)
```

## Input Data

ZooMSS requires two sets of input data:

1. **Groups** - Contains all taxa-specific parameter values for each model group, including size ranges and functional group properties.

2. **Environmental data** - A Time-series dataframe with time series of environmental conditions  with `time`, `sst`, and `chl` columns.

## Running the default Model

Get the default published `Groups` dataframe using:


```{r}
Groups <- getGroups()

Groups$repro_eff[Groups$Type == "Fish"] <- 0.1

# Groups$PPMR[Groups$Species == "Fish_Large"] <- 500

# Groups$PPMR[Groups$Species == "Fish_Med"] <- 100

# Groups$FeedWidth[Groups$Species == "Fish_Med"] <- 1.4
# Groups$FeedWidth[Groups$Species == "Fish_Large"] <- 1.3
# Groups$FeedWidth[Groups$Species == "Fish_Small"] <- 1.3

# Groups$Wmat[Groups$Species == "Fish_Med"] <- 1
# Groups$Wmat[Groups$Species == "Fish_Large"] <- 3
# Groups$Wmat[Groups$Species == "Fish_Small"] <- -1

# Groups$ZSpre[Groups$Type == "Fish"] <- 0.1
# Groups$ZSexp[Groups$Type == "Fish"] <- 0.3
# Groups$ZSpre[Groups$Species == "Fish_Med"] <- 0.15
# Groups$ZSpre[Groups$Species == "Fish_Small"] <- 0.1
# 
# Groups$ZSpre[Groups$Species == "Fish_Large"] <- 0.2
# Groups$ZSexp[Groups$Species == "Fish_Large"] <- 0.5

# Check the change
Groups[Groups$Type == "Fish", c("Species", "Type", "repro_on","repro_eff", "ZSpre","ZSexp")]
```

Now create an environmental data time-series using the helper function. This time-series uses a constant sea surface temperature (`sst`) and chlorophyll *a*  (`chl`) with a 0.1 yr-1 timestep (`dt`). 

```{r}
env_data <- createInputParams(time = seq(0, 200, by = 0.1) ,
                              sst = 15,
                              chl = 1)

```

We can look at the environment data and check everything is ok with: 

```{r}
plotEnvironment(env_data)
```


Now we run ZooMSS and save every `isave` timestep to reduce storage requirements. 
```{r}
mdl <- zoomss_model(input_params = env_data, Groups = Groups, isave = 2)

```

## Plotting

The model includes several built-in plotting functions for analysis and visualization.


### Time Series Analysis

```{r}
library(patchwork)

p_biomass <- plotTimeSeries(mdl, by = "biomass", transform = "log10") +
  theme(legend.position = "none")

p_fill <- plotTimeSeries(mdl, by = "biomass", type = "fill") 

wrap_plots(p_biomass, p_fill, nrow = 2, guides = "collect")
```


These plots display total abundance and mean growth/mortality across all size classes through time.

```{r}

library(patchwork)
p1 <- plotTimeSeries(mdl, by = "abundance", transform = "log10") # Plot abundance time series
p2 <- plotTimeSeries(mdl, by = "growth") # Plot growth rate time series
p3 <- plotTimeSeries(mdl, by = "mortality") # Plot predation mortality time series

wrap_plots(p1, p2, p3, nrow = 3, guides = "collect")

```


We can also plot total biomass through time.

```{r}
p4 <- plotTimeSeries(mdl, by = "biomass", transform = "log10") + theme(legend.position = "none") # Plot biomass 
p5 <- plotTimeSeries(mdl, by = "biomass", type = "stack", transform = "log10") # Plot stacked biomass 
p6 <- plotTimeSeries(mdl, by = "biomass", type = "fill") # Plot proportional stacked biomass 

wrap_plots(p4, p5, p6, nrow = 3, guides = "collect")

```


### Static Plots for a given model time point

Plot mean species-resolved size spectra for the final `n_years`.
```{r}
plotSizeSpectra(mdl, n_years = 10)
```

Plot predator-prey mass ratios for the `idx` timestep
```{r message=FALSE, warning=FALSE}
plotPPMR(mdl, idx = 500) # Plot final timestep
```



```{r}
env_data <- createInputParams(time = seq(0, 300, by = 0.1) ,
                              sst = 15,
                              chl = 1)
```


```{r}
# Create seasonal environmental conditions
env_seasonal <- createEnviroData(
  n_years = 300,
  dt = 0.1,
  base_sst = 15,
  base_chl = 1.0,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5
)

env_seasonal_params <- createInputParams(
  time = env_seasonal$time,
  sst = env_seasonal$sst,
  chl = env_seasonal$chl
)

```






```{r}
# Create seasonal environmental conditions
env_seasonal_noise <- createEnviroData(
  n_years = 300,
  dt = 0.1,
  base_sst = 15,
  base_chl = 1.0,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5,
  stochastic = TRUE,
  sst_noise_sd = 0.5,
  chl_noise_cv = 0.3,
  ar1_phi = 0.7
)

env_seasonal_params_noise <- createInputParams(
  time = env_seasonal_noise$time,
  sst = env_seasonal_noise$sst,
  chl = env_seasonal_noise$chl
)

```



Now we run ZooMSS and save every `isave` timestep to reduce storage requirements. 
```{r}
mdl_seasonal_noise <- zoomss_model(input_params = env_seasonal_params_noise, Groups = Groups, isave = 2)

mdl_seasonal <- zoomss_model(input_params = env_seasonal_params, Groups = Groups, isave = 2)

mdl <- zoomss_model(input_params = env_data, Groups = Groups, isave = 2)
```



```{r}
library(patchwork)

p_biomass_seas_noise <- plotTimeSeries(mdl_seasonal_noise, by = "biomass", transform = "log10") +
  theme(legend.position = "none")

p_fill_seas_noise <- plotTimeSeries(mdl_seasonal_noise, by = "biomass", type = "fill") 

wrap_plots(p_biomass_seas_noise, p_fill_seas_noise, nrow = 2, guides = "collect")
```


```{r}
library(patchwork)

p_biomass_seasonal <- plotTimeSeries(mdl_seasonal, by = "biomass", transform = "log10") +
  theme(legend.position = "none")

p_fill_seasonal <- plotTimeSeries(mdl_seasonal, by = "biomass", type = "fill") 

wrap_plots(p_biomass_seasonal, p_fill_seasonal, nrow = 2, guides = "collect")
```


```{r}
library(patchwork)

p_biomass <- plotTimeSeries(mdl, by = "biomass", transform = "log10") +
  theme(legend.position = "none")

p_fill <- plotTimeSeries(mdl, by = "biomass", type = "fill") 

wrap_plots(p_biomass, p_fill, nrow = 2, guides = "collect")
```








