---
title: "Fishing in ZooMSS"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Fishing in ZooMSS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r message=FALSE, warning=FALSE}
library(zoomss)
library(ggplot2)
```

## Overview

ZooMSS supports effort-driven fishing mortality where:

$$F(w, t) = \text{Effort}(t) \times q \times \text{Selectivity}(w)$$

- **Effort(t)** — time-varying fishing effort (from data, e.g. FishMIP)
- **q** — catchability coefficient (estimated by calibration)
- **Selectivity(w)** — knife-edge: 1 within `[Fmort_W0, Fmort_Wmax]`, 0 otherwise

Predicted catch is calculated as:

$$\text{Catch}(t) = \int F(w,t) \, N(w,t) \, w \, dw$$

## Setting Up Fishing

### 1. Configure Groups with Catchability

Start with the default groups and add catchability coefficients using `setFishingParams()`:

```{r}
Groups <- getGroups()

Groups <- setFishingParams(Groups,
                           q_small = 0.01,
                           q_med   = 0.02,
                           q_large = 0.03)
```

You can also customise the selectivity thresholds (the minimum size at which fish are caught):

```{r}
Groups <- setFishingParams(Groups,
                           q_small = 0.01,  q_med = 0.02,  q_large = 0.03,
                           w_min_small = 0.5, w_min_med = 2.5, w_min_large = 4.5)
```

### 2. Add Effort Time Series

Create environmental data and append effort columns with `addFishingEffort()`:

```{r}
env_data <- createInputParams(
  time = seq(0, 100, by = 0.1),
  sst = 15,
  chl = 0.5
)

# Constant effort across all groups
env_data <- addFishingEffort(env_data,
                             effort_small = 1.0,
                             effort_med   = 1.0,
                             effort_large = 1.0)
```

Effort can also be time-varying — just pass a vector the same length as `time`:

```{r}
n <- nrow(env_data)
# Effort that ramps up linearly over the simulation
ramp_effort <- seq(0, 2, length.out = n)

env_data <- addFishingEffort(env_data,
                             effort_small = ramp_effort,
                             effort_med   = ramp_effort,
                             effort_large = ramp_effort)
```

### 3. Run the Model

```{r}
mdl <- zoomss_model(input_params = env_data, Groups = Groups, isave = 2)
```

## Accessing Fishing Outputs

The model returns two additional output arrays when effort data is provided:

- `mdl$catch` — predicted catch biomass (nsave × 3 fish groups)
- `mdl$Fmort_ts` — fishing mortality rate (nsave × 12 groups × ngrid size classes)

```{r}
dim(mdl$catch)     # [nsave, 3]
dim(mdl$Fmort_ts)  # [nsave, 12, ngrid]
```

## Plotting Results

### Catch Time Series

```{r}
fish_names <- c("Fish_Small", "Fish_Med", "Fish_Large")
fish_colours <- mdl$param$Groups$PlotColour[mdl$param$fish_grps]
names(fish_colours) <- fish_names

catch_df <- data.frame(
  time = rep(mdl$time, times = 3),
  catch = c(mdl$catch[, 1], mdl$catch[, 2], mdl$catch[, 3]),
  group = rep(fish_names, each = length(mdl$time))
)

ggplot(catch_df, aes(x = time, y = catch, colour = group)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Time (years)", y = "Catch (g WW)", colour = "Fish Group",
       title = "Predicted Catch Time Series")
```

### Biomass Under Fishing

```{r}
library(patchwork)

p1 <- plotTimeSeries(mdl, by = "biomass",
                     species = c("Fish_Small", "Fish_Med", "Fish_Large"),
                     transform = "log10") +
  ggtitle("Fish Biomass Under Fishing")

p2 <- plotTimeSeries(mdl, by = "biomass", type = "fill") +
  ggtitle("Community Composition Under Fishing")

wrap_plots(p1, p2, ncol = 1)
```

### Reproduction Under Fishing

```{r}
p3 <- plotReproduction(mdl, by = "SSB") +
  ggtitle("Spawning Stock Biomass")

p4 <- plotReproduction(mdl, by = "recruitment") +
  ggtitle("Recruitment")

wrap_plots(p3, p4, ncol = 1)
```

## Without Fishing (Backward Compatibility)

When no effort columns are present, the model uses the existing static `Fmort` pathway. With `Fmort = 0` (the default), the model runs identically to previous versions:

```{r}
env_nofishing <- createInputParams(
  time = seq(0, 50, by = 0.1),
  sst = 15, chl = 0.5
)

# No effort columns, no fishing — behaves exactly as before
mdl_nofishing <- zoomss_model(input_params = env_nofishing, Groups = getGroups(), isave = 2)
is.null(mdl_nofishing$catch)  # TRUE — no catch output
```
