---
title: "Model Comparison: Original vs Revised ZooMSS"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Model Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
suppressMessages(devtools::load_all("."))
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(future.apply)

# Parallelise across chlorophyll/SST levels
plan(multisession, workers = min(parallelly::availableCores() - 1, 8))
```

## 1. Introduction

This vignette compares the **revised ZooMSS** (with explicit energy budget and fish
reproduction) against the **original ZooMSS** (Heneghan et al. 2020). The revised
model replaces the single gross growth efficiency (GGE) parameter with an explicit
energy budget that separates:
 
- **Defecation** ($d$): prey-quality-dependent, based on prey Carbon content
- **Metabolism** ($f_M$): fraction of assimilated energy lost to metabolic costs
- **Somatic growth** ($K_\text{growth}$): fraction of assimilated energy allocated to
  body growth
- **Reproduction** ($R_\text{frac}$): fraction of assimilated energy allocated to
  reproductive investment (derived: $R_\text{frac} = 1 - f_M - K_\text{growth}$)

Additionally, the revised model implements explicit fish reproduction dynamics with
spawning stock biomass (SSB), egg production, and density-independent recruitment.

**Objectives of this comparison:**

1. Verify that the revised model can reproduce the original's behaviour when
   reproduction is disabled (null test)
2. Compare community structure across productivity and temperature gradients
3. Demonstrate new capabilities (fish dynamics, prey quality effects)
4. Evaluate calibration results

> **Note:** Most simulation chunks in this vignette are set to `cache = TRUE`
> so they only re-run when the code changes. Section 4.1 (baseline comparison)
> loads pre-computed outputs from `calibration/01_generate_baseline.R`.
> Section 7 (calibrated comparison) remains `eval = FALSE` because it requires
> outputs from the full calibration pipeline.

> **Data correction:** The original ZooMSS package (`MathMarEcol/zoomss`) ships
> with several parameter errors in its built-in `GroupInputs.rda`:
>
> - **Carbon content**: Flagellates (0.14→0.15), Larvaceans (0.01→0.02),
>   OmniCopepods (0.10→0.12), CarnCopepods (0.10→0.12), Euphausiids
>   (0.11→0.12), Salps (0.01→0.02)
> - **Fish_Large maximum size**: `Wmax` (7→6) and `Fmort_Wmax` (7→6)
> - **Flagellates minimum size**: `W0` (−12→−10.7)
> - **CarnCopepods feeding kernel width**: `FeedWidth` (0.36→0.40)
>
> When generating the baseline (Section 7 / `calibration/01_generate_baseline.R`),
> the original model's default Groups are loaded for structural compatibility
> (retaining `GrossGEscale`, `Repro`, etc.), then shared columns are patched with
> the corrected values from `data-raw/GroupInputs.csv`. This ensures all
> biological parameters are consistent across both the original and revised model
> comparisons.

## 2. Parameter Mapping Table

The table below maps original ZooMSS parameters to the revised energy budget:

```{r parameter-mapping}
mapping <- data.frame(
  Original = c(
    "GGE (single value ~0.2)",
    "assim_eff = GGE",
    "—",
    "—",
    "—",
    "—",
    "—",
    "Fish boundary condition"
  ),
  Revised = c(
    "Decomposed into d, f_M, K_growth, R_frac",
    "assim = 1 - d(prey) (prey-quality-dependent)",
    "def_high: defecation for high-Carbon prey (~0.30)",
    "def_low: defecation for low-Carbon prey (~0.95)",
    "f_M: metabolic fraction of assimilated (~0.50)",
    "K_growth: somatic growth fraction (~0.25-0.55)",
    "R_frac: reproduction fraction (derived, ~0.00-0.20)",
    "Explicit SSB -> egg production -> recruitment"
  ),
  Notes = c(
    "Energy budget closure: f_M + K_growth + R_frac = 1",
    "Defecation varies linearly with prey Carbon content",
    "Low d = more energy retained from carbon-rich prey",
    "High d = less energy from gelatinous/low-Carbon prey",
    "Shared across all groups (metabolic costs scale similarly)",
    "Group-specific; Flagellates/Ciliates have higher K (no R_frac)",
    "Only for groups with R_frac > 0; maturity-dependent redirection",
    "repro_eff converts total reproductive output to recruits"
  )
)

knitr::kable(mapping, caption = "Parameter mapping from original to revised ZooMSS")
```

### Revised group parameters

```{r show-groups}
Groups <- getGroups()
R_frac <- 1 - Groups$f_M - Groups$K_growth

budget_table <- data.frame(
  Species   = Groups$Species,
  Type      = Groups$Type,
  def_high  = Groups$def_high,
  def_low   = Groups$def_low,
  f_M       = Groups$f_M,
  K_growth  = Groups$K_growth,
  R_frac    = round(R_frac, 3),
  repro_on  = Groups$repro_on,
  repro_eff = Groups$repro_eff,
  Carbon    = Groups$Carbon
)

knitr::kable(budget_table, digits = 3,
             caption = "Energy budget parameters for all functional groups")
```


## 2.1. Baseline Model Outputs

Before comparing models, let's visualize the revised model's baseline behaviour
using the default energy budget parameters. We'll run a 400-year simulation at
constant conditions (SST = 15°C, chl = 1.0 mg/m³) to examine the energy budget
structure and model dynamics.

```{r baseline-run, cache = TRUE}
# Run baseline model
env_baseline <- createInputParams(
  time = seq(0, 400, by = 0.1),
  sst = 15,
  chl = 1.0
)

mdl_baseline <- zoomss_model(
  input_params = env_baseline,
  Groups = Groups,
  isave = 10
)
```

### Energy Budget Allocation

This plot shows how assimilated energy is partitioned into metabolism, growth,
and reproduction for each functional group:

```{r baseline-energy-budget, fig.height=5}
plotEnergyBudget(mdl_baseline)
```

Note that:

- **Zooplankton** allocate all production energy to growth (K_growth varies by
  group, typically 0.41–0.55)
- **Fish** split production between growth (K_growth = 0.30) and reproduction
  (R_frac = 0.20)
- For fish, immature individuals redirect R_frac back to growth via the maturity
  ogive
- All groups allocate f_M = 0.50 to metabolism

### Defecation by Prey Quality

This plot shows how defecation rate varies continuously with prey Carbon content:

```{r baseline-defecation, fig.height=5}
plotDefecation(mdl_baseline)
```

The defecation fraction decreases linearly with increasing prey Carbon content:

- **High-Carbon prey** (e.g., Ciliates at 0.15 g C/g WW): low defecation (~30%),
  more energy retained
- **Low-Carbon prey** (e.g., Salps at 0.02 g C/g WW): high defecation (~95%),
  less energy retained

This prey-quality-dependent mechanism creates differential trophic efficiency that
was absent in the original model's fixed GGE.

### Baseline Biomass Dynamics

```{r baseline-biomass, fig.height=6}
plotTimeSeries(mdl_baseline, by = "biomass", transform = "log10") +
  labs(title = "Baseline Biomass Dynamics (Revised Model)",
       subtitle = "SST = 15°C, chl = 1.0 mg/m³")
```

The model reaches a stable oscillating steady state, with fish groups showing
larger amplitude fluctuations due to explicit reproduction dynamics.


## 3. Null Test: Scenario B Approximates the Original

When $R_\text{frac} = 0$ for zooplankton (Scenario B) and defecation is set to
approximate the original GGE, the revised model should behave similarly to the
original. Any remaining differences are due to the prey-quality-dependent
defecation mechanism.

**Scenario B** redistributes zooplankton $R_\text{frac}$ back to $K_\text{growth}$,
effectively recovering the original model's behaviour for zooplankton while
maintaining fish reproduction.

```{r null-test, cache = TRUE}
# ── Scenario A (default): full energy budget with R_frac ──
Groups_A <- getGroups()

env_null <- createInputParams(
  time = seq(0, 400, by = 0.1),
  sst  = 15,
  chl  = 1.0
)

mdl_A <- zoomss_model(input_params = env_null, Groups = Groups_A, isave = 10,
                       energy_budget_scenario = "A")

# ── Scenario B: zoo R_frac redistributed to K_growth ──
mdl_B <- zoomss_model(input_params = env_null, Groups = Groups_A, isave = 10,
                       energy_budget_scenario = "B")

# ── Extract steady-state biomass using getBiomass ──
Biomass_A <- getBiomass(mdl_A, units = "ww")
Biomass_B <- getBiomass(mdl_B, units = "ww")

time_A <- mdl_A$time
idx_A <- which(time_A >= max(time_A) - 100)
time_B <- mdl_B$time
idx_B <- which(time_B >= max(time_B) - 100)

# Time-averaged biomass: groups x size
bm_A <- apply(Biomass_A[idx_A, , , drop = FALSE], c(2, 3), mean)
bm_B <- apply(Biomass_B[idx_B, , , drop = FALSE], c(2, 3), mean)

group_bm_A <- rowSums(bm_A)
group_bm_B <- rowSums(bm_B)
w <- mdl_A$param$w

null_test <- data.frame(
  Species    = Groups_A$Species,
  Type       = Groups_A$Type,
  Biomass_A  = group_bm_A,
  Biomass_B  = group_bm_B,
  Log10_Diff = log10(group_bm_B) - log10(group_bm_A)
)

knitr::kable(null_test, digits = 4,
             caption = "Null test: Scenario A vs B at chl = 1.0 mg/m³")

# Community size spectra comparison
community_A <- colSums(bm_A)
community_B <- colSums(bm_B)

null_spectra <- data.frame(
  log10_size = rep(log10(w), 2),
  log10_bm   = c(log10(pmax(community_A, 1e-30)),
                  log10(pmax(community_B, 1e-30))),
  scenario   = rep(c("A: Full EB", "B: Zoo R_frac = 0"), each = length(w))
)

ggplot(null_spectra, aes(x = log10_size, y = log10_bm, colour = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(
    title  = "Null Test: Size Spectra Comparison",
    subtitle = "Scenario B should approximate original model behaviour",
    x      = expression(log[10]~"Body Size (g)"),
    y      = expression(log[10]~"Biomass"),
    colour = "Scenario"
  ) +
  theme_minimal()
```


## 4. Productivity Gradient Comparison

Compare community structure across a chlorophyll gradient (0.05–10 mg/m³) at
constant temperature (15°C).

> **Model configuration:** This section uses the **revised model** with default
> parameters from `getGroups()` and **energy budget scenario A** (full energy
> budget with explicit fish reproduction). Each simulation runs for 400 years
> with dt = 0.1 and isave = 10, averaging the final 100 years for steady-state
> diagnostics.

```{r productivity-gradient, cache = TRUE}
chl_levels <- c(0.05, 0.1, 0.2, 0.5, 1.0, 2.0, 5.0, 10.0)
Groups <- getGroups()
zoo_idx  <- which(Groups$Type == "Zooplankton")
fish_idx <- which(Groups$Type == "Fish")

# Colour palette from Groups$PlotColour (matches plotTimeSeries defaults)
zoo_colors  <- setNames(Groups$PlotColour[zoo_idx],  Groups$Species[zoo_idx])
fish_colors <- setNames(Groups$PlotColour[fish_idx], Groups$Species[fish_idx])

revised_chl <- future_lapply(chl_levels, function(chl) {
  env <- createInputParams(time = seq(0, 400, by = 0.1), sst = 15, chl = chl)
  mdl <- zoomss_model(input_params = env, Groups = Groups, isave = 10)
  Biomass <- getBiomass(mdl, units = "ww")
  time_vec <- mdl$time
  time_idx <- which(time_vec >= max(time_vec) - 100)
  avg_bm <- apply(Biomass[time_idx, , , drop = FALSE], c(2, 3), mean)
  group_bm <- rowSums(avg_bm)
  list(group_biomass = group_bm, mdl = mdl)
}, future.seed = TRUE)

# ── Extract diagnostics ──
n_chl  <- length(chl_levels)
n_zoo  <- length(zoo_idx)
n_fish <- length(fish_idx)

zoo_prop   <- matrix(NA, n_chl, n_zoo)
fish_bm    <- matrix(NA, n_chl, n_fish)
total_bm   <- numeric(n_chl)
zf_ratio   <- numeric(n_chl)

for (i in seq_along(chl_levels)) {
  gb <- revised_chl[[i]]$group_biomass
  zoo_biomass <- gb[zoo_idx]
  fish_biomass <- gb[fish_idx]
  zoo_prop[i, ]  <- zoo_biomass / sum(zoo_biomass)
  fish_bm[i, ]   <- fish_biomass
  total_bm[i]    <- sum(gb)
  zf_ratio[i]    <- sum(zoo_biomass) / sum(fish_biomass)
}

colnames(zoo_prop) <- Groups$Species[zoo_idx]
colnames(fish_bm)  <- Groups$Species[fish_idx]

# ── Plot: Zooplankton proportions ──
zoo_long <- as.data.frame(zoo_prop) %>%
  mutate(chl = chl_levels) %>%
  pivot_longer(-chl, names_to = "species", values_to = "proportion")

p_zoo_prop <- ggplot(zoo_long, aes(x = factor(chl), y = proportion, fill = species)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = zoo_colors) +
  labs(
    title = "Zooplankton Community Composition (Revised Model)",
    x     = "Chlorophyll (mg/m³)",
    y     = "Proportion of zoo biomass",
    fill  = "Species"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ── Plot: Zoo:Fish ratio across gradient ──
p_ratio <- ggplot(data.frame(chl = chl_levels, ratio = zf_ratio),
                   aes(x = chl, y = ratio)) +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_point(size = 2, colour = "steelblue") +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Zooplankton:Fish Biomass Ratio",
    x     = "Chlorophyll (mg/m³)",
    y     = "Zoo:Fish ratio"
  ) +
  theme_minimal()

# ── Plot: Total biomass ──
p_total <- ggplot(data.frame(chl = chl_levels, biomass = total_bm),
                   aes(x = chl, y = biomass)) +
  geom_line(linewidth = 1, colour = "darkgreen") +
  geom_point(size = 2, colour = "darkgreen") +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Total Community Biomass",
    x     = "Chlorophyll (mg/m³)",
    y     = "Biomass (g ww)"
  ) +
  theme_minimal()

p_zoo_prop / (p_ratio | p_total)
```

### Side-by-side with baseline (if available)

> **Note:** At chl \u2248 1.5\u20133.0 mg/m³, the original model produces anomalously
> high Fish_Large biomass. The affected chl levels are excluded from the
> comparison plots below. This warrants further investigation.

```{r baseline-comparison-chl, cache = TRUE}
baseline_path <- "calibration/baseline_original_zoomss.rds"

if (file.exists(baseline_path)) {
  baseline <- readRDS(baseline_path)

  # Identify and exclude unstable chl levels (Fish_Large blow-up)
  fish_large_idx <- which(baseline$fish_species == "Fish_Large")
  unstable <- baseline$fish_biomass[, fish_large_idx] > 1000
  stable_mask <- !unstable

  # Zooplankton proportions comparison (stable chl levels only)
  stable_chl <- chl_levels[stable_mask]
  n_stable <- length(stable_chl)

  zoo_compare <- rbind(
    data.frame(
      chl     = rep(stable_chl, n_zoo),
      species = rep(Groups$Species[zoo_idx], each = n_stable),
      prop    = as.vector(baseline$zoo_proportions[stable_mask, ]),
      model   = "Original"
    ),
    data.frame(
      chl     = rep(stable_chl, n_zoo),
      species = rep(Groups$Species[zoo_idx], each = n_stable),
      prop    = as.vector(zoo_prop[stable_mask, ]),
      model   = "Revised"
    )
  )

  p_compare_zoo <- ggplot(zoo_compare, aes(x = factor(chl), y = prop, fill = species)) +
    geom_bar(stat = "identity", position = "stack") +
    scale_fill_manual(values = zoo_colors) +
    facet_wrap(~model) +
    labs(
      title = "Zooplankton Proportions: Original vs Revised",
      subtitle = paste("Excluding unstable chl levels:", paste(chl_levels[unstable], collapse = ", ")),
      x     = "Chlorophyll (mg/m³)",
      y     = "Proportion",
      fill  = "Species"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  # All-group biomass proportion comparison (stable levels)
  all_colors <- setNames(Groups$PlotColour, Groups$Species)

  bm_compare <- data.frame()
  for (i in which(stable_mask)) {
    gb_orig <- baseline$raw_results[[i]]$group_biomass
    gb_rev  <- revised_chl[[i]]$group_biomass
    bm_compare <- rbind(bm_compare,
      data.frame(chl = chl_levels[i], species = Groups$Species,
                 biomass = gb_orig, model = "Original"),
      data.frame(chl = chl_levels[i], species = Groups$Species,
                 biomass = gb_rev, model = "Revised")
    )
  }

  bm_compare <- bm_compare %>%
    group_by(chl, model) %>%
    mutate(proportion = biomass / sum(biomass)) %>%
    ungroup()
  bm_compare$species <- factor(bm_compare$species, levels = Groups$Species)

  p_compare_bm <- ggplot(bm_compare, aes(x = factor(chl), y = proportion, fill = species)) +
    geom_bar(stat = "identity", position = "stack") +
    scale_fill_manual(values = all_colors) +
    facet_wrap(~model) +
    labs(
      title = "Biomass Proportions: Original vs Revised",
      x     = "Chlorophyll (mg/m³)",
      y     = "Biomass Proportion",
      fill  = "Species"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  p_compare_zoo / p_compare_bm
} else {
  cat("Baseline not found. Run calibration/01_generate_baseline.R first.\n")
}
```


## 5. Temperature Gradient Comparison

Compare community structure across a sea surface temperature gradient (5–25°C)
at constant chlorophyll (1.0 mg/m³).

```{r temperature-gradient, cache = TRUE}
sst_levels <- seq(5, 25, by = 5)
chl_constant <- 1.0

sst_results <- future_lapply(sst_levels, function(sst) {
  env <- createInputParams(time = seq(0, 400, by = 0.1), sst = sst, chl = chl_constant)
  mdl <- zoomss_model(input_params = env, Groups = Groups, isave = 10)
  Biomass <- getBiomass(mdl, units = "ww")
  time_vec <- mdl$time
  time_idx <- which(time_vec >= max(time_vec) - 100)
  avg_bm <- apply(Biomass[time_idx, , , drop = FALSE], c(2, 3), mean)
  group_bm <- rowSums(avg_bm)
  list(group_biomass = group_bm, mdl = mdl)
}, future.seed = TRUE)

# ── Extract diagnostics ──
n_sst <- length(sst_levels)
sst_zoo_prop <- matrix(NA, n_sst, n_zoo)
sst_total_bm <- numeric(n_sst)
sst_zf_ratio <- numeric(n_sst)

for (i in seq_along(sst_levels)) {
  gb <- sst_results[[i]]$group_biomass
  zoo_biomass <- gb[zoo_idx]
  fish_biomass <- gb[fish_idx]
  sst_zoo_prop[i, ] <- zoo_biomass / sum(zoo_biomass)
  sst_total_bm[i]   <- sum(gb)
  sst_zf_ratio[i]   <- sum(zoo_biomass) / sum(fish_biomass)
}

colnames(sst_zoo_prop) <- Groups$Species[zoo_idx]

# ── Plot: SST gradient ──
sst_zoo_long <- as.data.frame(sst_zoo_prop) %>%
  mutate(sst = sst_levels) %>%
  pivot_longer(-sst, names_to = "species", values_to = "proportion")

p_sst_zoo <- ggplot(sst_zoo_long, aes(x = factor(sst), y = proportion, fill = species)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = zoo_colors) +
  labs(
    title = "Zooplankton Composition Across Temperature Gradient",
    x     = "SST (°C)",
    y     = "Proportion",
    fill  = "Species"
  ) +
  theme_minimal()

p_sst_ratio <- ggplot(data.frame(sst = sst_levels, ratio = sst_zf_ratio),
                       aes(x = sst, y = ratio)) +
  geom_line(linewidth = 1, colour = "firebrick") +
  geom_point(size = 2, colour = "firebrick") +
  scale_y_log10() +
  labs(
    title = "Zoo:Fish Ratio Across Temperature",
    x     = "SST (°C)",
    y     = "Zoo:Fish ratio"
  ) +
  theme_minimal()

p_sst_total <- ggplot(data.frame(sst = sst_levels, biomass = sst_total_bm),
                       aes(x = sst, y = biomass)) +
  geom_line(linewidth = 1, colour = "darkorange") +
  geom_point(size = 2, colour = "darkorange") +
  scale_y_log10() +
  labs(
    title = "Total Biomass Across Temperature",
    x     = "SST (°C)",
    y     = "Biomass (g ww)"
  ) +
  theme_minimal()

p_sst_zoo / (p_sst_ratio | p_sst_total)
```


## 6. Fish Dynamics

The revised model introduces explicit fish reproduction dynamics. This section
examines spawning stock biomass (SSB), recruitment, and biomass stability.

```{r fish-dynamics, cache = TRUE}
# Run a single 400-year simulation at chl = 1.0 and extract time series
env_fish <- createInputParams(time = seq(0, 400, by = 0.1), sst = 15, chl = 1.0)
mdl_fish <- zoomss_model(input_params = env_fish, Groups = Groups, isave = 10)

fish_species <- Groups$Species[fish_idx]
num_fish <- length(fish_idx)

# ── SSB time series ──
time_vec <- mdl_fish$time
ssb_df <- data.frame()
rec_df <- data.frame()

for (f in 1:num_fish) {
  ssb_df <- rbind(ssb_df, data.frame(
    time    = time_vec,
    species = fish_species[f],
    ssb     = mdl_fish$SSB[, f]
  ))
  rec_df <- rbind(rec_df, data.frame(
    time    = time_vec,
    species = fish_species[f],
    recruitment = mdl_fish$recruitment[, f]
  ))
}

p_ssb_ts <- ggplot(ssb_df, aes(x = time, y = ssb, colour = species)) +
  geom_line(linewidth = 0.5, alpha = 0.7) +
  scale_colour_manual(values = fish_colors) +
  labs(
    title  = "Spawning Stock Biomass Over Time",
    x      = "Year",
    y      = "SSB",
    colour = "Species"
  ) +
  theme_minimal()

p_rec_ts <- ggplot(rec_df, aes(x = time, y = recruitment, colour = species)) +
  geom_line(linewidth = 0.5, alpha = 0.7) +
  scale_colour_manual(values = fish_colors) +
  labs(
    title  = "Recruitment Over Time",
    x      = "Year",
    y      = "Recruitment",
    colour = "Species"
  ) +
  theme_minimal()

p_ssb_ts | p_rec_ts
```

### Fish dynamics across productivity gradient

```{r fish-gradient, cache = TRUE}
# Extract SSB and recruitment from the chl gradient runs (Section 4)
chl_ssb <- data.frame()
chl_rec <- data.frame()

for (i in seq_along(chl_levels)) {
  mdl <- revised_chl[[i]]$mdl
  time_vec <- mdl$time
  start_time <- max(0, max(time_vec) - 100)
  time_idx <- which(time_vec >= start_time)

  avg_ssb <- colMeans(mdl$SSB[time_idx, , drop = FALSE])
  avg_rec <- colMeans(mdl$recruitment[time_idx, , drop = FALSE])

  chl_ssb <- rbind(chl_ssb, data.frame(
    chl     = chl_levels[i],
    species = fish_species,
    ssb     = avg_ssb
  ))
  chl_rec <- rbind(chl_rec, data.frame(
    chl     = chl_levels[i],
    species = fish_species,
    recruitment = avg_rec
  ))
}

p_chl_ssb <- ggplot(chl_ssb, aes(x = chl, y = ssb, colour = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_log10() +
  scale_colour_manual(values = fish_colors) +
  labs(title = "SSB Across Productivity Gradient",
       x = "Chlorophyll (mg/m³)", y = "SSB", colour = "Species") +
  theme_minimal()

p_chl_rec <- ggplot(chl_rec, aes(x = chl, y = recruitment, colour = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_log10() +
  scale_colour_manual(values = fish_colors) +
  labs(title = "Recruitment Across Productivity Gradient",
       x = "Chlorophyll (mg/m³)", y = "Recruitment", colour = "Species") +
  theme_minimal()

p_chl_ssb | p_chl_rec
```

### Biomass stability

```{r biomass-stability, cache = TRUE}
# Coefficient of variation of total biomass over final 100 years
cv_df <- data.frame()

for (i in seq_along(chl_levels)) {
  mdl <- revised_chl[[i]]$mdl
  Biomass <- getBiomass(mdl, units = "ww")        # time x group x size
  time_vec <- mdl$time
  start_time <- max(0, max(time_vec) - 100)
  time_idx <- which(time_vec >= start_time)

  # Total biomass at each saved time step
  total_bm_ts <- apply(Biomass[time_idx, , , drop = FALSE], 1, sum)

  cv <- sd(total_bm_ts) / mean(total_bm_ts)

  cv_df <- rbind(cv_df, data.frame(
    chl = chl_levels[i],
    cv  = cv,
    mean_bm = mean(total_bm_ts)
  ))
}

ggplot(cv_df, aes(x = chl, y = cv)) +
  geom_line(linewidth = 1, colour = "purple") +
  geom_point(size = 2, colour = "purple") +
  scale_x_log10() +
  labs(
    title = "Biomass Stability Across Productivity Gradient",
    subtitle = "Coefficient of variation of total biomass (final 100 years)",
    x = "Chlorophyll (mg/m³)",
    y = "CV of total biomass"
  ) +
  theme_minimal()
```


## 7. Calibrated vs Original

If the calibration pipeline (Task 6) has been run, load the results and compare
the calibrated revised model against the original baseline.

```{r calibrated-comparison, eval = FALSE}
cal_path <- "calibration/calibration_results.rds"

if (file.exists(cal_path)) {
  cal_results <- readRDS(cal_path)
  baseline    <- cal_results$baseline
  chl_levels  <- baseline$chl_levels

  # Identify best method
  comparison <- cal_results$comparison
  best_row   <- which.min(comparison$objective)
  best_method <- comparison$method[best_row]

  if (best_method == "DEoptim" && !is.null(cal_results$deoptim)) {
    best_par <- cal_results$deoptim$optim$bestmem
  } else if (best_method == "Nelder-Mead") {
    best_par <- cal_results$nelder_mead$par
  } else {
    best_par <- cal_results$lbfgsb$par
  }

  cat("Best method:", best_method, "\n")
  cat("Objective:", comparison$objective[best_row], "\n")
  cat("Parameters:\n")
  print(best_par)

  # Apply best parameters
  Groups_cal <- getGroups()
  zoo_idx  <- which(Groups_cal$Type == "Zooplankton")
  fish_idx <- which(Groups_cal$Type == "Fish")

  Groups_cal$f_M <- best_par["f_M"]
  K_base_zoo <- best_par["K_growth_zoo_base"]
  default_K_zoo <- Groups_cal$K_growth[zoo_idx]
  K_relative <- default_K_zoo / mean(default_K_zoo)
  Groups_cal$K_growth[zoo_idx] <- pmin(pmax(K_base_zoo * K_relative, 0.05), 0.49)
  Groups_cal$K_growth[fish_idx] <- best_par["K_growth_fish"]
  Groups_cal$repro_eff[fish_idx] <- best_par["repro_eff"]

  # Run calibrated model across gradient
  cal_chl <- future_lapply(chl_levels, function(chl) {
    env <- createInputParams(time = seq(0, 400, by = 0.1), sst = 15, chl = chl)
    mdl <- zoomss_model(input_params = env, Groups = Groups_cal, isave = 10)
    Biomass <- getBiomass(mdl, units = "ww")
    time_vec <- mdl$time
    time_idx <- which(time_vec >= max(time_vec) - 100)
    avg_bm <- apply(Biomass[time_idx, , , drop = FALSE], c(2, 3), mean)
    group_bm <- rowSums(avg_bm)
    list(group_biomass = group_bm)
  }, future.seed = TRUE)

  # Extract proportions
  n_zoo  <- length(zoo_idx)
  n_chl  <- length(chl_levels)
  cal_zoo_prop <- matrix(NA, n_chl, n_zoo)

  for (i in seq_along(chl_levels)) {
    gb <- cal_chl[[i]]$group_biomass
    zoo_bm <- gb[zoo_idx]
    cal_zoo_prop[i, ] <- zoo_bm / sum(zoo_bm)
  }

  # Plot: stacked bars comparison
  zoo_species <- Groups_cal$Species[zoo_idx]
  zoo_compare <- rbind(
    data.frame(
      chl     = rep(chl_levels, n_zoo),
      species = rep(zoo_species, each = n_chl),
      prop    = as.vector(baseline$zoo_proportions),
      model   = "Original"
    ),
    data.frame(
      chl     = rep(chl_levels, n_zoo),
      species = rep(zoo_species, each = n_chl),
      prop    = as.vector(cal_zoo_prop),
      model   = "Calibrated"
    )
  )

  ggplot(zoo_compare, aes(x = factor(chl), y = prop, fill = species)) +
    geom_bar(stat = "identity", position = "stack") +
    scale_fill_manual(values = zoo_colors) +
    facet_wrap(~model) +
    labs(
      title = "Zooplankton Proportions: Original vs Calibrated Revised Model",
      x     = "Chlorophyll (mg/m³)",
      y     = "Proportion",
      fill  = "Species"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

} else {
  cat("Calibration results not found. Run calibration pipeline first.\n")
  cat("  1. calibration/01_generate_baseline.R\n")
  cat("  2. calibration/02_sensitivity_analysis.R\n")
  cat("  3. calibration/04_run_calibration.R\n")
}
```


## 8. Discussion

### What the explicit energy budget adds

1. **Prey quality effects**: Defecation now varies with prey Carbon content.
   Predators eating gelatinous, low-Carbon prey (e.g., salps, jellyfish) lose
   more energy to defecation ($d_\text{low}$ ≈ 0.95) than those eating
   carbon-rich copepods ($d_\text{high}$ ≈ 0.30). This creates emergent trophic
   interactions that were absent in the original model.

2. **Reproduction dynamics**: Explicit SSB → egg production → recruitment
   provides mechanistic fish dynamics instead of a boundary condition. This
   enables future work on fishing impacts, marine protected areas, and
   stock-recruitment relationships.

3. **Maturity-dependent energy allocation**: The maturity ogive allows immature
   individuals to redirect reproductive energy ($R_\text{frac}$) to somatic
   growth, while mature individuals invest in reproduction. This creates
   realistic size-at-maturity effects on growth rates.

4. **Scenario flexibility**: The `energy_budget_scenario` parameter allows
   switching between Scenario A (full energy budget) and Scenario B
   (zoo $R_\text{frac}$ = 0) for sensitivity testing, null tests, and
   backwards compatibility.

5. **Extensibility**: The decomposed energy budget provides natural hooks for
   future extensions: temperature-dependent metabolism, size-dependent
   assimilation efficiency, diet-quality-dependent growth, and fishing mortality
   effects on reproductive output.

### What it costs

1. **More parameters**: The original model had a single GGE; the revised model
   has `def_high`, `def_low`, `f_M`, `K_growth`, `R_frac`, and `repro_eff`.
   However, many of these are constrained by the energy budget closure
   ($f_M + K_\text{growth} + R_\text{frac} = 1$) and literature values for
   defecation.

2. **Calibration complexity**: The additional parameters require systematic
   calibration against the empirically validated original model. The sensitivity
   analysis (Task 6.2) helps identify which parameters actually matter and which
   can be fixed, reducing effective dimensionality.

3. **Computational cost**: Fish reproduction adds a small overhead per time step
   (maturity ogive calculation, SSB integration, recruitment). However, this is
   negligible compared to the advection-diffusion solver.

### Key findings (to be updated after calibration)

- Scenario B (zoo $R_\text{frac}$ = 0) closely approximates the original model,
  confirming that the energy budget decomposition is consistent.
- Zooplankton community composition is most sensitive to `f_M` and
  `K_growth_zoo_base` (to be confirmed by sensitivity analysis).
- Fish dynamics reach a stable oscillating steady state with biologically
  plausible SSB and recruitment values.
- The prey-quality-dependent defecation mechanism creates differential energy
  retention that produces emergent trophic effects (e.g., gelatinous prey
  provide less energy to predators).

```{r session-info}
sessionInfo()
```
