---
title: "ZooMSS Model Comparison: Dual-Pathway Growth and Fish Reproduction vs. Legacy Model"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Model Comparison: Dual-Pathway vs Legacy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r message=FALSE, warning=FALSE}
library(zoomss)
library(ggplot2)
library(patchwork)
library(dplyr)
```

## Introduction

This vignette compares two versions of the ZooMSS model:

- **Legacy model** — the original ZooMSS formulation, which uses a single gross growth efficiency (GGE) pathway for all groups and a simple proportional boundary condition at the minimum fish size class (no explicit reproduction). This behaviour is reproduced exactly by setting `repro_on = 0` for all fish groups.
- **Dual-pathway model** — the updated formulation with an explicit energy budget for fish (separating defecation, metabolism, somatic growth, and reproduction), a maturity ogive, and dynamic recruitment driven by spawning-stock biomass (SSB). This is the default when `repro_on = 1` for fish groups.

The key differences between the two versions are:

| Feature | Legacy (`repro_on = 0`) | Dual-pathway (`repro_on = 1`) |
|---|---|---|
| Zooplankton growth | GGE × Carbon (prey) | Unchanged |
| Fish growth | GGE × Carbon (prey) | $(1 - D_{prey}) \times K_{growth}$ |
| Fish reproduction | None (proportional BC) | SSB → Recruitment flux |
| Maturity ogive | Not used | Logistic; $\psi(w, W_{mat})$ |
| Energy partitioning | Implicit | $f_M + K_{growth} + R_{frac} = 1$ |
| Prey-quality effects | Not fish-specific | Prey Carbon sets defecation |

Throughout, both models run under identical environmental forcing so that differences arise solely from the biological formulation.

## Helper Functions

```{r}
# Thin wrapper to label model output
label_run <- function(mdl, label) {
  mdl$label <- label
  mdl
}

# Extract total biomass per group over time for one model
extract_biomass_ts <- function(mdl) {
  ngrps <- mdl$param$ngrps
  nsave <- length(mdl$time)
  species <- mdl$param$Groups$Species

  bm <- matrix(NA, nrow = nsave, ncol = ngrps)
  for (g in 1:ngrps) {
    bm[, g] <- apply(mdl$abundance[, g, , drop = FALSE], 1, function(row) {
      sum(row * mdl$param$w * mdl$param$dx)
    })
  }
  df <- data.frame(
    time    = rep(mdl$time, times = ngrps),
    biomass = as.vector(bm),
    species = rep(species, each = nsave),
    type    = rep(mdl$param$Groups$Type, each = nsave)
  )
  df$species <- factor(df$species, levels = species)
  df
}

# Bind biomass from two models with a version label
bind_biomass <- function(mdl_legacy, mdl_new, groups = NULL) {
  df_leg <- extract_biomass_ts(mdl_legacy) %>% mutate(version = "Legacy")
  df_new <- extract_biomass_ts(mdl_new)    %>% mutate(version = "Dual-pathway")

  df <- bind_rows(df_leg, df_new)
  if (!is.null(groups)) df <- filter(df, species %in% groups)
  df$version <- factor(df$version, levels = c("Legacy", "Dual-pathway"))
  df
}

# Consistent colour scales
get_all_colours <- function(mdl) {
  ac <- mdl$param$Groups$PlotColour
  names(ac) <- mdl$param$Groups$Species
  ac
}

get_fish_colours <- function(mdl) {
  fc <- mdl$param$Groups$PlotColour[mdl$param$fish_grps]
  names(fc) <- mdl$param$Groups$Species[mdl$param$fish_grps]
  fc
}

fish_names <- c("Fish_Small", "Fish_Med", "Fish_Large")
```

## Part 1: Setting Up Legacy and Dual-Pathway Runs

The legacy behaviour is reproduced by loading the default groups and setting `repro_on = 0` for every fish group. All other parameters are held identical.

```{r}
# ---- Shared environmental forcing ----
env_base <- createInputParams(
  time = seq(0, 200, by = 0.1),
  sst  = 15,
  chl  = 1.0
)

# ---- LEGACY groups: reproduction disabled for all fish ----
Groups_legacy <- getGroups()
Groups_legacy$repro_on[Groups_legacy$Type == "Fish"] <- 0

mdl_legacy <- zoomss_model(input_params = env_base,
                           Groups       = Groups_legacy,
                           isave        = 5)
mdl_legacy <- label_run(mdl_legacy, "Legacy")

# ---- DUAL-PATHWAY groups: default (repro_on = 1 for fish already) ----
Groups_new <- getGroups()
# Confirm repro_on = 1 for fish (default)
stopifnot(all(Groups_new$repro_on[Groups_new$Type == "Fish"] == 1))

mdl_new <- zoomss_model(input_params = env_base,
                        Groups       = Groups_new,
                        isave        = 5)
mdl_new <- label_run(mdl_new, "Dual-pathway")
```

A quick sanity check: `repro_on` status for both runs.

```{r}
data.frame(
  Species  = Groups_legacy$Species,
  Type     = Groups_legacy$Type,
  repro_on_legacy     = Groups_legacy$repro_on,
  repro_on_dual       = Groups_new$repro_on
)
```

## Part 2: Equilibrium Biomass Comparison

The first question is whether the two formulations reach a similar equilibrium, and whether the new fish energy budget redistributes biomass within fish groups.

### All Groups

```{r fig.height=10}
all_cols <- get_all_colours(mdl_new)

bm_all <- bind_biomass(mdl_legacy, mdl_new)

ggplot(bm_all %>% filter(biomass > 0),
       aes(x = time, y = biomass, colour = species)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ version, ncol = 1) +
  scale_colour_manual(values = all_cols) +
  scale_y_log10() +
  scale_x_continuous(expand = c(0, 0)) +
  theme_bw() +
  labs(x = "Time (years)", y = "Biomass (g WW, log scale)",
       colour = "Group",
       title = "All Groups — Biomass Time Series",
       subtitle = "Legacy (top) vs Dual-pathway (bottom)")
```

### Fish Groups Only — Side-by-Side

```{r fig.height=6}
bm_fish <- bind_biomass(mdl_legacy, mdl_new, groups = fish_names)
fish_cols <- get_fish_colours(mdl_new)

ggplot(bm_fish %>% filter(biomass > 0, time > 50),
       aes(x = time, y = biomass, colour = species, linetype = version)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = fish_cols) +
  scale_linetype_manual(values = c("Legacy" = "dashed", "Dual-pathway" = "solid")) +
  scale_y_log10() +
  scale_x_continuous(expand = c(0, 0)) +
  theme_bw() +
  labs(x = "Time (years)", y = "Biomass (g WW, log scale)",
       colour = "Fish Group", linetype = "Model Version",
       title = "Fish Biomass — Legacy (dashed) vs Dual-pathway (solid)")
```

### Equilibrium Biomass Ratio

The ratio of equilibrium biomass (dual-pathway / legacy) summarises how the energy budget changes fish and zooplankton standing stock. Values > 1 indicate higher biomass under the new formulation; values < 1 indicate lower biomass.

```{r fig.height=5}
# Average final 20% of time series
avg_bm <- function(mdl) {
  ngrps <- mdl$param$ngrps
  nsave <- length(mdl$time)
  final_idx <- max(1, floor(nsave * 0.8)):nsave

  sapply(1:ngrps, function(g) {
    mean(apply(mdl$abundance[final_idx, g, , drop = FALSE], 1, function(row) {
      sum(row * mdl$param$w * mdl$param$dx)
    }))
  })
}

bm_leg_eq <- avg_bm(mdl_legacy)
bm_new_eq <- avg_bm(mdl_new)

ratio_df <- data.frame(
  species = factor(mdl_new$param$Groups$Species,
                   levels = mdl_new$param$Groups$Species),
  type    = mdl_new$param$Groups$Type,
  ratio   = bm_new_eq / bm_leg_eq
)

group_cols <- mdl_new$param$Groups$PlotColour
names(group_cols) <- mdl_new$param$Groups$Species

ggplot(ratio_df, aes(x = species, y = ratio, fill = species)) +
  geom_col() +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey30") +
  scale_fill_manual(values = group_cols, guide = "none") +
  scale_y_continuous(trans = "log10") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = expression("Equilibrium biomass ratio (Dual / Legacy, " * log[10] * " scale)"),
       title = "Biomass Ratio: Dual-pathway / Legacy",
       subtitle = "Values above 1 = higher biomass in dual-pathway model")
```

## Part 3: Size Spectra Comparison

Does the dual-pathway model alter the shape of the community size spectrum? This is important for validating the new formulation against observed marine size spectra.

```{r fig.height=7}
p_spec_leg <- plotSizeSpectra(mdl_legacy, n_years = 20) +
  ggtitle("Legacy — Community Size Spectrum")

p_spec_new <- plotSizeSpectra(mdl_new, n_years = 20) +
  ggtitle("Dual-pathway — Community Size Spectrum")

wrap_plots(p_spec_leg, p_spec_new, ncol = 1, guides = "collect")
```

### Fish Size Spectra Detail

```{r fig.height=7}
p_fish_leg <- plotSizeSpectra(mdl_legacy, by = "biomass", n_years = 20) +
  ggtitle("Legacy — Fish Biomass Spectrum") +
  coord_cartesian(xlim = c(0, max(log10(mdl_legacy$param$w[
    mdl_legacy$param$w_log10 >= min(mdl_legacy$param$Groups$W0[mdl_legacy$param$fish_grps])]))))

p_fish_new <- plotSizeSpectra(mdl_new, by = "biomass", n_years = 20) +
  ggtitle("Dual-pathway — Fish Biomass Spectrum") +
  coord_cartesian(xlim = c(0, max(log10(mdl_new$param$w[
    mdl_new$param$w_log10 >= min(mdl_new$param$Groups$W0[mdl_new$param$fish_grps])]))))

wrap_plots(p_fish_leg, p_fish_new, ncol = 1, guides = "collect")
```

The dual-pathway model typically shows a more structured fish size spectrum because somatic growth is capped at $K_{growth}$ of assimilated energy and reproductive output is explicitly subtracted above the maturation size — effects that the legacy GGE pathway does not resolve.

## Part 4: Fish Growth Rates

The explicit energy budget changes the effective growth rate experienced by fish at different body sizes, particularly around the maturation threshold $W_{mat}$ where energy is progressively redirected from somatic growth to reproduction.

### Growth Rate Time Series

```{r fig.height=6}
p_gg_leg <- plotTimeSeries(mdl_legacy, by = "growth",
                           species = fish_names,
                           transform = "log10") +
  ggtitle("Legacy — Fish Growth Rate") +
  xlim(50, 200)

p_gg_new <- plotTimeSeries(mdl_new, by = "growth",
                           species = fish_names,
                           transform = "log10") +
  ggtitle("Dual-pathway — Fish Growth Rate") +
  xlim(50, 200)

wrap_plots(p_gg_leg, p_gg_new, ncol = 1, guides = "collect")
```

### Growth Rate by Body Size

Averaging the final 20 years and inspecting the growth–size relationship reveals where energy re-allocation around $W_{mat}$ produces a visible kink in the dual-pathway model.

```{r fig.height=7}
p_gspec_leg <- plotGrowthComparison(mdl_legacy, groups = "fish", n_years = 20) +
  ggtitle("Legacy — Fish Growth vs Body Size")

p_gspec_new <- plotGrowthComparison(mdl_new, groups = "fish", n_years = 20) +
  ggtitle("Dual-pathway — Fish Growth vs Body Size")

wrap_plots(p_gspec_leg, p_gspec_new, ncol = 1, guides = "collect")
```

In the legacy model the growth–size relationship is smooth because it follows the GGE × Carbon × prey density product with no maturity threshold. In the dual-pathway model, growth rate declines more steeply above $W_{mat}$ as $R_{frac}$ progressively captures energy that would otherwise support somatic growth.

## Part 5: Reproduction — New Functionality

The legacy model has no explicit reproductive output. These diagnostics are only meaningful for the dual-pathway model but are shown alongside legacy placeholders (all NA) to make the contrast visible.

### Spawning Stock Biomass (SSB)

```{r fig.height=8}
p_ssb_leg <- plotReproduction(mdl_legacy, by = "SSB") +
  ggtitle("Legacy — SSB (boundary condition only, no explicit reproduction)") +
  xlim(50, 200)

p_ssb_new <- plotReproduction(mdl_new, by = "SSB") +
  ggtitle("Dual-pathway — SSB") +
  xlim(50, 200)

wrap_plots(p_ssb_leg, p_ssb_new, ncol = 1, guides = "collect")
```

### Recruitment Flux

```{r fig.height=8}
p_rec_leg <- plotReproduction(mdl_legacy, by = "recruitment") +
  ggtitle("Legacy — Recruitment (not mechanistically modelled)") +
  xlim(50, 200)

p_rec_new <- plotReproduction(mdl_new, by = "recruitment") +
  ggtitle("Dual-pathway — Recruitment") +
  xlim(50, 200)

wrap_plots(p_rec_leg, p_rec_new, ncol = 1, guides = "collect")
```

### Total Reproductive Output

```{r fig.height=8}
p_ro_leg <- plotReproduction(mdl_legacy, by = "repro_output") +
  ggtitle("Legacy — Reproductive Output (not tracked)") +
  xlim(50, 200)

p_ro_new <- plotReproduction(mdl_new, by = "repro_output") +
  ggtitle("Dual-pathway — Reproductive Output") +
  xlim(50, 200)

wrap_plots(p_ro_leg, p_ro_new, ncol = 1, guides = "collect")
```

### Reproductive Investment by Body Size

This shows where along the size spectrum reproductive investment peaks for each fish group under the dual-pathway model. The maturity ogive smoothly transitions this from zero at small sizes to the full $R_{frac}$ fraction above $W_{mat}$.

```{r fig.height=5}
plotReproSizeSpectra(mdl_new, n_years = 20) +
  ggtitle("Dual-pathway — Reproductive Rate by Size (averaged over final 20 years)")
```

## Part 6: Energy Budget Diagnostics (Dual-Pathway Only)

### Energy Budget Allocation

```{r fig.height=5}
plotEnergyBudget(mdl_new) +
  ggtitle("Dual-pathway — Energy Budget Fractions by Group")
```

Note that zooplankton groups all have $f_M = 0$, $K_{growth} = 1$, and $R_{frac} = 0$, meaning the full energy budget partitioning is a fish-only feature. The stacked bar for zooplankton groups therefore shows 100% "Growth (K)" — their effective growth rate is still set by the GGE pathway in the model equations.

### Prey-Dependent Defecation

The dual-pathway model scales defecation with prey Carbon content: high-quality (high-C) prey produces less waste than low-quality (low-C) prey. This is absent in the legacy model.

```{r fig.height=5}
plotDefecation(mdl_new) +
  ggtitle("Dual-pathway — Defecation Rate by Prey Carbon Content")
```

### Assimilation Efficiency Matrix

```{r fig.height=5}
plotAssimilationMatrix(mdl_new) +
  ggtitle("Dual-pathway — Net Growth Efficiency by Predator–Prey Pair")
```

## Part 7: Community Composition

How do the two models differ in the balance of zooplankton vs. fish biomass at equilibrium?

```{r fig.height=7}
make_fill_plot <- function(mdl, title_str) {
  plotTimeSeries(mdl, by = "biomass", type = "fill") +
    ggtitle(title_str) +
    xlim(100, 200)
}

p_fill_leg <- make_fill_plot(mdl_legacy, "Legacy — Community Composition")
p_fill_new <- make_fill_plot(mdl_new,    "Dual-pathway — Community Composition")

wrap_plots(p_fill_leg, p_fill_new, ncol = 1, guides = "collect")
```

## Part 8: Sensitivity to Repro Efficiency

The dual-pathway model introduces `repro_eff` (the egg-to-recruit survival fraction) as a new parameter that has no analogue in the legacy model. Here we explore how varying it shifts equilibrium fish biomass, providing guidance for calibration.

```{r}
repro_eff_levels <- c(0.001, 0.005, 0.01, 0.05, 0.1, 0.5)

results_repro <- list()

for (i in seq_along(repro_eff_levels)) {
  Grps_re <- getGroups()
  Grps_re$repro_eff[Grps_re$Type == "Fish"] <- repro_eff_levels[i]

  env_re <- createInputParams(
    time = seq(0, 200, by = 0.1),
    sst  = 15,
    chl  = 1.0
  )

  results_repro[[i]] <- zoomss_model(
    input_params = env_re,
    Groups       = Grps_re,
    isave        = 5
  )
}
```

```{r fig.height=9}
num_fish <- 3

re_summary <- data.frame(
  repro_eff  = rep(repro_eff_levels, each = num_fish),
  fish_group = rep(fish_names, length(repro_eff_levels)),
  eq_biomass = NA,
  eq_SSB     = NA,
  eq_rec     = NA
)

for (i in seq_along(repro_eff_levels)) {
  mdl_i    <- results_repro[[i]]
  nsave    <- length(mdl_i$time)
  fish_grps <- mdl_i$param$fish_grps
  final_idx <- max(1, floor(nsave * 0.8)):nsave

  for (f in 1:num_fish) {
    fg      <- fish_grps[f]
    row_idx <- (i - 1) * num_fish + f

    re_summary$eq_biomass[row_idx] <- mean(apply(
      mdl_i$abundance[final_idx, fg, , drop = FALSE], 1, function(row) {
        sum(row * mdl_i$param$w * mdl_i$param$dx)
      }))

    re_summary$eq_SSB[row_idx] <- mean(mdl_i$SSB[final_idx, f], na.rm = TRUE)
    re_summary$eq_rec[row_idx] <- mean(mdl_i$recruitment[final_idx, f], na.rm = TRUE)
  }
}

fish_cols_re <- get_fish_colours(results_repro[[1]])

p_bm_re <- ggplot(re_summary,
                  aes(x = repro_eff, y = eq_biomass, colour = fish_group)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3) +
  scale_colour_manual(values = fish_cols_re) +
  scale_x_log10() + scale_y_log10() + theme_bw() +
  labs(x = "Reproductive Efficiency", y = "Equilibrium Biomass (log scale)",
       title = "Biomass vs Reproductive Efficiency", colour = "Fish Group")

p_ssb_re <- ggplot(re_summary,
                   aes(x = repro_eff, y = eq_SSB, colour = fish_group)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3) +
  scale_colour_manual(values = fish_cols_re) +
  scale_x_log10() + scale_y_log10() + theme_bw() +
  labs(x = "Reproductive Efficiency", y = "Equilibrium SSB (log scale)",
       title = "SSB vs Reproductive Efficiency", colour = "Fish Group")

p_rec_re <- ggplot(re_summary,
                   aes(x = repro_eff, y = eq_rec, colour = fish_group)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3) +
  scale_colour_manual(values = fish_cols_re) +
  scale_x_log10() + scale_y_log10() + theme_bw() +
  labs(x = "Reproductive Efficiency", y = "Equilibrium Recruitment (log scale)",
       title = "Recruitment vs Reproductive Efficiency", colour = "Fish Group")

wrap_plots(p_bm_re, p_ssb_re, p_rec_re, ncol = 1, guides = "collect")
```

At low `repro_eff`, the recruitment flux is too small to sustain the population and fish biomass collapses to the proportional boundary-condition floor. Above a threshold value, equilibrium biomass stabilises. This threshold is the minimum `repro_eff` consistent with a self-sustaining fish population and should be used as a lower bound during calibration.

## Part 9: Response to Environmental Forcing

Both models are run with a stepped chlorophyll increase to test whether the dual-pathway model responds differently to a productivity perturbation — a fundamental check for Southern Ocean and FishMIP climate projections.

```{r}
time_step <- seq(0, 300, by = 0.1)
# Chl doubles from year 100 onwards
chl_step <- ifelse(time_step < 100, 0.5, 1.0)

env_step_leg <- createInputParams(time = time_step, sst = 15, chl = chl_step)
env_step_new <- createInputParams(time = time_step, sst = 15, chl = chl_step)

mdl_step_leg <- zoomss_model(input_params = env_step_leg,
                              Groups       = Groups_legacy, isave = 5)
mdl_step_new <- zoomss_model(input_params = env_step_new,
                              Groups       = Groups_new,    isave = 5)
```

```{r fig.height=12}
p_step_leg <- plotTimeSeries(mdl_step_leg, by = "biomass",
                              species = fish_names,
                              transform = "log10") +
  geom_vline(xintercept = 100, linetype = "dashed", colour = "grey40") +
  ggtitle("Legacy — Fish Biomass Response to Chl Step") +
  annotate("text", x = 105, y = -Inf, label = "Chl ×2", vjust = -0.5,
           hjust = 0, size = 3)

p_step_new <- plotTimeSeries(mdl_step_new, by = "biomass",
                              species = fish_names,
                              transform = "log10") +
  geom_vline(xintercept = 100, linetype = "dashed", colour = "grey40") +
  ggtitle("Dual-pathway — Fish Biomass Response to Chl Step") +
  annotate("text", x = 105, y = -Inf, label = "Chl ×2", vjust = -0.5,
           hjust = 0, size = 3)

p_ssb_step <- plotReproduction(mdl_step_new, by = "SSB") +
  geom_vline(xintercept = 100, linetype = "dashed", colour = "grey40") +
  ggtitle("Dual-pathway — SSB Response to Chl Step")

p_rec_step <- plotReproduction(mdl_step_new, by = "recruitment") +
  geom_vline(xintercept = 100, linetype = "dashed", colour = "grey40") +
  ggtitle("Dual-pathway — Recruitment Response to Chl Step")

wrap_plots(p_step_leg, p_step_new, p_ssb_step, p_rec_step, ncol = 1)
```

The dual-pathway model shows an additional lag in the fish biomass response relative to the legacy model because recruitment requires SSB to accumulate before new cohorts can contribute to standing stock. This lag is ecologically realistic and absent from the legacy formulation, where the proportional boundary condition instantly propagates productivity increases through to fish.

## Summary

This vignette has compared the legacy ZooMSS formulation (reproduced with `repro_on = 0` for all fish groups) against the new dual-pathway model (`repro_on = 1`). Key differences observed are:

**Equilibrium state.** Equilibrium fish biomass shifts under the dual-pathway model because the effective growth rate is capped at $K_{growth}$ of assimilated energy (rather than the full GGE × Carbon product), and because the reproductive investment fraction $R_{frac}$ diverts energy above $W_{mat}$. The direction and magnitude of the change are parameter-dependent.

**Size spectrum shape.** The dual-pathway model produces a more structured fish size spectrum with a visible break around the maturation size. This is consistent with DBPM theory and empirical observations of maturity-driven changes in fish growth trajectories.

**Reproductive dynamics.** The legacy model has no mechanistic reproduction: fish are maintained at the minimum size class by a simple proportional boundary condition. The dual-pathway model explicitly tracks SSB, total reproductive output, and the resulting recruitment flux, enabling stock–recruitment analyses and calibration against independent SSB or recruitment data.

**Transient response.** The dual-pathway model introduces a recruitment lag in the fish biomass response to productivity perturbations. This is an important distinction for climate projection work (e.g. FishMIP ISIMIP3a protocols) where the timing of fish responses to forcing matters.

**New calibration targets.** The dual-pathway model introduces `repro_eff` (egg-to-recruit survival) as a parameter with a threshold behaviour: below a critical value, fish populations collapse. This parameter should be constrained using independent data before calibration of other fish parameters.

### Practical Guidance

To reproduce legacy ZooMSS behaviour exactly:

```r
Groups <- getGroups()
Groups$repro_on[Groups$Type == "Fish"] <- 0
mdl <- zoomss_model(input_params, Groups)
```

To use the full dual-pathway model (default):

```r
Groups <- getGroups()  # repro_on = 1 for fish by default
mdl <- zoomss_model(input_params, Groups)
```

## Session Info

```{r}
sessionInfo()
```
