---
title: "Fish Reproduction and Energy Budget in ZooMSS"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Fish Reproduction and Energy Budget}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r message=FALSE, warning=FALSE}
library(zoomss)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(parallel)
```

```{r cache-setup}
# ============================================================================
# MODEL OUTPUT CACHING
# ============================================================================
# Long model runs (>100 years) are cached to avoid re-running on every knit.
# Delete the cache directory to force re-runs.

cache_dir <- here::here("vignettes", "cache")
if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE)

#' Run or load a cached ZooMSS model
#'
#' @param cache_name Character name for the cache file (without extension)
#' @param input_params Input parameters for the model
#' @param Groups Functional groups
#' @param isave Save frequency
#' @param cache_dir Path to cache directory (explicit for parallel safety)
#' @param force_rerun Logical, if TRUE, re-run even if cache exists
run_or_load <- function(cache_name, input_params, Groups, isave = 2,
                        energy_budget_scenario = "A", cache_dir, force_rerun = FALSE) {
  cache_file <- file.path(cache_dir, paste0(cache_name, ".rds"))
  if (file.exists(cache_file) && !force_rerun) {
    message("Loading cached model: ", cache_name)
    return(readRDS(cache_file))
  }
  message("Running model: ", cache_name, " (this may take a while...)")
  mdl <- zoomss_model(
    input_params = input_params,
    Groups = Groups,
    isave = isave,
    energy_budget_scenario = energy_budget_scenario
  )
  tmp_file <- tempfile(tmpdir = cache_dir, fileext = ".rds")
  saveRDS(mdl, tmp_file)
  file.rename(tmp_file, cache_file)
  message("Cached model saved: ", cache_file)
  return(mdl)
}
```

```{r parallel-setup}
# ============================================================================
# PARALLEL COMPUTING SETUP
# ============================================================================
# Gradient experiments (chlorophyll × temperature) run independent models that
# can be parallelised. We use parallel::mclapply which forks the current
# process — this inherits all loaded packages (including devtools::load_all()),
# all objects, and avoids the serialisation issues of multisession backends.
#
# NOTE: mclapply uses forking, which works on Linux and macOS.
# On Windows it falls back to sequential (mc.cores is ignored).

n_cores <- max(1, detectCores() - 1)
message("Parallel backend: mclapply with ", n_cores, " cores")
```

## Introduction

This vignette demonstrates the explicit energy budget and fish reproduction system in ZooMSS. The model now implements a DBPM-style (Dynamic Benthic-Pelagic Model) energy budget that explicitly tracks:
- **Defecation**: Energy lost as faeces (varies with prey quality)
- **Assimilation**: Energy absorbed through the gut wall
- **Metabolism**: Energy spent on maintenance (f_M fraction)
- **Growth**: Energy allocated to somatic growth (K_growth fraction)
- **Reproduction**: Energy allocated to reproductive output (R_frac fraction)

For fish groups with `repro_on = 1`, reproductive investment drives population dynamics through explicit recruitment.

## The Energy Budget Structure

The energy budget follows a two-stage partitioning:

**Stage 1: Ingestion Partitioning**
$$
\text{Ingestion} = \text{Defecation} + \text{Assimilation}
$$

Where defecation varies continuously based on prey Carbon content:
$$
D_{prey} = D_{high} + (D_{low} - D_{high}) \times \left(1 - \frac{C_{prey}}{C_{max}}\right)
$$

This formulation interpolates between low defecation for high-quality (high Carbon) prey and high defecation for low-quality (low Carbon) prey. The approach is based on the Dynamic Benthic-Pelagic Model (DBPM; Blanchard et al. 2009, 2012) which uses two food quality categories (D=0.30 for high quality, D=0.50 for low quality). We extend this to a continuous function using Carbon:wet weight ratio as a proxy for food quality.

**Stage 2: Assimilation Partitioning**
$$
\text{Assimilated} = \text{Metabolism} + \text{Growth} + \text{Reproduction}
$$

Where: $f_M + K + R_{frac} = 1$

## Model Setup

Let's load the default groups and examine the energy budget parameters:

```{r}
Groups <- getGroups()

# Display energy budget parameters
energy_params <- Groups[, c("Species", "Type", "def_high", "def_low",
                            "f_M", "K_growth", "repro_on", "Carbon")]
print(energy_params)
```

Notice that:
- **Zooplankton** have `K_growth = 0.50` and `repro_on = 0` (all production goes to growth)
- **Fish** have `K_growth = 0.36` and `repro_on = 1` (reproduction is enabled)
- `R_frac = 1 - f_M - K_growth = 1 - 0.50 - 0.36 = 0.14` for fish

## Part 1: Constant Environmental Conditions

First, let's run the model with constant temperature and chlorophyll (1 mg m⁻³) to establish baseline reproduction dynamics. We run for 1000 years to ensure the model reaches an oscillating steady-state, and average over the final 500 years for diagnostics.

```{r}
# Create constant environmental conditions - 1000 years
env_constant <- createInputParams(
  time = seq(0, 1000, by = 0.1),
  sst = 15,
  chl = 1.0
)

# Plot the environmental forcing
plotEnvironment(env_constant)
```

```{r}
# Run or load from cache
mdl_constant <- run_or_load("mdl_constant_1000yr", env_constant, Groups, isave = 2, cache_dir = cache_dir)
```

### Visualizing the Energy Budget

Before examining the model dynamics, let's visualize the energy budget structure.

#### Energy Budget Allocation

This plot shows how assimilated energy is partitioned into metabolism, growth, and reproduction for each functional group:

```{r fig.height=5}
plotEnergyBudget(mdl_constant)
```

Note that zooplankton allocate all production energy to growth (K = 0.50). For fish, the allocation depends on maturity: **immature fish** also allocate all production to growth (K + R_frac = 0.50), while **mature fish** split production between growth (K = 0.36) and reproduction (R_frac = 0.14). The transition is governed by a size-based maturity ogive.

#### Defecation by Prey Quality

This plot shows how the defecation fraction experienced by a predator varies depending on the Carbon content of the prey it consumes:

```{r fig.height=5}
plotDefecation(mdl_constant)
```

When a predator consumes high-Carbon prey (e.g. ciliates at 0.15 g C/g WW), a larger proportion of the ingested material is nutritionally useful, resulting in low defecation (~30%). Conversely, when consuming low-Carbon prey such as gelatinous organisms (e.g. jellyfish at 0.005 g C/g WW), the high water content means most of the ingested material passes through as waste, resulting in high defecation (~95%). The defecation fraction scales linearly between these bounds ($D_{high}$ = 0.30 and $D_{low}$ = 0.95) as a function of the prey's Carbon:wet weight ratio relative to the maximum across all groups.

#### Prey-Specific Growth Efficiency Matrix

This heatmap shows the net growth efficiency for each predator-prey combination:

```{r fig.height=6}
# ==========================================================================
# FIX: The original plotAssimilationMatrix used mdl$assim_by_prey which only
# contains (1 - D_prey). Since def_high/def_low are the same for all groups,
# every row was identical. The true growth efficiency is (1 - D_prey) * K_pred
# which varies by PREDATOR (since K_growth is group-specific).
# ==========================================================================

# Calculate corrected growth efficiency matrix
K_growth_vec <- mdl_constant$param$Groups$K_growth
growth_eff_matrix <- sweep(mdl_constant$assim_by_prey, 1, K_growth_vec, "*")

species <- mdl_constant$param$Groups$Species
carbon <- mdl_constant$param$Groups$Carbon
group_type <- mdl_constant$param$Groups$Type
ngrps <- mdl_constant$param$ngrps

# Order prey by Carbon content (low to high)
prey_order <- species[order(carbon)]

# Order predators by type then Carbon
pred_order <- species[order(group_type, carbon)]

# Convert to long format
assim_long <- expand.grid(predator = species, prey = species)
assim_long$efficiency <- as.vector(growth_eff_matrix)
assim_long$predator <- factor(assim_long$predator, levels = pred_order)
assim_long$prey <- factor(assim_long$prey, levels = prey_order)

ggplot(assim_long, aes(x = prey, y = predator, fill = efficiency)) +
  geom_tile() +
  geom_text(aes(label = round(efficiency, 3)), size = 2.5, colour = "white") +
  scale_fill_viridis_c(option = "plasma", limits = c(0, max(growth_eff_matrix) * 1.1)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Prey Group (ordered by Carbon content)",
       y = "Predator Group",
       fill = "Growth\nEfficiency",
       title = "Prey-Specific Growth Efficiency",
       subtitle = expression("(1 - D"[prey]*") \u00D7 K"[predator]))
```

This matrix displays the growth efficiency for each predator-prey combination, calculated as $(1 - D_{prey}) \times K_{predator}$. Prey are ordered by Carbon content along the x-axis (low Carbon like Larvaceans and Salps on the left, high Carbon like Ciliates on the right). Higher values (brighter colours) indicate that a predator gains more somatic growth per unit prey consumed. Importantly, predators with higher K_growth (e.g., Flagellates/Ciliates at 0.55) show higher growth efficiency **within each column** than those with lower K_growth (e.g., Fish at 0.36 or Euphausiids at 0.25), demonstrating the predator-specific nature of growth efficiency.

### Biomass Dynamics

Let's examine the biomass dynamics over time:

```{r fig.height=9}
# Plot biomass time series for all groups
p1 <- plotTimeSeries(mdl_constant, by = "biomass", transform = "log10") +
  ggtitle("All Groups - Biomass Time Series")

# Focus on fish groups only
p2 <- plotTimeSeries(mdl_constant, by = "biomass",
                     species = c("Fish_Small", "Fish_Med", "Fish_Large"),
                     transform = "log10") +
  ggtitle("Fish Groups - Biomass Time Series")

# Stacked biomass showing proportions
p3 <- plotTimeSeries(mdl_constant, by = "biomass", type = "fill") +
  ggtitle("Relative Biomass Proportions")

wrap_plots(p1, p2, p3, ncol = 1)
```

### Reproduction Metrics

Now let's examine the reproduction metrics for fish:

```{r fig.height=9}
# Spawning Stock Biomass (SSB)
p_ssb <- plotReproduction(mdl_constant, by = "SSB") +
  ggtitle("Spawning Stock Biomass (SSB)")

# Recruitment
p_recruit <- plotReproduction(mdl_constant, by = "recruitment") +
  ggtitle("Recruitment")

# Total reproductive output
p_repro <- plotReproduction(mdl_constant, by = "repro_output") +
  ggtitle("Total Reproductive Output")

wrap_plots(p_ssb, p_recruit, p_repro, ncol = 1)
```

### Stock-Recruitment Relationships

The relationship between Spawning Stock Biomass (SSB) and recruitment is fundamental to understanding fish population dynamics. Let's examine the emergent stock-recruitment relationships from the model:

```{r fig.height=8}
# Extract SSB and recruitment data for stock-recruitment plots
fish_grps <- mdl_constant$param$fish_grps
fish_names <- mdl_constant$param$Groups$Species[fish_grps]
fish_colours <- mdl_constant$param$Groups$PlotColour[fish_grps]
names(fish_colours) <- fish_names

# Use final 500 years of data for steady-state S-R relationships
n_save <- length(mdl_constant$time)
dt_save <- mdl_constant$param$isave * mdl_constant$param$dt
n_500yr <- min(n_save, round(500 / dt_save))
start_idx <- max(1, n_save - n_500yr + 1)

sr_data <- data.frame()
for (f in seq_along(fish_names)) {
  df <- data.frame(
    SSB = mdl_constant$SSB[start_idx:n_save, f],
    Recruitment = mdl_constant$recruitment[start_idx:n_save, f],
    Fish_Group = fish_names[f],
    Time = mdl_constant$time[start_idx:n_save]
  )
  sr_data <- rbind(sr_data, df)
}

sr_data$Fish_Group <- factor(sr_data$Fish_Group, levels = fish_names)

# Stock-recruitment scatter plots
# Points show individual time steps; connecting paths trace the temporal
# trajectory through S-R space (chronological order), revealing the
# oscillatory dynamics as the population cycles between high and low SSB
p_sr <- ggplot(sr_data, aes(x = SSB, y = Recruitment, colour = Fish_Group)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_path(alpha = 0.15, linewidth = 0.3) +
  scale_colour_manual(values = fish_colours) +
  facet_wrap(~Fish_Group, scales = "free") +
  theme_bw() +
  labs(x = "Spawning Stock Biomass (SSB)",
       y = "Recruitment (individuals)",
       title = "Stock-Recruitment Relationships",
       subtitle = "Final 500 years — paths trace temporal trajectory through S-R space")

# SSB vs Recruitment with log axes for comparison
p_sr_log <- ggplot(sr_data %>% filter(SSB > 0 & Recruitment > 0),
                   aes(x = SSB, y = Recruitment, colour = Fish_Group)) +
  geom_point(alpha = 0.3, size = 1) +
  scale_colour_manual(values = fish_colours) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~Fish_Group, scales = "free") +
  theme_bw() +
  labs(x = "SSB (log scale)",
       y = "Recruitment (log scale)",
       title = "Stock-Recruitment (log-log scale)")

wrap_plots(p_sr, p_sr_log, ncol = 1)
```

The stock-recruitment relationships show the emergent dynamics between spawning biomass and recruitment. In the constant environment, the oscillating steady-state creates characteristic trajectories in S-R space. The connecting paths trace the chronological trajectory — each path follows the population forward through time, revealing how the system orbits through S-R space as SSB and recruitment oscillate out of phase. The relationship is driven by the explicit energy budget where mature fish invest R_frac of their assimilated energy into reproduction, which is then converted to recruits via the reproductive efficiency parameter.

### Growth Rate Comparison

Compare growth rates between fish and zooplankton:

```{r fig.height=6}
# Growth rates for zooplankton (higher K_growth = 0.50)
p_zoo <- plotGrowthComparison(mdl_constant, groups = "zooplankton", n_years = 500) +
  ggtitle("Zooplankton Growth Rates")

# Growth rates for fish (lower K_growth = 0.36, with reproduction)
p_fish <- plotGrowthComparison(mdl_constant, groups = "fish", n_years = 500) +
  ggtitle("Fish Growth Rates")

wrap_plots(p_zoo, p_fish, ncol = 1, guides = "collect")
```

### Reproductive Investment by Size

Examine how reproductive investment varies with body size:

```{r}
plotReproSizeSpectra(mdl_constant, n_years = 500) +
  ggtitle("Fish Reproductive Rate by Body Size")
```

The reproductive rate increases with body size within each fish group, reflecting:
1. The maturity ogive (larger individuals are more likely to be mature)
2. The body-size scaling of growth rates

### Size Spectra

```{r fig.height=6}
# Abundance size spectrum
p_abund <- plotSizeSpectra(mdl_constant, by = "abundance", n_years = 500) +
  ggtitle("Abundance Size Spectrum")

# Biomass size spectrum
p_biom <- plotSizeSpectra(mdl_constant, by = "biomass", n_years = 500) +
  ggtitle("Biomass Size Spectrum")

wrap_plots(p_abund, p_biom, ncol = 1, guides = "collect")
```

## Part 2: Seasonal Environmental Forcing

Now let's compare with seasonal environmental conditions to see how reproduction responds to environmental variability. We use the same base chlorophyll (1 mg m⁻³) but with seasonal variation, run for 1000 years.

```{r}
# Create seasonal environmental conditions
env_seasonal <- createEnviroData(
  n_years = 1000,
  dt = 0.1,
  base_sst = 15,
  base_chl = 1.0,
  seasonal = TRUE,
  sst_amplitude = 4,
  chl_amplitude = 0.5
)

env_seasonal_params <- createInputParams(
  time = env_seasonal$time,
  sst = env_seasonal$sst,
  chl = env_seasonal$chl
)

# Plot the seasonal environment
plotEnvironment(env_seasonal_params)
```

```{r}
# Run or load from cache
mdl_seasonal <- run_or_load("mdl_seasonal_1000yr", env_seasonal_params, Groups, isave = 2, cache_dir = cache_dir)
```

### Comparing Constant vs Seasonal: Biomass

```{r fig.height=9}
# Constant environment biomass - focus on years 500-1000
p_const <- plotTimeSeries(mdl_constant, by = "biomass",
                          species = c("Fish_Small", "Fish_Med", "Fish_Large"),
                          transform = "log10") +
  ggtitle("Constant Environment - Fish Biomass") +
  xlim(500, 1000)

# Seasonal environment biomass
p_seas <- plotTimeSeries(mdl_seasonal, by = "biomass",
                         species = c("Fish_Small", "Fish_Med", "Fish_Large"),
                         transform = "log10") +
  ggtitle("Seasonal Environment - Fish Biomass") +
  xlim(500, 1000)

# Show the seasonal forcing for context
p_env <- plotEnvironment(env_seasonal_params) & xlim(500, 1000)

wrap_plots(p_const, p_seas, p_env, ncol = 1, heights = c(1, 1, 1))
```

### Comparing Constant vs Seasonal: Reproduction

```{r fig.height=9}
# SSB comparison
p_ssb_const <- plotReproduction(mdl_constant, by = "SSB") +
  ggtitle("Constant - SSB") + xlim(500, 1000)

p_ssb_seas <- plotReproduction(mdl_seasonal, by = "SSB") +
  ggtitle("Seasonal - SSB") + xlim(500, 1000)

# Recruitment comparison
p_rec_const <- plotReproduction(mdl_constant, by = "recruitment") +
  ggtitle("Constant - Recruitment") + xlim(500, 1000)

p_rec_seas <- plotReproduction(mdl_seasonal, by = "recruitment") +
  ggtitle("Seasonal - Recruitment") + xlim(500, 1000)

wrap_plots(p_ssb_const, p_ssb_seas, p_rec_const, p_rec_seas,
           ncol = 2, guides = "collect")
```

### Stock-Recruitment: Constant vs Seasonal

```{r fig.height=8}
# Build S-R data for both scenarios
build_sr_data <- function(mdl, scenario_name, n_years_avg = 500) {
  fish_grps <- mdl$param$fish_grps
  fish_names <- mdl$param$Groups$Species[fish_grps]
  n_save <- length(mdl$time)
  dt_save <- mdl$param$isave * mdl$param$dt
  n_avg <- min(n_save, round(n_years_avg / dt_save))
  start_idx <- max(1, n_save - n_avg + 1)

  sr <- data.frame()
  for (f in seq_along(fish_names)) {
    df <- data.frame(
      SSB = mdl$SSB[start_idx:n_save, f],
      Recruitment = mdl$recruitment[start_idx:n_save, f],
      Fish_Group = fish_names[f],
      Scenario = scenario_name
    )
    sr <- rbind(sr, df)
  }
  sr$Fish_Group <- factor(sr$Fish_Group, levels = fish_names)
  return(sr)
}

sr_const <- build_sr_data(mdl_constant, "Constant")
sr_seas <- build_sr_data(mdl_seasonal, "Seasonal")
sr_combined <- rbind(sr_const, sr_seas)

ggplot(sr_combined %>% filter(SSB > 0 & Recruitment > 0),
       aes(x = SSB, y = Recruitment, colour = Scenario)) +
  geom_point(alpha = 0.2, size = 0.8) +
  facet_wrap(~Fish_Group, scales = "free") +
  scale_colour_manual(values = c("Constant" = "grey40", "Seasonal" = "coral")) +
  theme_bw() +
  labs(x = "Spawning Stock Biomass (SSB)",
       y = "Recruitment",
       title = "Stock-Recruitment: Constant vs Seasonal Environment",
       subtitle = "Final 500 years of 1000-year simulations")
```

## Part 3: Exploring Chlorophyll Gradients

Let's explore how productivity (chlorophyll concentration) affects reproduction metrics and community composition across a wide gradient from oligotrophic to eutrophic conditions. For the chlorophyll *a* gradient, a distribution of log~10~(Chlorophyll *a*) values were computed at 0.1 intervals between -1.7 and 0.5 log~10~(Chlorophyll *a*) (n = 23) and the mean proportion calculated for each of the bins. We run each scenario for 1000 years to ensure steady-state.

```{r}
# Chlorophyll gradient: log10(chl) from -1.7 to 0.5 at 0.1 intervals (n = 23)
log10_chl_seq <- seq(-1.7, 0.5, by = 0.1)
chl_levels <- 10^log10_chl_seq
```

### Multi-Temperature Chlorophyll Gradient Experiments

We run the chlorophyll gradient at three temperatures (10°C, 15°C, 20°C) to examine how temperature modulates the productivity response.

```{r}
# Run models across chlorophyll gradient at multiple temperatures — IN PARALLEL
sst_variants <- c(10, 15, 20)

# Build task grid: every chl × sst combination (23 × 3 = 69 tasks)
task_grid <- expand.grid(
  chl_idx = seq_along(chl_levels),
  sst = sst_variants
)

# -------------------------------------------------------------------
# STEP 1: Run uncached models in parallel using mclapply (forking).
# Forking inherits the full parent environment — all loaded packages
# (including devtools::load_all()), all objects, no serialisation.
# Workers save to cache and return only the file path.
# -------------------------------------------------------------------
cache_files_chl <- mclapply(seq_len(nrow(task_grid)), function(row) {
  i    <- task_grid$chl_idx[row]
  temp <- task_grid$sst[row]

  cache_name <- sprintf("mdl_chl_log10_%.1f_sst%d_1000yr", log10(chl_levels[i]), temp)
  cache_file <- file.path(cache_dir, paste0(cache_name, ".rds"))

  if (!file.exists(cache_file)) {
    env_chl <- createInputParams(
      time = seq(0, 1000, by = 0.1),
      sst  = temp,
      chl  = chl_levels[i]
    )
    mdl <- zoomss_model(input_params = env_chl, Groups = Groups, isave = 2)
    tmp <- tempfile(tmpdir = cache_dir, fileext = ".rds")
    saveRDS(mdl, tmp)
    file.rename(tmp, cache_file)
  }

  cache_file
}, mc.cores = n_cores)

# Check for errors (mclapply returns error objects on failure)
errors <- sapply(cache_files_chl, inherits, "try-error")
if (any(errors)) {
  warning(sum(errors), " of ", length(errors), " tasks failed. Falling back to sequential for those.")
  for (row in which(errors)) {
    i    <- task_grid$chl_idx[row]
    temp <- task_grid$sst[row]
    cache_name <- sprintf("mdl_chl_log10_%.1f_sst%d_1000yr", log10(chl_levels[i]), temp)
    env_chl <- createInputParams(time = seq(0, 1000, by = 0.1), sst = temp, chl = chl_levels[i])
    run_or_load(cache_name, env_chl, Groups, isave = 2, cache_dir = cache_dir)
  }
}

# -------------------------------------------------------------------
# STEP 2: Load cached results sequentially (fast disk reads).
# -------------------------------------------------------------------
results_chl <- list()
for (temp in sst_variants) {
  temp_char <- as.character(temp)
  results_chl[[temp_char]] <- lapply(seq_along(chl_levels), function(i) {
    cache_name <- sprintf("mdl_chl_log10_%.1f_sst%d_1000yr", log10(chl_levels[i]), temp)
    readRDS(file.path(cache_dir, paste0(cache_name, ".rds")))
  })
}
```

### Fish Biomass, SSB and Recruitment Across the Gradient

```{r fig.height=12}
# Extract fish metrics across chlorophyll gradient for all temperatures
num_fish <- 3
fish_names <- c("Fish_Small", "Fish_Med", "Fish_Large")

fish_summary_all <- data.frame()

for (temp in sst_variants) {
  temp_char <- as.character(temp)
  fish_grps <- results_chl[[temp_char]][[1]]$param$fish_grps

  for (i in seq_along(chl_levels)) {
    mdl <- results_chl[[temp_char]][[i]]
    final_idx <- length(mdl$time)

    # Average biomass over last 500 years
    avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 500)
    fish_biomass <- rowSums(avg_biomass[fish_grps, , drop = FALSE])

    for (f in 1:num_fish) {
      fish_summary_all <- rbind(fish_summary_all, data.frame(
        chl = chl_levels[i],
        sst = temp,
        fish_group = fish_names[f],
        final_biomass = fish_biomass[f],
        final_SSB = mdl$SSB[final_idx, f],
        final_recruitment = mdl$recruitment[final_idx, f]
      ))
    }
  }
}

fish_summary_all$sst_label <- paste0(fish_summary_all$sst, "\u00B0C")

# Get fish colours
fish_colours <- results_chl[["15"]][[1]]$param$Groups$PlotColour[
  results_chl[["15"]][[1]]$param$fish_grps]
names(fish_colours) <- fish_names

# Plot Fish Biomass vs chlorophyll across temperatures
p_biomass <- ggplot(fish_summary_all,
                    aes(x = log10(chl), y = final_biomass, colour = fish_group,
                        linetype = sst_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Biomass",
       title = "Fish Biomass Response to Productivity Gradient",
       colour = "Fish Group", linetype = "Temperature")

# SSB
p_ssb <- ggplot(fish_summary_all,
                aes(x = log10(chl), y = final_SSB, colour = fish_group,
                    linetype = sst_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Spawning Stock Biomass (SSB)",
       title = "Fish SSB Response to Productivity Gradient",
       colour = "Fish Group", linetype = "Temperature")

# Recruitment
p_recruit <- ggplot(fish_summary_all,
                    aes(x = log10(chl), y = final_recruitment, colour = fish_group,
                        linetype = sst_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Recruitment",
       title = "Fish Recruitment Response to Productivity Gradient",
       colour = "Fish Group", linetype = "Temperature")

wrap_plots(p_biomass, p_ssb, p_recruit, ncol = 1, guides = "collect")
```

### Stock-Recruitment Across the Chlorophyll Gradient

```{r fig.height=12, fig.width=6}
# Build S-R data across chlorophyll gradient at 15°C
sr_chl_data <- data.frame()
for (i in seq_along(chl_levels)) {
  mdl <- results_chl[["15"]][[i]]
  sr_temp <- build_sr_data(mdl, paste0("Chl = ", round(chl_levels[i], 3), " mg/m³"), n_years_avg = 500)
  sr_temp$chl <- chl_levels[i]
  sr_chl_data <- rbind(sr_chl_data, sr_temp)
}

ggplot(sr_chl_data %>% filter(SSB > 0 & Recruitment > 0),
       aes(x = SSB, y = Recruitment, colour = log10(chl))) +
  geom_point(alpha = 0.3, size = 0.8) +
  facet_wrap(~Fish_Group, scales = "free", ncol = 1) +
  scale_colour_viridis_c(option = "viridis",
                         name = expression(log[10] * "(Chl " * italic(a) * ")")) +
  theme_bw() +
  labs(x = "Spawning Stock Biomass (SSB)",
       y = "Recruitment",
       title = "Stock-Recruitment Across Productivity Gradient (15\u00B0C)",
       subtitle = "Final 500 years of 1000-year simulations")
```

### Community Composition Across the Gradient

Let's examine how community composition changes across the productivity gradient. Following Heneghan et al., we split the lower trophic levels into two panels: (1) phytoplankton size classes with flagellates and ciliates, and (2) the remaining zooplankton groups.

```{r fig.height=14}
# ============================================================================
# HELPER: Extract community composition with phytoplankton size classes
# ============================================================================
extract_community_composition <- function(results_list, chl_levels, Groups,
                                          n_years_avg = 500) {

  all_species <- Groups$Species
  ngrps <- length(all_species)

  # ---- Zooplankton/Fish biomass ----
  biomass_summary <- data.frame()
  for (i in seq_along(chl_levels)) {
    mdl <- results_list[[i]]
    avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = n_years_avg)
    total_biomass <- rowSums(avg_biomass)

    for (g in 1:ngrps) {
      biomass_summary <- rbind(biomass_summary, data.frame(
        chl = chl_levels[i],
        species = all_species[g],
        biomass = total_biomass[g],
        type = Groups$Type[g]
      ))
    }
  }

  # ---- Phytoplankton size class biomass ----
  phyto_summary <- data.frame()
  for (i in seq_along(chl_levels)) {
    mdl <- results_list[[i]]
    # Use the phytoplankton parameters to calculate size class biomass
    env_row <- data.frame(chl = chl_levels[i])
    phyto_params <- calculatePhytoParam(env_row)

    phyto_summary <- rbind(phyto_summary, data.frame(
      chl = chl_levels[i],
      species = c("PicoPhytoplankton", "NanoPhytoplankton", "MicroPhytoplankton"),
      biomass = c(phyto_params$pico_biom, phyto_params$nano_biom, phyto_params$micro_biom),
      type = "Phytoplankton"
    ))
  }

  return(list(zoo_fish = biomass_summary, phyto = phyto_summary))
}

# ============================================================================
# HELPER: Create split-panel community composition plot
# ============================================================================
plot_split_community <- function(results_list, chl_levels, Groups,
                                 temperature_label, n_years_avg = 500) {

  comp <- extract_community_composition(results_list, chl_levels, Groups,
                                        n_years_avg = n_years_avg)

  # Get model colours
  plot_colours <- Groups$PlotColour
  names(plot_colours) <- Groups$Species

  # Add phytoplankton colours
  phyto_colours <- c(
    "PicoPhytoplankton" = "#c7e9c0",
    "NanoPhytoplankton" = "#74c476",
    "MicroPhytoplankton" = "#238b45"
  )

  # ---- PANEL 1: Phyto size classes + Flagellates + Ciliates ----
  upper_groups <- c("PicoPhytoplankton", "NanoPhytoplankton", "MicroPhytoplankton",
                    "Flagellates", "Ciliates")

  # Combine phyto and small zoo data
  flag_cil <- comp$zoo_fish %>%
    filter(species %in% c("Flagellates", "Ciliates"))
  upper_data <- rbind(comp$phyto, flag_cil)
  upper_data$species <- factor(upper_data$species, levels = upper_groups)

  # Calculate proportions within upper panel
  upper_data <- upper_data %>%
    group_by(chl) %>%
    mutate(proportion = biomass / sum(biomass, na.rm = TRUE)) %>%
    ungroup()

  upper_colours <- c(phyto_colours,
                     plot_colours[c("Flagellates", "Ciliates")])

  p_upper <- ggplot(upper_data,
                    aes(x = log10(chl), y = proportion, fill = species)) +
    geom_area(position = "stack") +
    scale_fill_manual(values = upper_colours, name = "Species") +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_continuous(expand = c(0, 0)) +
    theme_bw() +
    theme(axis.title.x = element_blank(),
          legend.position = "right") +
    labs(y = "Proportion")

  # ---- PANEL 2: Remaining zooplankton (no Flagellates, Ciliates, no Fish) ----
  lower_groups <- c("OmniCopepods", "CarnCopepods", "Euphausiids",
                    "Chaetognaths", "Salps", "Jellyfish", "Larvaceans")

  lower_data <- comp$zoo_fish %>%
    filter(species %in% lower_groups)
  lower_data$species <- factor(lower_data$species, levels = lower_groups)

  # Calculate proportions within lower panel
  lower_data <- lower_data %>%
    group_by(chl) %>%
    mutate(proportion = biomass / sum(biomass, na.rm = TRUE)) %>%
    ungroup()

  lower_colours <- plot_colours[lower_groups]

  p_lower <- ggplot(lower_data,
                    aes(x = log10(chl), y = proportion, fill = species)) +
    geom_area(position = "stack") +
    scale_fill_manual(values = lower_colours, name = "Species") +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_continuous(expand = c(0, 0)) +
    theme_bw() +
    theme(legend.position = "right") +
    labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
         y = "Proportion")

  # Combine with title
  combined <- wrap_plots(p_upper, p_lower, ncol = 1) +
    plot_annotation(title = paste0("Community Composition Across Productivity Gradient (",
                                   temperature_label, ")"))

  return(combined)
}

# ---- Plot community composition at 15°C ----
plot_split_community(results_chl[["15"]], chl_levels, Groups,
                     "15\u00B0C", n_years_avg = 500)
```

```{r fig.height=14}
# ---- Plot community composition at 10°C ----
plot_split_community(results_chl[["10"]], chl_levels, Groups,
                     "10\u00B0C", n_years_avg = 500)
```

```{r fig.height=14}
# ---- Plot community composition at 20°C ----
plot_split_community(results_chl[["20"]], chl_levels, Groups,
                     "20\u00B0C", n_years_avg = 500)
```

### Full Community Composition (Including Fish)

For completeness, here is the full community composition including fish groups across the gradient at 15°C:

```{r fig.height=10}
# Standard full community composition plot (all groups)
all_species <- Groups$Species
ngrps <- length(all_species)
plot_colours <- Groups$PlotColour
names(plot_colours) <- all_species

biomass_full <- data.frame()
for (i in seq_along(chl_levels)) {
  mdl <- results_chl[["15"]][[i]]
  avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 500)
  total_biomass <- rowSums(avg_biomass)

  for (g in 1:ngrps) {
    biomass_full <- rbind(biomass_full, data.frame(
      chl = chl_levels[i],
      species = all_species[g],
      biomass = total_biomass[g],
      type = Groups$Type[g]
    ))
  }
}

biomass_full <- biomass_full %>%
  group_by(chl) %>%
  mutate(proportion = biomass / sum(biomass, na.rm = TRUE)) %>%
  ungroup()

biomass_full$species <- factor(biomass_full$species, levels = all_species)

# Proportion plot - using log10(chl) on x-axis to match other gradient plots
p_prop <- ggplot(biomass_full,
                 aes(x = log10(chl), y = proportion, fill = species)) +
  geom_area(position = "stack") +
  scale_fill_manual(values = plot_colours) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Proportion of Total Biomass",
       title = "Community Composition Across Productivity Gradient (15\u00B0C)",
       fill = "Species")

# Absolute biomass
p_abs <- ggplot(biomass_full, aes(x = log10(chl), y = biomass, colour = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_x_continuous(expand = c(0, 0)) +
  scale_colour_manual(values = plot_colours) +
  theme_bw() +
  labs(x = expression("Chlorophyll " * italic(a) * " (log"[10] * "(mg m"^-3*"))"),
       y = "Biomass",
       title = "Absolute Biomass Across Productivity Gradient (15\u00B0C)",
       colour = "Species")

wrap_plots(p_prop, p_abs, ncol = 1, guides = "collect")
```

## Part 4: Temperature Effects

Explore how temperature affects fish reproduction through metabolic scaling. We run 1000-year simulations at 1 mg m⁻³ chlorophyll across a temperature gradient.

```{r}
# Run models across a temperature gradient — IN PARALLEL
sst_levels <- c(5, 10, 15, 20, 25)

# Step 1: Run uncached models in parallel, save to cache, return file paths
cache_files_sst <- mclapply(seq_along(sst_levels), function(i) {
  cache_name <- sprintf("mdl_sst%d_chl1_1000yr", sst_levels[i])
  cache_file <- file.path(cache_dir, paste0(cache_name, ".rds"))

  if (!file.exists(cache_file)) {
    env_sst <- createInputParams(
      time = seq(0, 1000, by = 0.1),
      sst = sst_levels[i],
      chl = 1.0
    )
    mdl <- zoomss_model(input_params = env_sst, Groups = Groups, isave = 2)
    tmp <- tempfile(tmpdir = cache_dir, fileext = ".rds")
    saveRDS(mdl, tmp)
    file.rename(tmp, cache_file)
  }

  cache_file
}, mc.cores = n_cores)

# Step 2: Load from cache sequentially
results_sst <- lapply(seq_along(sst_levels), function(i) {
  cache_name <- sprintf("mdl_sst%d_chl1_1000yr", sst_levels[i])
  readRDS(file.path(cache_dir, paste0(cache_name, ".rds")))
})
```

### Fish Biomass, SSB and Recruitment Across the Temperature Gradient

```{r fig.height=10}
# Extract fish metrics across temperature gradient
fish_sst_summary <- data.frame()

for (i in seq_along(sst_levels)) {
  mdl <- results_sst[[i]]
  final_idx <- length(mdl$time)
  fish_grps <- mdl$param$fish_grps

  avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 500)
  fish_biomass <- rowSums(avg_biomass[fish_grps, , drop = FALSE])

  for (f in 1:num_fish) {
    fish_sst_summary <- rbind(fish_sst_summary, data.frame(
      sst = sst_levels[i],
      fish_group = fish_names[f],
      final_biomass = fish_biomass[f],
      final_SSB = mdl$SSB[final_idx, f],
      final_recruitment = mdl$recruitment[final_idx, f]
    ))
  }
}

# Biomass vs temperature
p_biomass <- ggplot(fish_sst_summary,
                    aes(x = sst, y = final_biomass, colour = fish_group)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (\u00B0C)", y = "Biomass",
       title = "Fish Biomass Response to Temperature", colour = "Fish Group")

# SSB vs temperature
p_ssb <- ggplot(fish_sst_summary,
                aes(x = sst, y = final_SSB, colour = fish_group)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (\u00B0C)", y = "SSB",
       title = "Fish SSB Response to Temperature", colour = "Fish Group")

# Recruitment vs temperature
p_recruit <- ggplot(fish_sst_summary,
                    aes(x = sst, y = final_recruitment, colour = fish_group)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3) +
  scale_y_log10() +
  scale_colour_manual(values = fish_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (\u00B0C)", y = "Recruitment",
       title = "Fish Recruitment Response to Temperature", colour = "Fish Group")

wrap_plots(p_biomass, p_ssb, p_recruit, ncol = 1, guides = "collect")
```

### Stock-Recruitment Across the Temperature Gradient

```{r fig.height=12, fig.width=6}
# S-R relationships across temperature gradient
sr_sst_data <- data.frame()
for (i in seq_along(sst_levels)) {
  sr_temp <- build_sr_data(results_sst[[i]],
                           paste0(sst_levels[i], "\u00B0C"),
                           n_years_avg = 500)
  sr_temp$sst <- sst_levels[i]
  sr_sst_data <- rbind(sr_sst_data, sr_temp)
}

ggplot(sr_sst_data %>% filter(SSB > 0 & Recruitment > 0),
       aes(x = SSB, y = Recruitment, colour = factor(sst))) +
  geom_point(alpha = 0.3, size = 0.8) +
  facet_wrap(~Fish_Group, scales = "free", ncol = 1) +
  scale_colour_manual(
    values = c("5" = "#2166ac", "10" = "#67a9cf", "15" = "#f7f7f7",
               "20" = "#ef8a62", "25" = "#b2182b"),
    name = "SST (\u00B0C)"
  ) +
  theme_bw() +
  labs(x = "Spawning Stock Biomass (SSB)",
       y = "Recruitment",
       title = "Stock-Recruitment Across Temperature Gradient",
       subtitle = "Final 500 years of 1000-year simulations at 1 mg/m\u00B3 chlorophyll")
```

### Community Composition Across the Temperature Gradient

```{r fig.height=10}
# Community composition across temperature gradient
biomass_sst <- data.frame()
for (i in seq_along(sst_levels)) {
  mdl <- results_sst[[i]]
  avg_biomass <- averageTimeSeries(mdl, var = "biomass", n_years = 500)
  total_biomass <- rowSums(avg_biomass)

  for (g in 1:ngrps) {
    biomass_sst <- rbind(biomass_sst, data.frame(
      sst = sst_levels[i],
      species = all_species[g],
      biomass = total_biomass[g],
      type = Groups$Type[g]
    ))
  }
}

biomass_sst <- biomass_sst %>%
  group_by(sst) %>%
  mutate(proportion = biomass / sum(biomass, na.rm = TRUE)) %>%
  ungroup()

biomass_sst$species <- factor(biomass_sst$species, levels = all_species)

# Proportion plot
p_prop <- ggplot(biomass_sst, aes(x = factor(sst), y = proportion, fill = species)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = plot_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (\u00B0C)",
       y = "Proportion of Total Biomass",
       title = "Community Composition Across Temperature Gradient",
       fill = "Species")

# Absolute biomass
p_abs <- ggplot(biomass_sst, aes(x = sst, y = biomass, colour = species)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_manual(values = plot_colours) +
  theme_bw() +
  labs(x = "Sea Surface Temperature (\u00B0C)",
       y = "Biomass",
       title = "Absolute Biomass Across Temperature Gradient",
       colour = "Species")

wrap_plots(p_prop, p_abs, ncol = 1, guides = "collect")
```

## Summary

This vignette demonstrated the explicit energy budget and fish reproduction system in ZooMSS:

1. **Energy Budget Components**: The model now explicitly tracks defecation, metabolism, growth, and reproduction fractions

2. **Prey Quality Effects**: Defecation varies continuously with prey Carbon content - high-Carbon prey (like ciliates) have low defecation (~30%) while low-Carbon prey (like jellyfish) have high defecation (~50%)

3. **Fish Reproduction**: Fish groups allocate a fraction of assimilated energy (R_frac = 0.14) to reproduction, scaled by a maturity ogive

4. **Recruitment**: New fish enter the population through explicit recruitment based on Spawning Stock Biomass (SSB)

5. **Stock-Recruitment Relationships**: Emergent S-R dynamics arise from the energy budget, with characteristic patterns that vary across environmental gradients

6. **Environmental Sensitivity**: Reproduction metrics respond to:
   - Productivity (chlorophyll concentration)
   - Temperature (through metabolic scaling)
   - Seasonal cycles in environmental conditions

### Key Differences from Zooplankton

| Aspect | Zooplankton | Fish |
|--------|-------------|------|
| K_growth | 0.50 | 0.36 |
| R_frac | 0.00 | 0.14 |
| repro_on | 0 (disabled) | 1 (enabled) |
| Recruitment | Boundary condition | SSB-driven |

**Note on zooplankton R_frac:** Most zooplankton groups have R_frac > 0 (0.15–0.20), representing real energy allocated to reproduction. Below maturation size, this energy is redirected to somatic growth; above maturation size, it is lost from the modelled population (zooplankton recruitment uses boundary conditions that do not respond to reproductive investment). Flagellates and Ciliates have R_frac = 0 because they reproduce by binary fission. The energy budget framework supports future implementation of explicit zooplankton reproduction.

The explicit energy budget provides a mechanistic foundation for understanding how environmental change affects fish population dynamics through reproduction.

## Session Info

```{r}
sessionInfo()
```


