## Part 9: Reproductive Efficiency Sensitivity Analysis

The yield–effort curve should display the classic dome shape: catch increases with
effort to a maximum (MSY), then declines as the stock collapses under
overfishing. If the curve is monotonically increasing (or only weakly declining),
fish groups are replacing biomass too quickly — the stock–recruitment link is too
efficient.

Here we hold catchability constant at `q = 1` for all fish groups (so that
`F = effort × selectivity` directly) and sweep across effort levels for several
values of `repro_eff`. The current default is `repro_eff = 0.001` for all fish
groups.

```{r}
library(tidyverse)
library(patchwork)
```


## Helper Functions

We define a few helper functions used throughout the vignette for extracting and plotting fishing-specific outputs.

```{r}
# Extract catch data frame from model output
extract_catch <- function(mdl, fish_names = c("Fish_Small", "Fish_Med", "Fish_Large")) {
  data.frame(
    time = rep(mdl$time, times = length(fish_names)),
    catch = as.vector(mdl$catch),
    group = rep(fish_names, each = length(mdl$time))
  )
}

# Extract total biomass per group over time
extract_biomass_ts <- function(mdl) {
  species <- mdl$param$Groups$Species
  ngrps <- mdl$param$ngrps
  nsave <- length(mdl$time)
  
  # Sum biomass across size classes for each group
  bm <- matrix(NA, nrow = nsave, ncol = ngrps)
  for (g in 1:ngrps) {
    bm[, g] <- rowSums(mdl$biomass[g, , drop = FALSE] *
                          matrix(mdl$param$w, nrow = 1)[rep(1, nsave), ] *
                          mdl$param$dx)
  }
  # Actually use the abundance array directly
  bm <- matrix(NA, nrow = nsave, ncol = ngrps)
  for (g in 1:ngrps) {
    bm[, g] <- apply(mdl$abundance[, g, , drop = FALSE], 1, function(row) {
      sum(row * mdl$param$w * mdl$param$dx)
    })
  }
  
  df <- data.frame(
    time = rep(mdl$time, times = ngrps),
    biomass = as.vector(bm),
    species = rep(species, each = nsave),
    type = rep(mdl$param$Groups$Type, each = nsave)
  )
  df$species <- factor(df$species, levels = species)
  df
}

# Get consistent fish colours
get_fish_colours <- function(mdl) {
  fc <- mdl$param$Groups$PlotColour[mdl$param$fish_grps]
  names(fc) <- mdl$param$Groups$Species[mdl$param$fish_grps]
  fc
}

# Get all group colours
get_all_colours <- function(mdl) {
  ac <- mdl$param$Groups$PlotColour
  names(ac) <- mdl$param$Groups$Species
  ac
}
```

### Strategy

Because `GroupInputs` is just a data frame, we can modify `repro_eff` in-place
for the fish rows and pass the result straight to `zoomss_model()` — no package
code changes required.

```{r repro-eff-setup}
# Effort gradient (same as Part 4)
effort_levels <- c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 1.0, 1.5, 2.0)

# repro_eff values to explore: current default and two lower values
repro_eff_values <- c(1e-3)

# Load base groups
Groups_base <- getGroups()
fish_rows <- which(Groups_base$Type == "Fish")

# Fixed catchability = 1 for all fish groups
q_fixed <- 1.0
```

```{r repro-eff-sweep, cache=TRUE}
# Store all results in a nested list: results_repro[[re_idx]][[eff_idx]]
results_repro <- list()

for (re_idx in seq_along(repro_eff_values)) {

  # Modify repro_eff for fish groups
  Groups_re <- Groups_base
  Groups_re$repro_eff[fish_rows] <- repro_eff_values[re_idx]

  # Set constant catchability
  Groups_re <- setFishingParams(Groups_re,
                                q_small = q_fixed,
                                q_med   = q_fixed,
                                q_large = q_fixed)

  results_repro[[re_idx]] <- list()

  for (eff_idx in seq_along(effort_levels)) {

    env_re <- createInputParams(
      time = seq(0, 200, by = 0.1),
      sst  = 15,
      chl  = 1.0
    )

    env_re <- addFishingEffort(env_re,
                               effort_small = effort_levels[eff_idx],
                               effort_med   = effort_levels[eff_idx],
                               effort_large = effort_levels[eff_idx])

    results_repro[[re_idx]][[eff_idx]] <- zoomss_model(
      input_params = env_re,
      Groups       = Groups_re,
      isave        = 5
    )
  }
}
```

### Extract Equilibrium Summary

We average over the final 20 % of each run (years 160–200) to approximate
equilibrium values.

```{r repro-eff-extract}
fish_names <- c("Fish_Small", "Fish_Med", "Fish_Large")
num_fish   <- length(fish_names)

repro_summary <- data.frame()

for (re_idx in seq_along(repro_eff_values)) {
  for (eff_idx in seq_along(effort_levels)) {

    mdl_i     <- results_repro[[re_idx]][[eff_idx]]
    nsave     <- length(mdl_i$time)
    final_idx <- max(1, floor(nsave * 0.8)):nsave
    fish_grps <- mdl_i$param$fish_grps

    for (f in seq_len(num_fish)) {
      fg <- fish_grps[f]

      # Equilibrium biomass
      eq_bm <- mean(apply(
        mdl_i$abundance[final_idx, fg, , drop = FALSE], 1, function(row) {
          sum(row * mdl_i$param$w * mdl_i$param$dx)
        }))

      # Equilibrium catch
      eq_catch <- if (!is.null(mdl_i$catch)) {
        mean(mdl_i$catch[final_idx, f], na.rm = TRUE)
      } else { 0 }

      # Equilibrium SSB
      eq_ssb <- mean(mdl_i$SSB[final_idx, f], na.rm = TRUE)

      # Equilibrium recruitment
      eq_rec <- mean(mdl_i$recruitment[final_idx, f], na.rm = TRUE)

      repro_summary <- rbind(repro_summary, data.frame(
        repro_eff   = repro_eff_values[re_idx],
        effort      = effort_levels[eff_idx],
        fish_group  = fish_names[f],
        eq_catch    = eq_catch,
        eq_biomass  = eq_bm,
        eq_SSB      = eq_ssb,
        eq_recruitment = eq_rec
      ))
    }
  }
}

repro_summary$repro_eff_label <- paste0("repro_eff = ", format(repro_summary$repro_eff, scientific = TRUE))
```

### Yield–Effort Curves by Reproductive Efficiency

This is the key diagnostic: for each `repro_eff` value, does the yield–effort
curve show a dome?

```{r repro-eff-yield, fig.height=8, fig.width=10}
fish_colours <- get_fish_colours(results_repro[[1]][[1]])

p_yield_re <- ggplot(repro_summary,
                     aes(x = effort, y = eq_catch, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ repro_eff_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  theme_bw(base_size = 12) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Catch (g WW)",
       colour = "Fish Group",
       title  = "Yield–Effort Curves Across Reproductive Efficiency Values",
       subtitle = paste0("q = ", q_fixed, " for all fish groups"))

p_yield_re
```

### Biomass Depletion by Reproductive Efficiency

How severely is each stock depleted? Lower `repro_eff` should make stocks more
vulnerable — biomass should decline more steeply with effort.

```{r repro-eff-biomass, fig.height=8, fig.width=10}
p_bm_re <- ggplot(repro_summary,
                  aes(x = effort, y = eq_biomass, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ repro_eff_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw(base_size = 12) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Biomass (log scale)",
       colour = "Fish Group",
       title  = "Biomass Depletion Across Reproductive Efficiency Values")

p_bm_re
```

### SSB and Recruitment Response

```{r repro-eff-ssb-rec, fig.height=10, fig.width=10}
p_ssb_re <- ggplot(repro_summary,
                   aes(x = effort, y = eq_SSB, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ repro_eff_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw(base_size = 12) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium SSB (log scale)",
       colour = "Fish Group",
       title  = "SSB Response Across Reproductive Efficiency Values")

p_rec_re <- ggplot(repro_summary,
                   aes(x = effort, y = eq_recruitment, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ repro_eff_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw(base_size = 12) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Recruitment (log scale)",
       colour = "Fish Group",
       title  = "Recruitment Response Across Reproductive Efficiency Values")

wrap_plots(p_ssb_re, p_rec_re, ncol = 1, guides = "collect")
```

### Normalised Yield Curves (Relative to Maximum)

Normalising each group's yield to its own maximum makes the dome shape (or
absence thereof) easier to compare across `repro_eff` treatments.

```{r repro-eff-normalised, fig.height=6, fig.width=10}
repro_summary <- repro_summary %>%
  group_by(repro_eff_label, fish_group) %>%
  mutate(yield_norm = eq_catch / max(eq_catch, na.rm = TRUE)) %>%
  ungroup()

p_norm <- ggplot(repro_summary,
                 aes(x = effort, y = yield_norm, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ repro_eff_label, ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  theme_bw(base_size = 12) +
  labs(x      = "Effort Multiplier",
       y      = "Normalised Yield (relative to max)",
       colour = "Fish Group",
       title  = "Normalised Yield–Effort Curves",
       subtitle = "Dome shape indicates MSY; monotonic increase = stock too resilient")

p_norm
```

### Combined Overlay: All repro_eff Treatments on One Panel

A single-panel comparison per fish group to directly see the effect of reducing
reproductive efficiency.

```{r repro-eff-overlay, fig.height=5, fig.width=10}
ggplot(repro_summary,
       aes(x = effort, y = eq_catch,
           colour = repro_eff_label, group = repro_eff_label)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(~ fish_group, scales = "free_y") +
  theme_bw(base_size = 12) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Catch (g WW)",
       colour = "Repro. Efficiency",
       title  = "Effect of Reproductive Efficiency on Yield–Effort Relationship")
```

### Interpretation

If the dome shape only emerges at lower `repro_eff` values:

- The current default (`repro_eff = 0.001`) is too high — stocks can replace
  harvested biomass faster than fishing removes it, preventing the
  overexploitation phase of the yield curve.
- The appropriate `repro_eff` should produce a dome where MSY occurs at a
  realistic effort level (roughly `effort = 1–2` depending on the group).

If none of the tested values produce a dome, the resilience may be coming from
elsewhere in the model (e.g. density-dependent growth, the stock–recruitment
functional form, or the interaction between fish groups via shared prey).



### Strategy

Because `GroupInputs` is just a data frame, we can modify `repro_eff` in-place
for the fish rows and pass the result straight to `zoomss_model()` — no package
code changes required.

```{r repro-eff-setup}
# Effort gradient (same as Part 4)
effort_levels <- c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 1.0, 1.5, 2.0)

# repro_eff values to explore: current default and two lower values
repro_eff_values <- c(1e-5)

# Load base groups
Groups_base <- getGroups()
fish_rows <- which(Groups_base$Type == "Fish")

# Fixed catchability = 1 for all fish groups
q_fixed <- 1.0
```

```{r repro-eff-sweep, cache=TRUE}
# Store all results in a nested list: results_repro[[re_idx]][[eff_idx]]
results_repro <- list()

for (re_idx in seq_along(repro_eff_values)) {

  # Modify repro_eff for fish groups
  Groups_re <- Groups_base
  Groups_re$repro_eff[fish_rows] <- repro_eff_values[re_idx]

  # Set constant catchability
  Groups_re <- setFishingParams(Groups_re,
                                q_small = q_fixed,
                                q_med   = q_fixed,
                                q_large = q_fixed)

  results_repro[[re_idx]] <- list()

  for (eff_idx in seq_along(effort_levels)) {

    env_re <- createInputParams(
      time = seq(0, 200, by = 0.1),
      sst  = 15,
      chl  = 1.0
    )

    env_re <- addFishingEffort(env_re,
                               effort_small = effort_levels[eff_idx],
                               effort_med   = effort_levels[eff_idx],
                               effort_large = effort_levels[eff_idx])

    results_repro[[re_idx]][[eff_idx]] <- zoomss_model(
      input_params = env_re,
      Groups       = Groups_re,
      isave        = 5
    )
  }
}
```






### Extract Equilibrium Summary

We average over the final 20 % of each run (years 160–200) to approximate
equilibrium values.

```{r repro-eff-extract}
fish_names <- c("Fish_Small", "Fish_Med", "Fish_Large")
num_fish   <- length(fish_names)

repro_summary <- data.frame()

for (re_idx in seq_along(repro_eff_values)) {
  for (eff_idx in seq_along(effort_levels)) {

    mdl_i     <- results_repro[[re_idx]][[eff_idx]]
    nsave     <- length(mdl_i$time)
    final_idx <- max(1, floor(nsave * 0.8)):nsave
    fish_grps <- mdl_i$param$fish_grps

    for (f in seq_len(num_fish)) {
      fg <- fish_grps[f]

      # Equilibrium biomass
      eq_bm <- mean(apply(
        mdl_i$abundance[final_idx, fg, , drop = FALSE], 1, function(row) {
          sum(row * mdl_i$param$w * mdl_i$param$dx)
        }))

      # Equilibrium catch
      eq_catch <- if (!is.null(mdl_i$catch)) {
        mean(mdl_i$catch[final_idx, f], na.rm = TRUE)
      } else { 0 }

      # Equilibrium SSB
      eq_ssb <- mean(mdl_i$SSB[final_idx, f], na.rm = TRUE)

      # Equilibrium recruitment
      eq_rec <- mean(mdl_i$recruitment[final_idx, f], na.rm = TRUE)

      repro_summary <- rbind(repro_summary, data.frame(
        repro_eff   = repro_eff_values[re_idx],
        effort      = effort_levels[eff_idx],
        fish_group  = fish_names[f],
        eq_catch    = eq_catch,
        eq_biomass  = eq_bm,
        eq_SSB      = eq_ssb,
        eq_recruitment = eq_rec
      ))
    }
  }
}

repro_summary$repro_eff_label <- paste0("repro_eff = ", format(repro_summary$repro_eff, scientific = TRUE))
```





### Combined Overlay: All repro_eff Treatments on One Panel

A single-panel comparison per fish group to directly see the effect of reducing
reproductive efficiency.

```{r repro-eff-overlay, fig.height=5, fig.width=10}
ggplot(repro_summary,
       aes(x = effort, y = eq_catch,
           colour = repro_eff_label, group = repro_eff_label)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(~ fish_group, scales = "free_y") +
  theme_bw(base_size = 12) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Catch (g WW)",
       colour = "Repro. Efficiency",
       title  = "Effect of Reproductive Efficiency on Yield–Effort Relationship")
```






























