---
title: "Reproductive Efficiency Sensitivity Analysis in ZooMSS"
subtitle: "Uniform and Group-Specific repro_eff Across an Effort Gradient"
author: "ZooMSS Development"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    number_sections: true
    theme: flatly
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Reproductive Efficiency Sensitivity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 10,
  fig.height = 6,
  message  = FALSE,
  warning  = FALSE
)
```

```{r libraries, message=FALSE, warning=FALSE}
library(here)
devtools::load_all(here())
library(ggplot2)
library(patchwork)
library(dplyr)
```

# Introduction

The yield–effort curve should display the classic dome shape: catch increases
with effort to a maximum (MSY), then declines as the stock collapses under
overfishing. If the curve is monotonically increasing (or only weakly
declining), fish groups are replacing biomass too quickly — the
stock–recruitment link is too efficient.

The reproductive efficiency parameter `repro_eff` controls the fraction of
reproductive output (eggs/larvae) that survive to become recruits at the
smallest fish size class. It is the primary lever determining how resilient
fish populations are to fishing mortality.

This analysis explores `repro_eff` in two stages:

1. **Uniform `repro_eff`** (Part 1): The same value applied to all three fish
   groups, swept across four orders of magnitude.
2. **Group-specific `repro_eff`** (Part 2): Different values for each fish
   group, testing whether differential reproductive efficiency produces
   realistic differential vulnerability to fishing.

Throughout, catchability is held constant at `q = 1` so that
`F = effort × selectivity` directly, and the effort gradient ranges from 0
to 5.

## Prerequisites

Model runs are computationally expensive. This document **loads pre-computed
results** from `.rds` files generated by the companion script
`run_repro_eff_sweep.R`. Run that script first:

```
Rscript run_repro_eff_sweep.R
```

Results are saved to the `repro_eff_cache/` directory. If the cache files are
not found, this document will stop with an informative error.

# Load Cached Results

```{r load-cache}
cache_dir <- "repro_eff_cache"

# Validate cache exists
required_files <- c("results_uniform_repro_v2.rds",
                    "results_group_repro_v2.rds",
                    "sweep_config.rds")
missing <- required_files[!file.exists(file.path(cache_dir, required_files))]

if (length(missing) > 0) {
  stop("Missing cached results in '", cache_dir, "/':\n  ",
       paste(missing, collapse = "\n  "),
       "\n\nRun 'Rscript run_repro_eff_sweep.R' first to generate them.")
}

# Load
sweep_config        <- readRDS(file.path(cache_dir, "sweep_config.rds"))
results_repro       <- readRDS(file.path(cache_dir, "results_uniform_repro.rds"))
results_group_repro <- readRDS(file.path(cache_dir, "results_group_repro.rds"))

cat("Loaded cached results generated on:", format(sweep_config$timestamp), "\n")
cat("R version used for sweep:", sweep_config$r_version, "\n")
cat("Cores used:", sweep_config$n_cores_used, "\n")

# Check if Part 2 data is available (allows rendering with Part 1 only)
has_part2 <- length(results_group_repro) > 0 &&
             length(results_group_repro[[1]]) > 0
if (!has_part2) cat("\nNote: Part 2 results not yet available — those sections will be skipped.\n")
```

```{r unpack-config}
# Unpack configuration so downstream chunks can reference them directly
effort_levels    <- sweep_config$effort_levels
repro_eff_values <- sweep_config$repro_eff_values
repro_scenarios  <- sweep_config$repro_scenarios
scenario_names   <- sweep_config$scenario_names
q_fixed          <- sweep_config$q_fixed
fish_names       <- sweep_config$fish_names
num_fish         <- length(fish_names)
```

# Helper Functions

```{r helpers}
# Get consistent fish colours from a model object
get_fish_colours <- function(mdl) {
  fc <- mdl$param$Groups$PlotColour[mdl$param$fish_grps]
  names(fc) <- mdl$param$Groups$Species[mdl$param$fish_grps]
  fc
}

# Get all group colours
get_all_colours <- function(mdl) {
  ac <- mdl$param$Groups$PlotColour
  names(ac) <- mdl$param$Groups$Species
  ac
}

# Extract equilibrium metrics for one model run and one fish group
extract_equilibrium <- function(mdl_i, f, fish_grps) {
  nsave     <- length(mdl_i$time)
  final_idx <- max(1, floor(nsave * 0.5)):nsave
  fg        <- fish_grps[f]

  eq_bm <- mean(apply(
    mdl_i$abundance[final_idx, fg, , drop = FALSE], 1, function(row) {
      sum(row * mdl_i$param$w * mdl_i$param$dx)
    }))

  eq_catch <- if (!is.null(mdl_i$catch)) {
    mean(mdl_i$catch[final_idx, f], na.rm = TRUE)
  } else { 0 }

  eq_ssb <- mean(mdl_i$SSB[final_idx, f], na.rm = TRUE)
  eq_rec <- mean(mdl_i$recruitment[final_idx, f], na.rm = TRUE)

  data.frame(eq_catch = eq_catch, eq_biomass = eq_bm,
             eq_SSB = eq_ssb, eq_recruitment = eq_rec)
}
```


# Part 1: Uniform Reproductive Efficiency {.tabset}

## Extract Equilibrium Summary

Averaged over the final 200 years of each 400-year run (years 200–400) to
approximate long-term steady state.

```{r uniform-extract}
repro_summary <- data.frame()

for (re_idx in seq_along(repro_eff_values)) {
  for (eff_idx in seq_along(effort_levels)) {

    mdl_i     <- results_repro[[re_idx]][[eff_idx]]
    fish_grps <- mdl_i$param$fish_grps

    for (f in seq_len(num_fish)) {
      eq <- extract_equilibrium(mdl_i, f, fish_grps)
      repro_summary <- rbind(repro_summary, cbind(
        data.frame(repro_eff  = repro_eff_values[re_idx],
                   effort     = effort_levels[eff_idx],
                   fish_group = fish_names[f]),
        eq
      ))
    }
  }
}

repro_summary$repro_eff_label <- paste0(
  "repro_eff = ", format(repro_summary$repro_eff, scientific = TRUE))
```

## Yield–Effort Curves {.tabset}

The key diagnostic: for each `repro_eff` value, does the yield–effort curve
show a dome?

### Absolute Yield

```{r uniform-yield, fig.height=8}
fish_colours <- get_fish_colours(results_repro[[1]][[1]])

ggplot(repro_summary,
       aes(x = effort, y = eq_catch, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ repro_eff_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  theme_bw(base_size = 12) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Catch (g WW)",
       colour = "Fish Group",
       title  = "Yield-Effort Curves Across Reproductive Efficiency Values",
       subtitle = paste0("q = ", q_fixed, " for all fish groups"))
```

### Normalised Yield

Normalising each group's yield to its own maximum makes the dome shape (or
absence thereof) easier to compare.

```{r uniform-normalised, fig.height=8}
repro_summary <- repro_summary %>%
  group_by(repro_eff_label, fish_group) %>%
  mutate(yield_norm = eq_catch / max(eq_catch, na.rm = TRUE)) %>%
  ungroup()

ggplot(repro_summary,
       aes(x = effort, y = yield_norm, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ repro_eff_label, ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  theme_bw(base_size = 12) +
  labs(x      = "Effort Multiplier",
       y      = "Normalised Yield (relative to max)",
       colour = "Fish Group",
       title  = "Normalised Yield-Effort Curves",
       subtitle = "Dome shape indicates MSY; monotonic increase = stock too resilient")
```

## Biomass Depletion

Lower `repro_eff` should make stocks more vulnerable — biomass should decline
more steeply with effort.

```{r uniform-biomass, fig.height=8}
ggplot(repro_summary,
       aes(x = effort, y = eq_biomass, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ repro_eff_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw(base_size = 12) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Biomass (log scale)",
       colour = "Fish Group",
       title  = "Biomass Depletion Across Reproductive Efficiency Values")
```

## SSB and Recruitment

```{r uniform-ssb-rec, fig.height=10}
p_ssb_re <- ggplot(repro_summary,
                   aes(x = effort, y = eq_SSB, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ repro_eff_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw(base_size = 12) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium SSB (log scale)",
       colour = "Fish Group",
       title  = "SSB Response Across Reproductive Efficiency Values")

p_rec_re <- ggplot(repro_summary,
                   aes(x = effort, y = eq_recruitment, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ repro_eff_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw(base_size = 12) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Recruitment (log scale)",
       colour = "Fish Group",
       title  = "Recruitment Response Across Reproductive Efficiency Values")

wrap_plots(p_ssb_re, p_rec_re, ncol = 1, guides = "collect")
```

## Combined Overlay

A single-panel comparison per fish group to directly see the effect of
reducing reproductive efficiency.

```{r uniform-overlay, fig.height=5}
ggplot(repro_summary,
       aes(x = effort, y = eq_catch,
           colour = repro_eff_label, group = repro_eff_label)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(~ fish_group, scales = "free_y") +
  theme_bw(base_size = 12) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Catch (g WW)",
       colour = "Repro. Efficiency",
       title  = "Effect of Reproductive Efficiency on Yield-Effort Relationship")
```

## Interpretation

If the dome shape only emerges at lower `repro_eff` values:

- The current default (`repro_eff = 0.001`) is too high — stocks can replace
  harvested biomass faster than fishing removes it, preventing the
  overexploitation phase of the yield curve.
- The appropriate `repro_eff` should produce a dome where MSY occurs at a
  realistic effort level (roughly `effort = 1–2` depending on the group).

If none of the tested values produce a dome, the resilience may be coming from
elsewhere in the model (e.g. density-dependent growth, the stock–recruitment
functional form, or the interaction between fish groups via shared prey).


# Part 2: Group-Specific Reproductive Efficiency {.tabset}

Part 1 applied the same `repro_eff` to all three fish groups. In practice,
reproductive efficiency should differ by life history: small, fast-turnover
fish (r-strategists) typically have higher egg-to-recruit survival than
large, slow-growing fish (K-strategists) because small fish mature earlier,
produce more frequent cohorts, and face lower cumulative larval mortality
relative to their adult population turnover.

Here we test a set of **group-specific `repro_eff` combinations** across the
same effort gradient to identify whether differential reproductive efficiency
produces more realistic yield curves — in particular, whether large fish
collapse at lower effort while small fish remain productive.

### Scenario Definitions

```{r group-scenarios, eval=has_part2}
# Display scenario table
scenario_df <- do.call(rbind, lapply(seq_along(repro_scenarios), function(i) {
  data.frame(Scenario   = scenario_names[i],
             Fish_Small = repro_scenarios[[i]][1],
             Fish_Med   = repro_scenarios[[i]][2],
             Fish_Large = repro_scenarios[[i]][3])
}))

knitr::kable(scenario_df, caption = "Group-specific repro_eff scenarios",
             col.names = c("Scenario", "Fish_Small", "Fish_Med", "Fish_Large"))
```

## Extract Equilibrium Summary

```{r group-extract, eval=has_part2}
group_repro_summary <- data.frame()

for (sc_idx in seq_along(repro_scenarios)) {

  re_vec  <- repro_scenarios[[sc_idx]]
  sc_name <- scenario_names[sc_idx]

  for (eff_idx in seq_along(effort_levels)) {

    mdl_i     <- results_group_repro[[sc_idx]][[eff_idx]]
    fish_grps <- mdl_i$param$fish_grps

    for (f in seq_len(num_fish)) {
      eq <- extract_equilibrium(mdl_i, f, fish_grps)
      group_repro_summary <- rbind(group_repro_summary, cbind(
        data.frame(scenario     = sc_name,
                   repro_eff_fg = re_vec[f],
                   effort       = effort_levels[eff_idx],
                   fish_group   = fish_names[f]),
        eq
      ))
    }
  }
}

group_repro_summary$scenario <- factor(group_repro_summary$scenario,
                                        levels = scenario_names)

# Build informative facet labels showing per-group values
scenario_subtitles <- sapply(seq_along(repro_scenarios), function(i) {
  re <- repro_scenarios[[i]]
  paste0(scenario_names[i], "\n(S:", format(re[1], sci = TRUE),
         " M:", format(re[2], sci = TRUE),
         " L:", format(re[3], sci = TRUE), ")")
})
names(scenario_subtitles) <- scenario_names

group_repro_summary$scenario_label <- scenario_subtitles[
  as.character(group_repro_summary$scenario)]
group_repro_summary$scenario_label <- factor(
  group_repro_summary$scenario_label, levels = scenario_subtitles)
```

## Yield–Effort Curves {.tabset}

Each facet is a scenario with different `repro_eff` combinations. Fish groups
within each panel now respond differently to the same effort gradient.

### Absolute Yield

```{r group-yield, fig.height=10, eval=has_part2}
fish_colours <- get_fish_colours(results_group_repro[[1]][[1]])

ggplot(group_repro_summary,
       aes(x = effort, y = eq_catch, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ scenario_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  theme_bw(base_size = 11) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Catch (g WW)",
       colour = "Fish Group",
       title  = "Yield-Effort Curves With Group-Specific Reproductive Efficiency",
       subtitle = paste0("q = ", q_fixed, " for all fish groups"))
```

### Normalised Yield

Normalised to each group's own maximum within each scenario.

```{r group-normalised, fig.height=10, eval=has_part2}
group_repro_summary <- group_repro_summary %>%
  group_by(scenario, fish_group) %>%
  mutate(yield_norm = eq_catch / max(eq_catch, na.rm = TRUE)) %>%
  ungroup()

ggplot(group_repro_summary,
       aes(x = effort, y = yield_norm, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ scenario_label, ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  theme_bw(base_size = 11) +
  labs(x      = "Effort Multiplier",
       y      = "Normalised Yield (relative to max)",
       colour = "Fish Group",
       title  = "Normalised Yield-Effort Curves With Group-Specific Reproductive Efficiency",
       subtitle = "Dome shape indicates MSY; monotonic increase = stock too resilient")
```

## Biomass Depletion

```{r group-biomass, fig.height=10, eval=has_part2}
ggplot(group_repro_summary,
       aes(x = effort, y = eq_biomass, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ scenario_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw(base_size = 11) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Biomass (log scale)",
       colour = "Fish Group",
       title  = "Biomass Depletion With Group-Specific Reproductive Efficiency")
```

## SSB and Recruitment

```{r group-ssb-rec, fig.height=12, eval=has_part2}
p_ssb_grp <- ggplot(group_repro_summary,
                    aes(x = effort, y = eq_SSB, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ scenario_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw(base_size = 11) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium SSB (log scale)",
       colour = "Fish Group",
       title  = "SSB Response With Group-Specific Reproductive Efficiency")

p_rec_grp <- ggplot(group_repro_summary,
                    aes(x = effort, y = eq_recruitment, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  facet_wrap(~ scenario_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  scale_y_log10() +
  theme_bw(base_size = 11) +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Recruitment (log scale)",
       colour = "Fish Group",
       title  = "Recruitment Response With Group-Specific Reproductive Efficiency")

wrap_plots(p_ssb_grp, p_rec_grp, ncol = 1, guides = "collect")
```

## Cross-Scenario Comparison

Overlay all scenarios on one panel per fish group to directly compare how
different `repro_eff` combinations affect each group's yield curve.

```{r group-overlay, fig.height=6, fig.width=12, eval=has_part2}
ggplot(group_repro_summary,
       aes(x = effort, y = eq_catch,
           colour = scenario, group = scenario)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(~ fish_group, scales = "free_y") +
  theme_bw(base_size = 12) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x      = "Effort Multiplier",
       y      = "Equilibrium Catch (g WW)",
       colour = "Scenario",
       title  = "Cross-Scenario Comparison of Yield-Effort by Fish Group")
```

## MSY Summary Table

The effort level producing maximum yield for each fish group under each
scenario.

```{r group-msy-table, eval=has_part2}
msy_table <- group_repro_summary %>%
  group_by(scenario, fish_group) %>%
  summarise(
    MSY            = max(eq_catch, na.rm = TRUE),
    effort_MSY     = effort[which.max(eq_catch)],
    SSB_at_MSY     = eq_SSB[which.max(eq_catch)],
    biomass_at_MSY = eq_biomass[which.max(eq_catch)],
    repro_eff      = unique(repro_eff_fg),
    .groups = "drop"
  ) %>%
  arrange(scenario, fish_group)

knitr::kable(msy_table, digits = 6,
             caption = "MSY summary by scenario and fish group",
             col.names = c("Scenario", "Fish Group", "MSY (g WW)",
                          "Effort at MSY", "SSB at MSY", "Biomass at MSY",
                          "repro_eff"))
```

## Depletion Ratio (B/B~0~)

The depletion ratio (`B/B0`, biomass relative to unfished) is a standard
fisheries reference metric.

```{r group-depletion, fig.height=10, eval=has_part2}
group_repro_summary <- group_repro_summary %>%
  group_by(scenario, fish_group) %>%
  mutate(B0 = eq_biomass[effort == 0],
         depletion = eq_biomass / B0) %>%
  ungroup()

ggplot(group_repro_summary,
       aes(x = effort, y = depletion, colour = fish_group)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0.4, linetype = "dashed",
             colour = "grey50", linewidth = 0.5) +
  geom_hline(yintercept = 0.2, linetype = "dotted",
             colour = "red", linewidth = 0.5) +
  annotate("text", x = max(effort_levels) * 0.95, y = 0.42,
           label = "B_MSY proxy (0.4)", hjust = 1, size = 2.5, colour = "grey40") +
  annotate("text", x = max(effort_levels) * 0.95, y = 0.22,
           label = "B_lim (0.2)", hjust = 1, size = 2.5, colour = "red") +
  facet_wrap(~ scenario_label, ncol = 2) +
  scale_colour_manual(values = fish_colours) +
  coord_cartesian(ylim = c(0, 1.05)) +
  theme_bw(base_size = 11) +
  labs(x      = "Effort Multiplier",
       y      = expression(B / B[0]),
       colour = "Fish Group",
       title  = "Stock Depletion With Group-Specific Reproductive Efficiency",
       subtitle = "Dashed = B_MSY proxy (0.4); Dotted = B_lim (0.2)")
```

## Interpretation

Key patterns to look for across scenarios:

1. **Differential vulnerability**: In gradient scenarios, large fish should
   collapse at lower effort than small fish. If Fish_Large with
   `repro_eff = 1e-6` shows a dome peak at `effort ~ 0.5` while Fish_Small
   with `repro_eff = 1e-2` peaks at `effort ~ 3`, this reproduces the
   empirical pattern of "fishing down the food web" where large species are
   depleted first.

2. **Trophic feedback**: When large fish collapse, medium and small fish may
   experience reduced predation mortality and temporarily increase — a
   mesopredator release. Group-specific `repro_eff` interacts with these
   trophic dynamics in ways that uniform `repro_eff` cannot capture.

3. **MSY effort separation**: Realistic fisheries management requires
   different effort levels for different species. If all three groups have MSY
   at the same effort, the model cannot distinguish selective management
   strategies.

4. **Depletion thresholds**: The `B/B0` plots with reference lines at 0.4
   (BMSY proxy) and 0.2 (Blim) show which scenarios produce realistic stock
   assessment dynamics.

The scenario that best reproduces these patterns — differential vulnerability,
dome-shaped yield curves for all groups, and realistic depletion levels —
provides the starting point for calibrating `repro_eff` against observed
catch data in the FishMIP framework.


# Session Info

```{r session-info}
sessionInfo()
```
