---
title: "Getting started with ZooMSS"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    number_sections: false
vignette: >
  %\VignetteIndexEntry{zoomss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r message=FALSE, warning=FALSE}
library(zoomss)
library(ggplot2)
library(dplyr)
```

## Input Data

ZooMSS requires two sets of input data:

1. **Groups** - Contains all taxa-specific parameter values for each model group, including size ranges and functional group properties.

2. **Environmental data** - A Time-series dataframe with time series of environmental conditions  with `time`, `sst`, and `chl` columns.


## Full Reproductive Efficiency Scenario

Before running the default model, we first explore a scenario where the reproductive efficiency (`repro_eff`) of all fish groups is set to 1. This removes the egg-to-recruit mortality bottleneck and allows us to investigate how model dynamics respond when all reproductive output successfully recruits.

Load the default groups and modify `repro_eff` for fish:

```{r}
Groups_repro1 <- getGroups()

# Set repro_eff to 1 for all fish groups
# Groups_repro1$repro_eff[Groups_repro1$Type == "Fish"] <- .001


# Groups_repro1$W0[Groups_repro1$Species == "Flagellates"] <- -12

Groups_repro1$f_M[Groups_repro1$FeedType == "Carnivore"] <- 0.3

Groups_repro1$f_M[Groups_repro1$FeedType == "Carnivore"] <- 0.3

Groups_repro1$K_growth[Groups_repro1$FeedType == "Carnivore"] <- 0.6

Groups_repro1$f_M[Groups_repro1$Species == "Salps"] <- 0.6
Groups_repro1$K_growth[Groups_repro1$Species == "Salps"] <- 0.4



# Groups_repro1 <- Groups_repro1 %>% 
#   filter(!Species %in% c("Fish_Med", "Fish_Large"))


# Check the change
# Groups_repro1[Groups_repro1$Type == "Fish", c("Species", "Type", "repro_eff")]
```

Run the model with the modified groups under constant environmental conditions:

```{r}
env_repro1 <- createInputParams(time = seq(0, 200, by = 0.1),
                                sst = 15,
                                chl = 0.01)

mdl_repro1 <- zoomss_model(input_params = env_repro1, Groups = Groups_repro1, isave = 2) #,
                           # energy_budget_scenario = "B")
```

### Compare dynamics

Plot biomass time series to see how the elevated reproductive efficiency affects community structure:

```{r}
library(patchwork)

p_repro1_biomass <- plotTimeSeries(mdl_repro1, by = "biomass", transform = "log10") +
  theme(legend.position = "none")

p_repro1_fill <- plotTimeSeries(mdl_repro1, by = "biomass", type = "fill") 

wrap_plots(p_repro1_biomass, p_repro1_fill, nrow = 2, guides = "collect")
```

Plot the size spectra and compare the shape under full reproductive efficiency:

```{r}
plotSizeSpectra(mdl_repro1, n_years = 100) +
  ggtitle("Size spectra (repro_eff = 1)")
```

## Running the default Model

Get the default published `Groups` dataframe using:

```{r}
Groups <- getGroups()
```

Now create an environmental data time-series using the helper function. This time-series uses a constant sea surface temperature (`sst`) and chlorophyll *a*  (`chl`) with a 0.1 yr-1 timestep (`dt`). 

```{r}
env_data <- createInputParams(time = seq(0, 100, by = 0.1) ,
                              sst = 15,
                              chl = 0.15)

```

We can look at the environment data and check everything is ok with: 

```{r}
plotEnvironment(env_data)
```


Now we run ZooMSS and save every `isave` timestep to reduce storage requirements. 
```{r}
mdl <- zoomss_model(input_params = env_data, Groups = Groups, isave = 2)

```

## Plotting

The model includes several built-in plotting functions for analysis and visualization.


### Time Series Analysis

These plots display total abundance and mean growth/mortality across all size classes through time.

```{r}

library(patchwork)
p1 <- plotTimeSeries(mdl, by = "abundance", transform = "log10") # Plot abundance time series
p2 <- plotTimeSeries(mdl, by = "growth") # Plot growth rate time series
p3 <- plotTimeSeries(mdl, by = "mortality") # Plot predation mortality time series

wrap_plots(p1, p2, p3, nrow = 3, guides = "collect")

```


We can also plot total biomass through time.

```{r}
p4 <- plotTimeSeries(mdl, by = "biomass", transform = "log10") + theme(legend.position = "none") # Plot biomass 
p5 <- plotTimeSeries(mdl, by = "biomass", type = "stack", transform = "log10") # Plot stacked biomass 
p6 <- plotTimeSeries(mdl, by = "biomass", type = "fill") # Plot proportional stacked biomass 

wrap_plots(p4, p5, p6, nrow = 3, guides = "collect")

```


### Static Plots for a given model time point

Plot mean species-resolved size spectra for the final `n_years`.
```{r}
plotSizeSpectra(mdl, n_years = 10)
```

Plot predator-prey mass ratios for the `idx` timestep
```{r message=FALSE, warning=FALSE}
plotPPMR(mdl, idx = 500) # Plot final timestep
```
